---
title: "Fibonacci Numbers"
index_entry: "Matrix population model for generating the Fibonacci numbers"
bibliography: ../../references.bib
link-citations: TRUE
author: Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "./figures/"
)
```

This example demonstrates that `macpan2` can simulate any discrete-time dynamical model—not just compartmental models. Use the `~` operator in the `during` list to define how model variables are updated at each time step. For example in this article, we generate the [Fibonacci sequence](https://en.wikipedia.org/wiki/Fibonacci_sequence) using its [matrix-based dynamical form](https://en.wikipedia.org/wiki/Fibonacci_sequence#Matrix_form).

# Packages Used and Settings

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(dplyr)
library(macpan2)
```


# Model Specification

This model has been specified in the `fibonacci` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/fibonacci/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
spec = mp_tmb_library(
    "starter_models"
  , "fibonacci"
  , package = "macpan2"
)
```

This model contains two variables -- a matrix `A` and vector `x`.
```{r default_vars}
mp_default_list(spec)
```

At every iteration of the simulation we multiply `A` by `x` to get a new value of `x`. Then we pick out the next Fibonacci number as the second (zero-based) element of `x`.
```{r during}
mp_print_during(spec)
```

By the end of the simulation loop, if we run enough iterations, we compute a very good approximation to the golden ratio.
```{r after}
mp_print_after(spec)
```

# Dynamics 

Here is the mathematical description of this model.

$$
\begin{pmatrix} 
x_0(t+1) \\ 
x_1(t+1) 
\end{pmatrix}
= \begin{pmatrix} 
1 & 1 \\ 
1 & 0 
\end{pmatrix}
\begin{pmatrix} 
x_0(t) \\ 
x_1(t) 
\end{pmatrix}
$$

# Simulation

The Fibonacci numbers are given by the following simulation.
```{r fib_sims}
(spec 
  |> mp_simulator(
      time_steps = 10
    , output = "fibonacci"
  ) 
  |> mp_trajectory()
  |> pull(value)
)
```

And by running the simulation for longer we can get a very good approximation to the golden ratio.
```{r golden_sims}
(spec 
  |> mp_simulator(
      time_steps = 100
    , output = "golden_ratio"
  ) 
  |> mp_final()
  |> pull(value)
)
```
