---
title: "SHIVER = SEIR + H + V"
index_entry: "A modified SEIR model with Hospitalization and Vaccination"
bibliography: references.bib
link-citations: TRUE
author: Jennifer Freeman
output: 
  github_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

```{r packages, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```

```{r local_function, include=FALSE}
## -------------------------
## local function
## -------------------------

# to be included in mp_tmb_coef in the future
# see here, https://github.com/canmod/macpan2/issues/179
backtrans <- function(x) {
  vars1 <- intersect(c("default", "estimate", "conf.low", "conf.high"), names(x))
  prefix <- stringr::str_extract(x[["mat"]], "^log(it)?_")  |> tidyr::replace_na("none")
  sx <- split(x, prefix)
  for (ptype in setdiff(names(sx), "none")) {
    link <- make.link(stringr::str_remove(ptype, "_"))
    sx[[ptype]] <- (sx[[ptype]]
                    |> mutate(across(std.error, ~link$mu.eta(estimate)*.))
                    |> mutate(across(any_of(vars1), link$linkinv))
                    |> mutate(across(mat, ~stringr::str_remove(., paste0("^", ptype))))
    )
  }
  bind_rows(sx)
}

```

This model builds on the basic SEIR model, with two additional compartments for vaccination and hospitalizations. 

Vaccines are typically subject to resource constraints and distribution strategies might prioritize vaccinations for specific subpopulations, such as immunocompromised people, to reduce bad outcomes. We model this with a flow of susceptibles entering the vaccination class. Following vaccination, most individuals return to the susceptible class when the acquired immune response wears off, called *vaccine waning*.

In reality, vaccinations are only partly effective in developing a protective host immune response. This *vaccine failure* includes both vaccine-related inadequacies and host-related reasons like health status and genetic factors [@wiedermann_2016]. Vaccine failure is modelled as flow from vaccination to the exposed class.

The SHIVER model assumes all individuals can be exposed to the disease, but transmission rates depend on vaccination status. Additionally, individuals with severe infections are hospitalized and assumed to be isolated, before recovering from the disease. Hospital isolation means this portion of infectious individuals no longer contribute to the transmission dynamics. 

# States

| variable | description                        |
| -------- | ---------------------------------- |
| S        | Number of susceptible individuals  |
| H        | Number of hospitalized individuals |
| I        | Number of infectious individuals   |
| V        | Number of vaccinated individuals   |
| E        | Number of exposed individuals      |
| R        | Number of recovered individuals    |

The size of the total population is,  $N = S + H + I  + V + E + R$, and the disease spreads through homogenous mixing of the subpopulation $N_{\text{mix}}=N -H$.

# Parameters

| variable   | description                                                                                         |
| ---------- | --------------------------------------------------------------------------------------------------- |
| $\phi$     | per capita vaccination rate of susceptibles                                                         |
| $\rho$     | per capita vaccine waning rate                                                                      |
| $\beta_S$  | per capita transmission rate for susceptibles (in $N_{\text{mix}}$ population)                      |
| $\beta_V$  | per capita transmission rate for vaccinated individuals (in $N_{\text{mix}}$ population)            |
| $\alpha$   | per capita infection rate (average time spent in compartment $E$ is $1/\alpha$)                     |  |
| $\gamma_I$ | per capita recovery rate for infected  individuals                                                  |
| $\gamma_H$ | per capita recovery rate for hospitalized individuals                                               |
| $\sigma$   | per capita rate at which infected individuals develop severe infections and require hospitalization |

# Dynamics 

$$
\begin{align*}
\frac{dS}{dt} &= -\beta_{S} S\frac{I}{N_{\text{mix}}} - \phi S + \rho V\\
\frac{dV}{dt} &=  \phi S - \rho V - \beta_{V} V\frac{I}{N_{\text{mix}}} \\
\frac{dE}{dt} &= (\beta_{S} S + \beta_{V} V) \frac{I}{N_{\text{mix}}} - \alpha E \\
\frac{dI}{dt} &= \alpha E - \gamma_I I - \sigma I\\
\frac{dH}{dt} &= \sigma I- \gamma_H H \\
\frac{dR}{dt} &= \gamma_I I + \gamma_H H
\end{align*}
$$

This model could be optionally parameterized with $\beta_{S} = \beta$, and $\beta_{V} = a \beta$ with $a \in (0, 1)$ to explicitly show that the rate at which vaccinated individuals acquire infection and transmit the disease is reduced when compared to unvaccinated individuals [@brauer_2012].

# Calibration Example

The general goal of this example is to see if we can fit the SHIVER model to COVID19 hospitilization data to see if we can estimate plausible tranmission rates for vaccinated and unvaccinated individuals.

We load the model specification from the model library.

```{r model_spec}
source("tmb.R")
#spec = mp_tmb_library("starter_models","shiver", package="macpan2")
```

## Calibration Scenario
Let's create a scenario for calibration. Suppose we have 3 months of daily COVID hospitalization data where some daily reports are missing.
```{r calibration_scenario}
expected_daily_reports = 90 # days
missed_reports = 10
actual_daily_reports = expected_daily_reports - missed_reports
```

Daily COVID19 hospitalization data for Ontario was obtained from the [Ontario Data Catalogue](https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/274b819c-5d69-4539-a4db-f2950794138c). The data contains daily hospitalization counts stratitfied by vaccination status and severity of hospitalization care. For simplicity, we aggregate all counts into one. We will use the first data point (August 10, 2021) as the initial condition for H, and the second data point (August 11, 2021) as day 1 of the scenario. To incorporate missed data reporting, we randomly remove `r missed_reports` records.

```{r observed_data}
set.seed(expected_daily_reports)
# Obtained from here:
# https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/274b819c-5d69-4539-a4db-f2950794138c
daily_hospitalizations = (read.csv(
  system.file(
      "starter_models"
    , "shiver"
    , "data"
    , "hospitalizations_ontario.csv"
    , package = "macpan2"
  )
  , row.names = NULL
  ) 
  |> rename(time = X_id)
  |> group_by(time)
  # assuming all hospital admissions (ICU or not) are part of H compartment
  |> summarize(value = sum(icu_unvac, icu_partial_vac, icu_full_vac,
                            hospitalnonicu_unvac, hospitalnonicu_partial_vac,
                            hospitalnonicu_full_vac))
  |> ungroup()
)

head(daily_hospitalizations)

reported_hospitalizations = (daily_hospitalizations
  # remove first record (Aug 10, this will be the initial condition for H)
  |> filter(time!=1) 
  # update all times to set Aug 11 to be day 1 of the scenario
  |> mutate(time=time-1) 
  |> head(expected_daily_reports)
  |> mutate(matrix="H")
  |> sample_n(actual_daily_reports)
)

```

## Deciding on Defaults

We want to check the model specification defaults to see what needs updating. 

```{r}
# We can view model spec default values here to see if we need to make any changes.
mp_default(spec)

```






We need to update the population size $N$ to the population of Ontario at the time. [Statistics Canada](https://doi.org/10.25318/1710000901-eng) estimates a population of approximately 14.8 million in the second quarter of 2021.

```{r N}
# N = population size
# ---------------------
N = 14.8e7
```

Vaccination data from [Public Health Ontario](https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool) can help us get an estimate for the per capita vaccination rate $\phi$. The average number of vaccines that can be administered per day was estimated by averaging weekly vaccination counts for the month preceding the scenario. We then convert `daily_vaccinations` to a rate by dividing by the initial population size. 

```{r phi}
# phi = vaccination rate
# ---------------------
# for the month of July 2021, number of individuals vaccinated per week in ontario
july_vax = c(
    135177
  , 124468
  , 111299
  , 100825
)
## average july weekly data to per day (4 weeks = 28 days)
daily_vaccinations = sum(july_vax)/28 # seems plausible
# 0.01% of population per day
phi = daily_vaccinations/N
```

We assume the average protection acquired from vaccination lasts 6 months (180 days), individuals spend on average 3.3 days in the exposed compartment, and 10% of infectious individuals on average require hospital care before recovery. 

```{r other_defaults}
# rho = waning vaccination
# ---------------------
rho = 1/180 # average protection lasts 180 days 

# alpha = exposure rate
# ---------------------
alpha = 1/3.3 #3.3 days in exposed class 

# sigma = hospitalization rate
# ---------------------
# 10% of infections result in hospitalizations
sigma = 1/10
```

The default recovery rates mean an individual takes 10 days on average to move from I to R, and 14 days on average to go from H to R. These rates seem plausible for this scenario. Note that the recovery class in this model, includes deaths from the hospitalization class. We could speculate that the recovery rate from H to R could be shorter than the recovery rate from I to R if hospitalized individuals are more likely to die for instance than recover (and given death on average happens quicker than recovery from infection). For now, let's stick with the 2 week recovery rate for H, given we assumed in the observed data above that all hospitalized individuals are in H regardless of severity.

We want to estimate the transmission parameters `beta_s` and `beta_v`. It's more challenging to interpret these state dependent parameters, but we would 
expect transmission involving vaccinated suceptibles would be less effective than unvaccinated transmission (beta_v < beta_s). These defaults meet this minimal assumption. Additionally, $R_{0,S} \approx \beta_s/\gamma_i = 2$ which seems reasonable.

We want to update the initial states with information we can find, [Public Health Ontario](https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool) provides us with daily estimates for $V(0)$ and $I(0)$ and $H(0)$ is taken from observed data. There is no data on initial exposure, however we know it is improbable that there are no exposed individuals initially. We will estimate this value in addition to transmission parameters when calibrating. Finally, since we are not interested in the dynamics of S or R, it makes sense to leave the default $R = 0$, even though we have prior knowledge that by the summer of 2021 $R > 0$. Individuals in R are removed
from the transmission dynamics, so initializing R to a non-zero value will only deplete the initial susceptible population by this non-zero value.

```{r initial conditions}
## Initial Conditions
# ---------------------

# V0 = initial V
# ---------------------
# We can get weekly Ontario vaccination data here,
# https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool
# If we divide the number of individuals vaccinated in the first week of August 2021 by 7
V0 = 71096/7


# I0 = initial I
# ---------------------
# We can also look at the number of weekly cases of COVID for this time period from here,
# https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool
# and divide by 7 as above to make it a daily estimate.
I0 = 1903/7

# H0 = initial H
# ---------------------
# Use the first observed data point (Aug 10, 2021) as initial H.
H0 = daily_hospitalizations |> filter(row_number()==1) |> select(value) |> pull()
```


## Simulating Dynamics

Before optimizing, we want to make sure the dynamics look reasonable, so we simulate from the calibrator object using `mp_trajectory`. The trajectories look mostly as expected. The sharp initial increase in E and decrease in I can be attributed to setting $E(0)=1$

```{r simulating_dynamics}

# state variables
states = c("S","H","I","V","E","R")

# set up calibrator
shiver_calibrator = mp_tmb_calibrator(
    spec = spec
  , data = reported_hospitalizations
  , traj = "H"
  # parameters we want to estimate (transmission rates)
  # we also want to estimate initial E
  , par = c("beta_v","beta_s","E") 
  , outputs = states
  # update defaults
  , default = list(
      N=N
    , V=V0
    , I=I0
    , H=H0
    , phi=phi
    , alpha=alpha
    , sigma=sigma
    , rho = rho
  ) 
)
# print to check
shiver_calibrator

# trajectory has 90 time steps (which is what we expect)
nrow(shiver_calibrator 
  |> mp_trajectory()
  |> filter(matrix=="H")
  |> select(time) 
  |> unique()
)
  

# which time steps are missing in observed data
(shiver_calibrator 
  |> mp_trajectory()
  |> filter(matrix=="H")
  |> anti_join(reported_hospitalizations, by="time")

) 

# before optimizing, do the dynamics look reasonable? 
(shiver_calibrator 
    |> mp_trajectory()
    |> ggplot(aes(time, value))
    + facet_wrap(vars(matrix), scales='free')
    + geom_line()
    + theme_bw()
)


```

## Estimating Parameters

We are now ready for the optimization step.
```{r estimates}
# optimize to estimate parameters
# this converges!
mp_optimize(shiver_calibrator)

# look at estimates with CI
# beta_s ~ 0.19(0.18,0.19) - good
# beta_v < beta_s but CI contains 0
# E0 ~ 148 std error 11 - good
mp_tmb_coef(shiver_calibrator, conf.int=TRUE)
```

We get a realistic estimate for `beta_s` at 0.18 with a small standard error and the estimated intial number of exposed individuals, 148, seems plausible with a standard error of 11. The estimate for $\beta_v$ however has a large standard error and the confidence interval contains zero indicating we are not learning about this parameter.

To check the fit we plot the observed data as well as the trajectories of all states.

```{r fit, echo=FALSE}
# how does the fit compare with observed data?
(shiver_calibrator 
    |> mp_trajectory_sd(conf.int=TRUE)
    |> filter(matrix == "H")
    |> ggplot(aes(time, value))
    + geom_line(aes(y=value), colour="red")
    + geom_ribbon(aes(ymin=conf.low,ymax=conf.high), fill="red",alpha=0.3)
    + geom_point(data=reported_hospitalizations, aes(time,value))
  )


# does the fit create unrealistic dynamics?
(shiver_calibrator 
    |> mp_trajectory()
    |> ggplot(aes(time, value))
    + facet_wrap(vars(matrix), scales='free')
    + geom_line()
)


```

The simulated hospitalization trajectory fits the data well, and the other trajectories look as expected.

## Re-parameterizing and introducing transformations

For better interpretability we can re-parameterize the model with one transmission rate, `beta`, and a proportion, `a` in (0,1), representing the reduced transmission rate for vaccinated people. We also wish to parameterize `{I0, E0}` to `{I0, E0/I0}` to de-correlate `I0` and `E0`.

We use the log transformation for `beta` and `E0/I0` because both quantities take on only positive values. We use the logistic transformation for the proportion `a` to constrain it to the domain (0,1).

We define a new model specification object with these changes, and set the default values for all three parameters to be small values (0.01) because they all have a lower bound of 0. 

```{r reparameterization}
# Create a new model specification with these changes:
#
# - update the before step to transform "new" parameters
reparameterized_spec = mp_tmb_insert(spec
     , phase = "before"
     , at=1L
     , expressions = list(
         E ~ exp(log_E_I_ratio) * I
       , beta ~ exp(log_beta)
       , a ~ 1/(1+exp(-logit_a))
     )
     , default = list(
         logit_a = qlogis(1e-2)
       , log_beta = log(1e-2)
       , log_E_I_ratio = log(2) 
     )
)

# - overwrite existing exposure terms with new ones
reparameterized_spec = mp_tmb_update(reparameterized_spec
    , phase = "during"
     # exposure expressions start at step 4 in the during phase
    , at=4L
    , expressions = list(
        mp_per_capita_flow("S", "E", unvaccinated_exposure ~ I * beta/N_mix)
      , mp_per_capita_flow("V", "E", vaccinated_exposure ~  I * beta * a/N_mix)
    )
)


# all changes have been made
print(reparameterized_spec)
```

Next we calibrate and specify the transformed parameters to estimate.

```{r }
# let's calibrate
shiver_calibrator = mp_tmb_calibrator(
    spec = reparameterized_spec
  , data = reported_hospitalizations
  , traj = "H"
  # now we want to estimate the transformed parameters
  , par = c("log_beta","logit_a","log_E_I_ratio")
  , outputs = states
  , default = list(
      N=N
    , V=V0
    , I=I0
    , H=H0
    , phi=phi
    , alpha=alpha
    , sigma=sigma
    , rho = rho
  ) 
)


# optimize to estimate transmission parameters
# converges
mp_optimize(shiver_calibrator)
```

```{r}
# looking at coefficients and CIs
# we need to back transform to interpret
cc <- (mp_tmb_coef(shiver_calibrator, conf.int=TRUE)
       |> backtrans()
)
cc 

```
The estimates for our parameters are close to the the previous estimates, which makes sense because we only re-parameterized instead of making changes to the model specification. The ratio of initial exposed to initial infecteds is `r cc |> filter(mat=="E_I_ratio") |> select(estimate) |> pull() |> round(2)`, given we specified the initial I as `r round(I0,0)`, the estimated initial number of exposed is approximately `r cc |> filter(mat=="E_I_ratio") |> select(estimate) |> pull() %>% "*"(I0) |>  round(0)`  A reduction in transmission by `r cc |> filter(mat=="a") |> select(estimate) |> pull() |> round(2)` percent could be plausible but again the CI contains 0. 


## Runge-Kutta 4

By default, `macpan2` uses the first order Euler method for solving ODEs. We might be able to improve our estimates if we use a higher order method such as Runge-Kutta 4. Provided our model is specified appropriately (it is), we can easily perform Runge-Kutta 4 with `mp_rk4()`. However, in this scenario we get optimization warnings that lead to extreme estimates and no confidence intervals. 

```{r rk4}
# let's calibrate
shiver_calibrator_rk4 = mp_tmb_calibrator(
    spec = reparameterized_spec |> mp_rk4()
  , data = reported_hospitalizations
  , traj = "H"
  , par = c("log_beta","logit_a","log_E_I_ratio")
  , outputs = states
  # defaults are specified in model spec alreadt
)

# optimize
# converges with warning
mp_optimize(shiver_calibrator_rk4)

# looking at coefficients and CIs
rk4_coef <- (mp_tmb_coef(shiver_calibrator_rk4, conf.int=TRUE)
       |> backtrans()
)

rk4_coef
# rk4 doesn't help us learn more about a
# let's try adding more data

```

## Fitting to Multiple Trajectories

If we include more observed data, can we get an estimate for the proportion `a`? 

We obtain COVID case count data from the [Ontario Data Catalogue](https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/eed63cf2-83dd-4598-b337-b288c0a89a16) for an experiment to fit the SHIVER model to both hospitialzation and incidence data.

```{r, include=FALSE}
# COVID19 case data for Ontario
# It makes sense to assume these case counts are incidence (# number of new 
# cases each day) instead of prevalence (# all active cases each day)
daily_cases = (read.csv(
    system.file(
        "starter_models"
      , "shiver"
      , "data"
      , "cases_ontario.csv"
      , package = "macpan2"
    )
  , row.names = NULL
  )
  |> rename(time = X_id)
  |> rowwise()
  # aggregate all case counts (by vax status) into one
  |> mutate(value = sum(c_across(starts_with("covid19_cases")),na.rm = TRUE))
)


reported_cases = (daily_cases
     |> filter(time!=1 & time!=2)  # remove first two records (Aug 9, Aug10 = I0)
     |> mutate(time=time-2)        # update all times to set Aug 11 to be day 1 of the scenario
     |> head(expected_daily_reports)
     |> mutate(matrix="infection") # the infection variable is incidence
     |> select(c("time","value","matrix"))
)

# use Aug 10 as initial I
# Note: this case data is from a different source than above, but I0 and I0_new
# don't differ much
I0_new = daily_cases |> filter(time==2) |> select(value) |> pull()

# check defaults again
mp_default(reparameterized_spec)

# calibrate
shiver_calibrator = mp_tmb_calibrator(
    spec = reparameterized_spec |> mp_rk4()
    # row bind both observed data
  , data = rbind(reported_hospitalizations, reported_cases)
    # fit both trajectories
  , traj = c("H","infection")
  , par = c("log_beta","logit_a","log_E_I_ratio")
  , outputs=c(states, "infection")
)

# optimize to estimate transmission parameters
# converges 
mp_optimize(shiver_calibrator)


```


```{r}
(mp_tmb_coef(shiver_calibrator, conf.int=TRUE)
    |> backtrans()
)
```

We were not able to estimate `a` again, note the confidence interval is the entire domain of possible values for a. 

Looking at the fit to the observed data, we can clearly see adding more noisy data leads to major fitting problems.

```{r echo=FALSE}
# how does data look with these parameters
# not good!
(shiver_calibrator 
   |> mp_trajectory_sd(conf.int=TRUE)
   |> filter(matrix %in% c("H","infection"))
   |> ggplot(aes(time, value))
   + geom_line(aes(y=value), colour="red")
   + geom_ribbon(aes(ymin=conf.low,ymax=conf.high), fill="red",alpha=0.3)
   + geom_point(data=rbind(reported_hospitalizations, reported_cases), aes(time,value))
   + facet_wrap(vars(matrix), scales='free')
  )


```



## Parameter identifiability

As a first step in calibration, we advise specifying a default for the parameter of interest, simulate from the model and then calibrate to this simulated data to see if we can recover the parameter default. If we are unable to recover the default, this could mean there are identifiability issues with the model and parameter.

Here we specify a ground truth for `a` and `beta`, and simulate data from the model using these default values. We perform two calibrations, first we estimate `a` with a fixed `beta`, and then we do the reverse.

```{r}

# set true values
true_a = 0.2
true_beta = 0.3 


# simulate fake data
simulated_data = (reparameterized_spec
  |> mp_simulator(
      time_steps = expected_daily_reports
    , outputs=states
    , default=list(
        logit_a=qlogis(true_a)
      , log_beta=log(true_beta)
    )
    )
  |> mp_trajectory()
  # add some noise
  |> mutate(value = rpois(n(),value))
  |> select(-c(row,col))
)



## fix beta estimate a
fixed_beta = mp_tmb_calibrator(
    spec = reparameterized_spec |> mp_rk4()
  , data = simulated_data
  , traj = states
  , par = c("logit_a")
  , outputs=states
)
# converges, but not getting estimate for `a`
mp_optimize(fixed_beta)
(mp_tmb_coef(fixed_beta, conf.int=TRUE)
  |> backtrans()
)
## fix a estimate beta
fixed_a = mp_tmb_calibrator(
  spec = reparameterized_spec |> mp_rk4()
  , data = simulated_data
  , traj = states
  , par = c("log_beta")
  , outputs=states
)
# converges and recovering true beta
mp_optimize(fixed_a)
(mp_tmb_coef(fixed_a, conf.int=TRUE)
  |> backtrans()
)


```

We are able to recover the transmission rate `beta` but not the proportion `a`, suggesting model indentifiability issues with this parameter.

# Vaccine Constraints and Variable Vaccination Rate 

We can implement vaccine constraints by adding more model complexity. Resource limitations create an upper bound on the number of vaccines that can be administered to susceptibles per time step (in this case per day). There is also the constraint that we can only vaccinate, at most, the current number of susceptibles i.e. the vaccination rate can be at most 1. These constraints naturally lead us to consider a variable vaccination rate $\phi(S(t))$, instead of vaccinating a fixed proportion $\phi$ per time step. 

We choose a sigmoidal function for $\phi(S(t))$ because these curves are increasing and asymptotic. The Hill function, $H(x) = x^n/(K_{half}^n + x^n)$, is one such curve that goes through the origin which is convenient for our case because when we have zero susceptibles our vaccination rate should be zero. The Hill function is parameterized by the Hill coefficient $n>0$ and $K_{half}$ is the value of the dependent variable $x$ at the half maximum.

<!-- example Hill plot -->
```{r hill_parameters}
# Hill parameters
# asymptote
A = daily_vaccinations
# value of S at half maximum, seems like a sensible choice
half_max = A/2
# Hill coefficient - chosen randomly, but higher n creates steeper slopes
n = 3

```

```{r hill_plot, echo=FALSE}

# specify a sequence of S
S = seq(0,3*daily_vaccinations,length.out=300)

# test out hill function
hill_data = simple_sims(
    iteration_exprs = list(
      hill_fn ~ (A * (S^n))/((half_max)^n + (S^n))
    )
  , time_steps = 1L
  , mats = list(
      hill_fn = 0
    , A = A
    , S = S
    , n = n
    , half_max = half_max
    , N_mix = N
  )
)

# this is right
(hill_data
   |> filter(matrix %in% c("hill_fn"))
   |> cbind(S)
   |> ggplot(aes(S, value))
   + geom_point(alpha=0.3)
   + geom_hline(yintercept = A,col="blue")
   + geom_text(x=50,y=A-300,label="A", col="blue")
   # + geom_text(x=500,y=2500,label=expression(phi = S),col="red")
   # + geom_abline(a=0,b=1, col="red")
   + ggtitle("Variable vaccination rate using the Hill function")
   + theme_bw()
   + ylab("H(S)")

  )


```

To restrict the vaccination rate to be at most 1, we use the method in @gharouni_2022 Appendix 5 to modify the output of the hill function. 
```{r gharouni_parameters}
# Gharouni parameters
max_rate = 1 #tau parameter in Gharouni

```

We then use this rate as our vaccination rate,  $\phi(S(t)) =  \frac{\textrm{max}_{rate} Hill(S)}{\textrm{max}_{rate} S + Hill(S)}$ 
```{r gharouni_rate}

# test out Gharouni (2022) rate
hill_data = simple_sims(
    iteration_exprs = list(
        hill_fn ~ (A * (S^n))/((half_max)^n + (S^n))
      , vaccination ~ ((max_rate * hill_fn)/((max_rate * S) + (hill_fn)))
      , flow ~ vaccination * S
      , S ~ S - flow
    )
  , time_steps = 100
  , mats = list(
      hill_fn = 0
    , A = A
    , S = A*10
    , n = n
    , half_max = half_max
    , max_rate = max_rate
    , N_mix = N
    , vaccination = 0
    , flow = 0
  )
)

# rate is always less than 1 and is larger for smaller S, good
# not sure why there is a peak (maybe part of hill coeficient choice)
(hill_data 
   |> filter(matrix %in% c("vaccination"))
   |> ggplot(aes(time,value))
   + geom_point(alpha=0.3)
   + geom_hline(yintercept = 1)
   + geom_text(x=5,y=0.95,label="max_rate")
   + theme_bw()
)


```


```{r var_vax}
# update model spec with vaccine supply function phi(S) in place of constant phi
spec_vaccine_rate = mp_tmb_update(
    reparameterized_spec
  , phase = "during"
  , at = 2L
  , expression = list(
      hill_fn ~ (A * (S^n))/((half_max)^n + (S^n)),
      mp_per_capita_flow(
        from = "S"
      , to = "V"
      , rate = vaccination ~ ((max_rate * hill_fn)/((max_rate * S) + (hill_fn)))
  )
  )
  , default = list(
      A = A
    , half_max = half_max
    , n = n
    , max_rate = max_rate
    , vaccination = 0
  )
)
print(spec_vaccine_rate)


# simulate from model and plot hill_fn
# update R to skip through dynamics to see what happens when we run out of S
simulated_data = (
    spec_vaccine_rate
  |> mp_simulator(time_steps = 500,outputs=c(states,"vaccination"), default = list(R=0.95*N))
  |> mp_trajectory()
  ##|> mutate(value = rpois(n(),value))
)

# things look good

(simulated_data 
   |> filter(matrix %in% c("vaccination"))
   |> ggplot(aes(time, value))
   + geom_point(alpha=0.3)
   + ylab("daily vaccinations")
   + theme_bw()
)

```

Let's check the vaccination rate

```{r vax_rate, echo=FALSE}
# let's check variable vaccination rate
sim_data2 = (simulated_data
             %>% filter(matrix %in% c("S","vaccination"))
             %>% pivot_wider(names_from = matrix, values_from = value)
             %>% mutate(value = vaccination/S))
# wonder if peak causes issues, 
# looks right thought, always <1, its closest to 1 when S is small

(sim_data2 
   |> ggplot(aes(time, value))
   + geom_point(alpha=0.3)
   + scale_y_log10()
   + ylab("vaccination rate")
   + theme_bw()
)

```

The vaccination rate is at most 1, and it's greatest when susceptibles we have smaller numbers of susceptibles.
We can now try calibrating 


```{r}


# try calibratings ....
# can we estimate fit to real data and see if we can get similar estimates for
# beta and E_I ratio
cal_vaccine_supply = mp_tmb_calibrator(
  spec = spec_vaccine_rate |> mp_rk4()
  , data = reported_hospitalizations
  , traj = "H"
  , par = c("log_beta","log_E_I_ratio")
  , outputs=states
)
cal_vaccine_supply = mp_tmb_calibrator(
  spec = spec_vaccine_rate |> mp_rk4()
  , data = reported_hospitalizations
  , traj = "H"
  , par = c("vaccination")
  , outputs=states
)
# # converges and get similar estimates when phi is constant
# # makes sense I think because these parameters don't relate to vaccination rate
mp_optimize(cal_vaccine_supply)

(mp_tmb_coef(cal_vaccine_supply, conf.int=TRUE)
  |> backtrans()
)



```


# Model Specification

This model has been specified in the `shiver` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/shiver/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). 

# References

