---
title: "SHIVER = SEIR + H + V"
index_entry: "A modified SEIR model with Hospitalization and Vaccination"
bibliography: ../../references.bib
link-citations: TRUE
author: Jennifer Freeman, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "./figures/"
)
system.file("utils", "round-coef.R", package = "macpan2") |> source()
```


This model builds on the basic SEIR model, with two additional compartments for vaccination and hospitalizations. 

Vaccines are typically subject to resource constraints and distribution strategies might prioritize vaccinations for specific subpopulations, such as immunocompromised people, to reduce bad outcomes. We model this with a flow of susceptibles entering the vaccination class. This flow could be a fixed rate, i.e. a constant proportion of the population receives a vaccine each time step, but instead we wish to capture a more realistic vaccination rate by allowing it to vary (see [Variable Vaccination Rate](#variable-vaccination-rate)). 

Following vaccination, most individuals return to the susceptible class when the acquired immune response wears off, called *vaccine waning*.

In reality, vaccinations are only partly effective in developing a protective host immune response. This *vaccine failure* includes both vaccine-related inadequacies and host-related reasons like health status and genetic factors [@wiedermann_2016]. Vaccine failure is modelled as flow from vaccination to the exposed class.

The SHIVER model assumes all individuals can be exposed to the disease, but transmission rates depend on vaccination status. Additionally, individuals with severe infections are hospitalized and assumed to be isolated, before recovering from the disease. Hospital isolation means this portion of infectious individuals no longer contribute to the transmission dynamics. 

# Packages Used and Settings

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```

To keep the optimizer from printing too much in this article, we set the `macpan2_verbose` option to `FALSE`.

```{r options}
options(macpan2_verbose = FALSE)
```

# Model Specification

This model has been specified in the `shiver` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/shiver/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
spec = mp_tmb_library(
    "starter_models"
  , "shiver"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagrams using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/shiver/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 2, fig.width = 5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_grid(spec
  , east = "(infection|progression|recovery)$"
  , north = "waning$"
  , south = "(hospitalizations|vaccination)$"
)
plot_flow_diagram(layout)
```

# States

| variable | description                        |
| -------- | ---------------------------------- |
| S        | Number of susceptible individuals  |
| H        | Number of hospitalized individuals |
| I        | Number of infectious individuals   |
| V        | Number of vaccinated individuals   |
| E        | Number of exposed individuals      |
| R        | Number of recovered individuals    |

The size of the total population is,  $N = S + H + I  + V + E + R$, and the disease spreads through homogeneous mixing of the subpopulation $N_{\text{mix}}=N -H$.

# Parameters

| variable   | description                                                                                         |
| ---------- | --------------------------------------------------------------------------------------------------- |
| $\phi$     | per capita vaccination rate of susceptibles                                                         |
| $\rho$     | per capita vaccine waning rate                                                                      |
| $\beta_S$  | per capita transmission rate for susceptibles (in $N_{\text{mix}}$ population)                      |
| $\beta_V$  | per capita transmission rate for vaccinated individuals (in $N_{\text{mix}}$ population)            |
| $\alpha$   | per capita infection rate (average time spent in compartment $E$ is $1/\alpha$)                     |
| $\gamma_I$ | per capita recovery rate for infected  individuals                                                  |
| $\gamma_H$ | per capita recovery rate for hospitalized individuals                                               |
| $\sigma$   | per capita rate at which infected individuals develop severe infections and require hospitalization |



# Variable Vaccination Rate

We can implement vaccine constraints by adding more model complexity. Resource limitations create an upper bound on the number of vaccines that can be administered to susceptibles per time step. There is also the constraint that we can only vaccinate, at most, the current number of susceptibles i.e. the vaccination rate can be at most 1. These constraints naturally lead us to consider a variable vaccination rate $\phi(S(t))$, instead of vaccinating a fixed proportion $\phi > 0$ per time step. 

There are many choices for the function $\phi(S(t))$. We choose a sigmoidal function because these curves are increasing and asymptotic. The *Michaelis-Menten* function, $f(x) = ax/(b + x)$, is one such curve that passes through the origin, which is convenient for our case because when we have zero susceptibles our vaccination rate should be zero [@bolker2008]. The slope through the origin for $f(x)$ is $a/b$ and $f(x)$ approaches $a$ as $x \rightarrow \infty$ [@bolker2008].

We interpret $a$ as the maximum number of vaccinations that can be administered per time step, and we set $b=a$ to fix the vaccination rate to be at most 1.

```{r Michaelis-Menten_param}
# asymptote
a = 1000
# force slope to be one at the origin
b = a
```


Suppose we can vaccinate at most `r a` individuals per time step, then the Michaelis-Menten function will approach this asymptote with increasing $S(t)$. The choice of $b$ restricts the slope of the curve to be below the line $y=x$

```{r Michaelis-Menten_fn, echo=FALSE}

# specify a sequence of S
S = seq(0,3*a,length.out=300)

# simulate Michaelis-Menten and vaccination rate for each S
varvax = simple_sims(
    iteration_exprs = list(
        mm_fn ~ (a * S)/(b + S)
      , vaccination ~ mm_fn/S
    )
  , time_steps = 1L
  , mats = list(
      mm_fn = 0
    , a = a
    , b = b
    , S = S
    , vaccination = 0
    , flow = 0
  )
)
# Michaelis-Menten function
(varvax 
   |> filter(matrix %in% c("mm_fn"))
   |> cbind(S)
   |> ggplot(aes(S, value))
   + geom_point(alpha=0.3)
   + geom_hline(yintercept = a,col="red")
   + geom_text(x=10,y=a-30,label="a", col="red")
   + geom_abline(intercept=0,slope=1,col="blue")
   + geom_text(x=a+30,y=a-125,label="y = x",col="blue")
   + theme_bw()
   + ggtitle("Michaelis-Menten function")
   + ylab(expression(f(S(t))))
   + xlab(expression(S(t)))
   + theme_bw()
)


```

To convert the Michaelis-Menten curve above to a rate, we divide by $S(t)$.

```{r var_vax, echo=FALSE, warning=FALSE}
# vaccination rate phi(S(t))
(varvax 
   |> filter(matrix %in% c("vaccination"))
   |> cbind(S)
   |> ggplot(aes(S, value))
   + geom_point(alpha=0.3)
   + geom_hline(yintercept = 1,col="red")
   + geom_text(x = 1500, y = 0.98, label = "maximum rate", col = "red")
   + theme_bw()
   + ggtitle("Vaccination Rate")
   + ylab(expression(phi(S(t))))
   + xlab(expression(S(t)))
   + theme_bw()
)

```

# Dynamics

$$
\begin{align*}
\frac{dS}{dt} &= -\beta_{S} S\frac{I}{N_{\text{mix}}} - \phi(S) S + \rho V\\
\frac{dV}{dt} &=  \phi(S) S - \rho V - \beta_{V} V\frac{I}{N_{\text{mix}}} \\
\frac{dE}{dt} &= (\beta_{S} S + \beta_{V} V) \frac{I}{N_{\text{mix}}} - \alpha E \\
\frac{dI}{dt} &= \alpha E - \gamma_I I - \sigma I\\
\frac{dH}{dt} &= \sigma I- \gamma_H H \\
\frac{dR}{dt} &= \gamma_I I + \gamma_H H
\end{align*}
$$

This model could be optionally parameterized with $\beta_{S} = \beta$, and $\beta_{V} = p \beta$ with $p \in (0, 1)$ to explicitly show that the rate at which vaccinated individuals acquire infection and transmit the disease is reduced when compared to unvaccinated individuals [@brauer_2012].

# Calibration Example

The general goal of this example is to see if we can fit the SHIVER model to COVID19 hospitilization data to estimate plausible transmission rates for vaccinated and unvaccinated individuals.

## Calibration Scenario

Let's create a scenario for calibration. Suppose we have 3 months of daily COVID19 hospitalization data where some daily reports are missing.
```{r calibration_scenario}
expected_daily_reports = 90 # days
missed_reports = 10
actual_daily_reports = expected_daily_reports - missed_reports
```

Daily COVID19 hospitalization data for Ontario was obtained from the [Ontario Data Catalogue](https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/274b819c-5d69-4539-a4db-f2950794138c). The data contains daily hospitalization counts stratified by vaccination status and severity of hospitalization care. For simplicity, we aggregate all counts into one. We will use the first data point (August 10, 2021) as the initial condition for H, and the second data point (August 11, 2021) as day 1 of the scenario. To incorporate missed data reporting, we randomly remove `r missed_reports` records.

```{r observed_data}
set.seed(expected_daily_reports)
# Obtained from here:
# https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/274b819c-5d69-4539-a4db-f2950794138c
daily_hospitalizations = (read.csv(
  system.file(
      "starter_models"
    , "shiver"
    , "data"
    , "hospitalizations_ontario.csv"
    , package = "macpan2"
  )
  , row.names = NULL
  ) 
  |> rename(time = X_id)
  |> group_by(time)
  # assuming all hospital admissions (ICU or not) are part of H compartment
  |> summarize(value = sum(icu_unvac, icu_partial_vac, icu_full_vac,
                            hospitalnonicu_unvac, hospitalnonicu_partial_vac,
                            hospitalnonicu_full_vac))
  |> ungroup()
)

head(daily_hospitalizations)

reported_hospitalizations = (daily_hospitalizations
  # remove first record (Aug 10, this will be the initial condition for H)
  |> filter(time!=1) 
  # update all times to set Aug 11 to be day 1 of the scenario
  |> mutate(time=time-1) 
  |> head(expected_daily_reports)
  |> mutate(matrix="H")
  |> sample_n(actual_daily_reports)
  |> arrange(time)
)

```

## Deciding on Defaults

We want to check the model specification defaults to see what needs updating. 

```{r mp_default}
# We can view model spec default values here to see if we need to make any
# changes
mp_default(spec)

```

We need to update the population size $N$ to the population of Ontario at the time. [Statistics Canada](https://doi.org/10.25318/1710000901-eng) estimates a population of approximately 14.8 million in the second quarter of 2021.

```{r N}
# N = population size
# ---------------------
N = 14.8e7
```

Vaccination data from [Public Health Ontario](https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool) can help us get an estimate for the maximum number of vaccines that can be administered per day. We average weekly vaccination counts for the month preceding the scenario, and use this as our estimate for the Michaelis-Menten asymptote parameter $a$. 

```{r phi}
# phi = vaccination rate
# ---------------------
# for the month of July 2021, number of individuals vaccinated per week in ontario
july_vax = c(
    135177
  , 124468
  , 111299
  , 100825
)
## average july weekly data to per day (4 weeks = 28 days)
# use this as the maximum number of daily vaccinations
a = sum(july_vax)/28 # seems plausible
```

We assume the average protection acquired from vaccination lasts 6 months (180 days), individuals spend on average 3.3 days in the exposed compartment, and 10% of infectious individuals on average require hospital care before recovery. 

```{r other_defaults}
# rho = waning vaccination
# ---------------------
rho = 1/180 # average protection lasts 180 days 

# alpha = exposure rate
# ---------------------
alpha = 1/3.3 #3.3 days in exposed class 

# sigma = hospitalization rate
# ---------------------
# 10% of infections result in hospitalizations
sigma = 1/10
```

The default recovery rates mean an individual takes 10 days on average to move from I to R, and 14 days on average to go from H to R. These rates seem plausible for this scenario. Note that the recovery class in this model, also called the removed compartment, includes deaths from the hospitalization class. We could speculate that the recovery rate from H to R could be shorter than the recovery rate from I to R if hospitalized individuals are more likely to die for instance than recover (and given death on average happens quicker than recovery from infection). For now, let's stick with the 2 week recovery rate for H, given we assumed in the observed data above that all hospitalized individuals are in H regardless of severity.

We want to estimate the transmission parameters `beta_s` and `beta_v`. It's more challenging to interpret these state dependent parameters, but we would expect transmission involving vaccinated suceptibles would be less effective than unvaccinated transmission (`beta_v` < `beta_s`). These defaults meet this minimal assumption. Additionally, $R_{0,S} \approx \beta_s/\gamma_i = 2$ which seems reasonable.

We want to update the initial states with data we can find. [Public Health Ontario](https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool) provides us with daily estimates for $V(0)$ and $I(0)$ and $H(0)$. There is no data on initial exposure, however we know it is improbable that there are no exposed individuals initially. We will estimate this value in addition to transmission parameters when calibrating. Finally, since we are not interested in the dynamics of S or R, it makes sense to leave the default $R = 0$, even though we have prior knowledge that by the summer of 2021 $R > 0$. Individuals in $R$ are removed from the transmission dynamics, so initializing $R$ to a non-zero value will only deplete the initial susceptible population by this value.

```{r initial_conditions}
## Initial Conditions
# ---------------------

# V0 = initial V
# ---------------------
# We can get weekly Ontario vaccination data here,
# https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool
# If we divide the number of individuals vaccinated in the first week of August 2021 by 7
V0 = 71096/7


# I0 = initial I
# ---------------------
# We can also look at the number of weekly cases of COVID for this time period from here,
# https://www.publichealthontario.ca/en/Data-and-Analysis/Infectious-Disease/Respiratory-Virus-Tool
# and divide by 7 as above to make it a daily estimate. We should also multiply by something
# to account for under-reporting ... say 10.
I0 = 10 * 1903/7

# H0 = initial H
# ---------------------
# Use the first observed data point (Aug 10, 2021) as initial H.
H0 = daily_hospitalizations |> filter(row_number() == 1) |> select(value) |> pull()
```

Now we update the specification to include these defaults.
```{r defaults}
spec = mp_tmb_update(spec
  , default = list(
      N = N
    , V = V0
    , I = I0
    , H = H0
    , a = a
    , b = a
    , alpha = alpha
    , sigma = sigma
    , rho = rho
  ) 
)
```


## Simulating Dynamics

Before optimizing, we want to make sure the dynamics look reasonable, so we simulate from the calibrator object using `mp_trajectory`. The trajectories look mostly as expected. The sharp initial increase in $E$ and decrease in $I$ might be attributed to setting $E(0)=1$

```{r simulating_dynamics}
# set up calibrator
shiver_calibrator = mp_tmb_calibrator(
    spec = spec
  , data = reported_hospitalizations
  , traj = "H"
  # parameters we want to estimate (transmission rates)
  # we also want to estimate initial E
  , par = c("beta_v","beta_s","E", "sigma", "gamma_h") 
  , outputs = mp_state_vars(spec)
)
# print to check
print(shiver_calibrator)

# trajectory has 90 time steps (which is what we expect)
nrow(shiver_calibrator 
  |> mp_trajectory()
  |> filter(matrix == "H")
  |> select(time) 
  |> unique()
)
  

# which time steps are missing in observed data
(shiver_calibrator 
  |> mp_trajectory()
  |> filter(matrix == "H")
  |> anti_join(reported_hospitalizations, by = "time")

) 

# before optimizing, do the dynamics look reasonable? 
(shiver_calibrator 
    |> mp_trajectory()
    |> mutate(state = factor(matrix, levels = mp_state_vars(spec)))
    |> ggplot(aes(time, value))
    + facet_wrap(vars(state), scales = 'free')
    + geom_line()
    + theme_bw()
)
```

## Estimating Parameters

We are now ready for the optimization step.
```{r estimates}
# optimize to estimate parameters.
# optimizer converged (note convergence = 0)
mp_optimize(shiver_calibrator)

# look at estimates with CI
est_coef = mp_tmb_coef(shiver_calibrator, conf.int=TRUE) |> round_coef_tab()
est_coef
```

We get an unrealistically low estimate for `beta_s` at `r est_coef %>% filter(mat=="beta_s") %>% select(estimate) %>% round(2) %>% pull()` with a small standard error and the estimated initial number of exposed individuals, `r est_coef %>% filter(mat=="E") %>% select(estimate) %>% round(0) %>% pull()`, seems plausible with a standard error of `r est_coef %>% filter(mat=="E") %>% select(std.error) %>% round(0) %>% pull()`. The estimate for $\beta_v$ however has a large standard error and the point estimate is too high, indicating that vaccination leads to higher transmission.

What's worse is the fit is good, even though the parameter estimates do not make sense. To check the fit we plot the observed data as well as the trajectories of all states.

```{r fit, echo=FALSE}
# how does the fit compare with observed data?
(shiver_calibrator 
  |> mp_trajectory_sd(conf.int = TRUE)
  |> ggplot(aes(time, value))
  + facet_wrap(~matrix, scales = "free")
  + geom_line(aes(y=value), colour = "red")
  + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "red", alpha = 0.3)
  + geom_point(data = reported_hospitalizations, aes(time, value))
  + ylim(c(0, NA))
  + theme_bw()
)
```

The simulated hospitalization trajectory fits the data well.

## Re-parameterizing and Introducing Transformations

Before addressing this issue with unrealistic parameter estimates, we re-parameterize ror better interpretability. We model with one transmission rate, `beta`, and a proportion, `p` in (0,1), representing the reduced transmission rate for vaccinated people. We also wish to parameterize `{I0, E0}` to `{I0, E0/I0}` to de-correlate `I0` and `E0`.

We use the log transformation for `beta` and `E0/I0` because both quantities take on only positive values. We use the logistic transformation for the proportion `p` to constrain it to the domain (0,1).

We define a new model specification object with these changes, and set the default values for all three parameters to be small values (0.01) because they all have a lower bound of 0.

```{r reparameterization}
reparameterized_spec = (spec
  |> mp_tmb_insert(phase = "before"
    , expressions = list(E ~ E_I_ratio * I)
    , default = list(E_I_ratio = 1e-2)
  )
  |> mp_tmb_update(phase = "during", at = 4L
    , expressions = list(
        mp_per_capita_flow("S", "E", unvaccinated_infection ~ I * beta/N_mix)
      , mp_per_capita_flow("V", "E", vaccinated_infection ~  I * beta * p/N_mix)
    )
    , default = list(beta = 1e-2, p = 1e-2)
  )
)
print(reparameterized_spec)
```

Next we calibrate and specify the parameters to estimate. Note that by prepending `log_` or `logit_` in front of parameter names we are asserting that we should put priors and optimize on these transformed scales.

```{r reparam_calib}
prior_distributions = list(
      log_beta = mp_uniform()
    , log_E_I_ratio = mp_uniform()
    , logit_p = mp_normal(qlogis(1/4), 8)
    , sigma = mp_uniform()
    , gamma_h = mp_uniform()
)
shiver_calibrator = mp_tmb_calibrator(
    spec = reparameterized_spec
  , data = reported_hospitalizations
  , traj = "H"
  , par = prior_distributions
)

# optimize to estimate transmission parameters
# converges with warnings
mp_optimize(shiver_calibrator)
```

Note here that we have to put a normal prior on the logit-transformed vaccine-efficacy, `p`, parameter, albeit with a very large standard deviation. If we do not put this prior the optimization fails. The posterior for this parameter will be effectively identical however to the prior, suggesting that there is very little information in the data about vaccine efficacy in this model, as we see in the following coefficient table.

```{r reparam_estimates, echo=FALSE}
# looking at coefficients and CIs
# we need to back transform to interpret
cc <- mp_tmb_coef(shiver_calibrator, conf.int = TRUE) |> round_coef_tab()
print(cc)
```
This re-parameterization is beginning to help make a more reasonable model. Our prior vaccine transmission reduction parameter, `p`, is restricted to `0-1` and this ensures that vaccination will reduce transmission. On the other hand, although a reduction in transmission by `r cc |> filter(mat=="p") |> mutate(x = estimate * 100) |> select(x) |> pull() |> round(2)` percent could be plausible but the confidence interval is effectively all possible values for `p`. The ratio of initial exposed to initial number of infected is `r cc |> filter(mat=="E_I_ratio") |> select(estimate) |> pull() |> round(2)`. Given we specified the initial $I$ as `r round(I0,0)`, the estimated initial number of exposed is approximately `r cc |> filter(mat=="E_I_ratio") |> select(estimate) |> pull() %>% "*"(I0) |>  round(0)`. The biggest issue with these estimates though is that the transmission rate, `beta`, for unvaccinated people is much too low (we expect something more like `0.2`).

The fit is very similar to the model without a reparameterization, but with different inferences for the variables that are not fitted. In particular, the reparameterized model has a saturating function for the number of recovered individuals and the number of infectious individuals goes to zero, which is not at all realistic.

```{r repar_fit}
(shiver_calibrator 
  |> mp_forecaster(0, outputs = mp_state_vars(spec))
  |> mp_trajectory_sd(conf.int = TRUE)
  |> ggplot(aes(time, value))
  + facet_wrap(~matrix, scales = "free")
  + geom_line(aes(y=value), colour = "red")
  + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "red", alpha = 0.3)
  + geom_point(data = reported_hospitalizations, aes(time, value))
  + ylim(c(0, NA))
  + theme_bw()
)
```


## Runge-Kutta 4

By default, `macpan2` uses the first order Euler method for solving ODEs. We might be able to improve our estimates if we use a higher order method such as Runge-Kutta 4. Provided our model is specified with the appropriate flow rates between compartments (it is, see `?mp_per_capita_flow`), we can easily perform Runge-Kutta 4 with `mp_rk4()`. 
```{r rk4}
# let's calibrate
shiver_calibrator_rk4 = mp_tmb_calibrator(
    spec = reparameterized_spec |> mp_rk4()
  , data = reported_hospitalizations
  , traj = "H"
  , par = prior_distributions
  , outputs = mp_state_vars(spec)
)

# optimize
# converges with warning
mp_optimize(shiver_calibrator_rk4)

# looking at coefficients and CIs
rk4_coef <- mp_tmb_coef(shiver_calibrator_rk4, conf.int = TRUE) |> round_coef_tab()

print(rk4_coef)
# rk4 doesn't help us learn more about p
# let's try adding more data

```

In this scenario, Runge-Kutta 4 did not improve our parameter estimates.

## Fitting to Multiple Trajectories

If we include more observed data, can we get a more believable fit and parameter estimates?

We obtain COVID case count data from the [Ontario Data Catalogue](https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/eed63cf2-83dd-4598-b337-b288c0a89a16) for an experiment to fit the SHIVER model to both hospitalization and incidence data.

```{r multiple_traj_data, include=FALSE}
# COVID19 case data for Ontario
# It makes sense to assume these case counts are incidence (# number of new 
# cases each day) instead of prevalence (# all active cases each day)
daily_cases = (read.csv(
    system.file(
        "starter_models"
      , "shiver"
      , "data"
      , "cases_ontario.csv"
      , package = "macpan2"
    )
  , row.names = NULL
  )
  |> rename(time = X_id)
  |> rowwise()
  # aggregate all case counts (by vax status) into one
  |> mutate(value = sum(c_across(starts_with("covid19_cases")),na.rm = TRUE))
)


reported_cases = (daily_cases
     |> filter(time != 1 & time != 2)  # remove first two records (Aug 9, Aug10 = I0)
     |> mutate(time = time - 2)        # update all times to set Aug 11 to be day 1 of the scenario
     |> head(expected_daily_reports)
     |> mutate(matrix = "reported_incidence") # the infection variable is incidence
     |> select(c("time","value","matrix"))
)

# use Aug 10 as initial I
# Note: this case data is from a different source than above, but I0 and I0_new
# don't differ much
I0_new = daily_cases |> filter(time==2) |> select(value) |> pull()
I0_new = I0_new * 10  ## account for under-reporting a bit

# check defaults again
mp_default(reparameterized_spec)
```

We need to expand our model to accommodate more data.

```{r incidence_in_model}
multi_traj_spec = (reparameterized_spec
  |> mp_tmb_insert(phase = "during", at = Inf
    , expressions = list(incidence ~ unvaccinated_infection + vaccinated_infection)
  ) 
  
  ## with case report data, we need to account for reporting delays and 
  ## under-reporting.
  |> mp_tmb_insert_reports("incidence"
    , report_prob = 0.1, mean_delay = 11, cv_delay = 0.25
    , report_prob_name = "report_prob"
  )
  
  ## we also need to prepare for a more flexible fit to the transmission
  ## rate that varies over time, as the report data provide sufficiently more
  ## information for this purposes.
  |> mp_tmb_insert(phase = "during", at = 1L
    , expressions = list(beta ~ beta * rbf_beta)
    , default = list(
          rbf_beta = 1
        , V = V0
        , N = N
      )
  )
)
```



```{r multiple_traj_calib}
## we need a more elaborate prior distribution
sd_par = 1 ## for convenience we give all parameters the same prior sd, for now
sd_state = 4 ## extremely vague priors on state variables
prior_distributions = list(
    log_beta = mp_normal(log(0.2), sd_par)
  , log_sigma = mp_normal(log(sigma), sd_par)
  , log_gamma_h = mp_normal(log(0.07), sd_par)
  , logit_report_prob = mp_normal(qlogis(0.1), sd_par)
  , logit_p = mp_normal(qlogis(1/4), 4)
  , log_E_I_ratio = mp_normal(0, sd_par)
  , log_I = mp_normal(log(I0), sd_state)
  , log_H = mp_normal(log(H0), sd_state)
)

## put the data together
dd = rbind(reported_hospitalizations, reported_cases)

# calibrate
shiver_calibrator = mp_tmb_calibrator(
    spec = mp_hazard(multi_traj_spec)
    # row bind both observed data
  , data = dd
    # fit both trajectories with log-normal distributions
    # (changed from negative binomial because apparently it is easier
    # to fit standard deviations than dispersion parameters)
  , traj = list(
      H = mp_normal(sd = mp_fit(1))
    , reported_incidence = mp_normal(sd = mp_fit(1))
  )
  , par = prior_distributions
    # fit the transmission rate using five radial basis functions for
    # a flexible model of time variation.
  , tv = mp_rbf("rbf_beta", 5, sparse_tol = 1e-8)
  , outputs = c(mp_state_vars(spec), "reported_incidence", "beta")
)
```

Next we optimize, and look at our estimates.

```{r mult_traj_optim, include=FALSE}
# optimize to estimate transmission parameters
# converges with warnings
mp_optimize(shiver_calibrator)
```

```{r mult_traj_estimates, echo=FALSE}
#check estimates
cc = mp_tmb_coef(shiver_calibrator, conf.int = TRUE)
round_coef_tab(cc)
```

Our prior for `sigma` is similar to the posterior, but `gamma_h` seems to have been pushed up by the data from about `0.05` to about `r filter(cc, mat == "gamma_h") |> pull(estimate) |> round(2L)`. We still do not have confidence in our estimate of `p`. We now have five other parameters controlling transmission, and so to interpret them we really need a plot of how transmission varies over time in the model. We add this variable to our model fit plot.

```{r mult_traj_fit, echo=FALSE}
(shiver_calibrator
   |> mp_trajectory_sd(conf.int = TRUE)
   |> ggplot(aes(time, value))
   + geom_line(aes(y = value), colour = "red")
   + geom_ribbon(
        aes(ymin = conf.low, ymax = conf.high)
      , fill = "red"
      , alpha = 0.3
    )
   + geom_point(data = dd, aes(time, value))
   + facet_wrap(vars(matrix), scales = 'free')
   + theme_bw()
)
```


This fit makes more sense. The transmission rate for unvaccinated people varies from about `0.1` to `0.3`. We have a lot of uncertainty about state variables that we don't fit, but this makes sense. Importantly `I` does not drop off to zero, which was an unrealistic aspect of previous fits.

## Parameter Identifiability

As a first step in calibration, we advise specifying a default for the parameter of interest, simulate data from the model and then calibrate to this simulated data to see if the default parameter can be recovered. If we are unable to recover the default, this could mean there are identifiability issues with the model and parameter. This should have been the first step in this example, before calibrating to observed data.

Here we specify a ground truth for `p` and `beta`, and simulate data from the model using these default values. We perform two calibrations, first we estimate `p` with a fixed `beta`, and then we do the reverse.

```{r ground_truth}
# set true values
true_p = 0.2
true_beta = 0.3 
```

```{r identifiability}
# simulate fake data
simulated_data = (reparameterized_spec
  |> mp_simulator(
      time_steps = expected_daily_reports
    , outputs=mp_state_vars(spec)
    , default=list(
        logit_p=qlogis(true_p)
      , log_beta=log(true_beta)
    )
    )
  |> mp_trajectory()
  # add some noise
  |> mutate(value = rpois(n(),value))
  |> select(-c(row,col))
)


## fix beta estimate p
fixed_beta = mp_tmb_calibrator(
    spec = reparameterized_spec |> mp_rk4()
  , data = simulated_data
  , traj = mp_state_vars(spec)
  , par = c("logit_p")
  , outputs = mp_state_vars(spec)
)
# converges, but not getting estimate for `p`
mp_optimize(fixed_beta)
mp_tmb_coef(fixed_beta, conf.int = TRUE) |> round_coef_tab()
## fix p estimate beta
fixed_a = mp_tmb_calibrator(
  spec = reparameterized_spec |> mp_rk4()
  , data = simulated_data
  , traj = mp_state_vars(spec)
  , par = c("log_beta")
  , outputs = mp_state_vars(spec)
)
# converges and recovering true beta
mp_optimize(fixed_a)
mp_tmb_coef(fixed_a, conf.int = TRUE) |> round_coef_tab()
```

We are able to recover the transmission rate `beta` but not the proportion `p`, suggesting model identifiability issues with this parameter.

# Model Specification

This model has been specified in the `shiver` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/shiver/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). 

# References

