---
title: "HIV"
index_entry: "A simple HIV model"
bibliography: ../../references.bib
link-citations: TRUE
author: Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```


This article describes a `macpan2` implementation of the HIV model proposed by [@granich2009universal]. This model played a key role in advancing HIV prevention using antiretroviral therapy, predicting that a test-and-treat strategy would substantially lower HIV incidence [@hayes2019effect]. Default parameter estimates [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/tmb.R) are inspired by [@kretzschmar2013prospects].

# Packages Used

The code in this article uses the following packages.

```{r packages, warning=FALSE, message=FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
```


# Model Specification

This model has been specified in the `hiv` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_lib}
spec = mp_tmb_library(
    "starter_models"
  , "hiv"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagram using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 2.5, fig.width = 8.5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_grid(spec
  , east = "^(infection|progression|death)"
  , south = "^(protection)"
  , north = "^(unprotection)"
  , loops = "^(unprotection)"
  , x_gap = 0.3
  , y_gap = 0.3
  , north_south_sep = 0.15
)
diagram = plot_flow_diagram(layout)
(diagram 
  |> draw_outflows(layout)
  |> draw_inflows(layout)
)
```


# States

| Variable | Description                                                                                                        |
| -------- | ------------------------------------------------------------------------------------------------------------------ |
| S        | Number of susceptible individuals.                                                                                 |
| I1 - I4  | Numbers of infectious individuals who are not treated. The number gives the state of disease progression from 1-4. |
| A1 - A4  | Numbers of infectious individuals who are treated. The number gives the state of disease progression from 1-4.     |
| D        | Number of individuals who have died from the disease.                                                              |
| D_bg     | Number of individuals who have died from other causes.                                                             |

# Parameters

| Variable    | Description                                                                                                               |
| ----------- | ------------------------------------------------------------------------------------------------------------------------- |
| $\lambda_0$ | Baseline transmission rate.                                                                                 |
| $\alpha$    | Constant in non-linear transmission rate, accounting for heterogeneity in sexual behaviour.                                                                                 |
| $n$         | Constant in non-linear transmission rate, accounting for heterogeneity in sexual behaviour.                                                                                 |
| $\epsilon$  | Constant, in non-linear transmission rate, measuring the relative decrease in transmission caused by treatment. |
| $\beta$     | Per-capita birth rate.                                                                                                    |
| $\mu$       | Per-capita (background) death rate.                                                                                       |
| $\rho$      | Per-capita rate of disease progression for non-treated individuals.                                                       |
| $\sigma$    | Per-capita rate of disease progression for treated individuals.                                                           |
| $\tau$      | Per-capita rate at which individuals become protected.                                                                    |
| $\phi$      | Per-capita rate at which individuals become unprotected.                                                                  |


# Force-of-Infection

This model has the somewhat non-standard functional form for the force-of-infection (per-capita transition rate from `S` to `I1`), $\frac{\lambda J}{N}$ where $\lambda = \lambda_0 e^{-\alpha P^n}$, $P = I/N$, $I = \sum_i(I_i + A_i)$, $J = \sum_i(I_i + \epsilon A_i)$, and $N$ is the total number of alive boxes.

The remaining transition rates are constant per-capita rates.

# Solving the ODEs

Here is an example trajectory from this model with small [updates](?mp_tmb_update()) to the default parameters, using the [Runge-Kutta ODE solver](?mp_rk4()).

```{r simulations}
outputs = c(sprintf("I%s", 1:4), sprintf("A%s", 1:4))
sim = (spec
  |> mp_tmb_update(default = list(lambda0 = 0.36, n = 0.2))
  |> mp_rk4()
  |> mp_simulator(time_steps = 50L, outputs)
)
(sim
  |> mp_trajectory()
  |> mutate(matrix = sub("^A([1-4])$", "Infectious and treated, stage \\1", matrix))
  |> mutate(matrix = sub("^I([1-4])$", "Infectious and untreated, stage \\1", matrix))
  |> ggplot()
  + geom_line(aes(time, value))
  + facet_wrap(~ matrix, ncol = 2, scales = 'free', dir = "v")
  + scale_y_continuous(limits = c(0, NA), expand = c(0, 0))
  + theme_bw()
)
```


# References
