---
title: "HIV"
index_entry: "A simple HIV model"
bibliography: references.bib
link-citations: TRUE
author: Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```


This is a description of the HIV model proposed by [@granich2009universal].

# States

| Variable | Description                                                                                                        |
| -------- | ------------------------------------------------------------------------------------------------------------------ |
| S        | Number of susceptible individuals.                                                                                 |
| I1 - I4  | Numbers of infectious individuals who are not treated. The number gives the state of disease progression from 1-4. |
| A1 - A4  | Numbers of infectious individuals who are treated. The number gives the state of disease progression from 1-4.     |
| D        | Number of individuals who have died from the disease.                                                              |
| D_bg     | Number of individuals who have died from other causes.                                                             |

# Parameters

| Variable    | Description                                                                                                               |
| ----------- | ------------------------------------------------------------------------------------------------------------------------- |
| $\alpha$    | Constant in non-linear transmission rate.                                                                                 |
| $\lambda_0$ | Constant in non-linear transmission rate.                                                                                 |
| $n$         | Constant in non-linear transmission rate.                                                                                 |
| $\epsilon$  | Constant, in non-linear transmission rate, measuring the relative decrease in transmission caused by treatment. |
| $\beta$     | Per-capita birth rate.                                                                                                    |
| $\mu$       | Per-capita (background) death rate.                                                                                       |
| $\rho$      | Per-capita rate of disease progression for non-treated individuals.                                                       |
| $\sigma$    | Per-capita rate of disease progression for treated individuals.                                                           |
| $\tau$      | Per-capita rate at which individuals become protected.                                                                    |
| $\phi$      | Per-capita rate at which individuals become unprotected.                                                                  |

# Packages Used

The code in this artical uses the following packages.
```{r packages, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
library(ggraph)
library(tidygraph)
```


# Dynamics

We can read in the model specification using the `mp_tmb_library` command.
```{r model_lib}
spec = mp_tmb_library(
    "starter_models", "hiv"
  , package = "macpan2"
)
```

The specification can be used to draw the flow diagram using the following code.

```{r flow_diagram}
x_pos = c(
    S = 1
  , I1 = 2, I2 = 3, I3 = 4, I4 = 5
  , A1 = 2, A2 = 3, A3 = 4, A4 = 5
  , D = 6
) / 7
y_pos = c(
    S = 6
  , I1 = 6, I2 = 6, I3 = 6, I4 = 6
  , A1 = 1, A2 = 1, A3 = 1, A4 = 1
  , D = 6
) / 7
node_size = 10
(spec
  |> mp_flow_frame(topological_sort = FALSE)
  |> filter(!abs_rate %in% c("mu", "sigma"))
  |> distinct()
  |> as_tbl_graph()
  |> ggraph('manual', x = x_pos[name], y = y_pos[name])
  + geom_edge_link(
      arrow = arrow(length = unit(node_size * 0.3, 'mm'))
    , end_cap = circle(node_size / 2, 'mm')
    , start_cap = circle(node_size / 2, 'mm')
  )
  + geom_node_point(size = node_size, colour = "lightgrey")
  + geom_node_text(aes(label = name), family = "mono")
  + theme_graph(background = NA, plot_margin = margin(0, 0, 0, 0))
  + scale_x_continuous(limits = c(0, 1))
  + scale_y_continuous(limits = c(0, 1))
)
```


# Solving the ODEs

Here is an example trajectory from this model with defaults using the Runge-Kutta ODE solver.

```{r}
outputs = c(
    "D_bg", "D", "S"
  , "Itotal", "Atotal", "lambda"
)
sim = (spec
  |> mp_rk4()
  |> mp_simulator(time_steps = 30L, outputs)
)
(sim
  |> mp_trajectory()
  |> ggplot()
  + geom_line(aes(time, value))
  + facet_wrap(~ matrix, ncol = 3, scales = 'free')
)
```


# Model Specification

This model has been specified in the `hiv` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details).

# References
