---
title: "HIV"
index_entry: "A simple HIV model"
bibliography: ../../references.bib
link-citations: TRUE
author: Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```


This article describes a `macpan2` implementation of the HIV model proposed by [@granich2009universal]. This model played a key role in advancing HIV prevention using antiretroviral therapy, predicting that a test-and-treat strategy would substantially lower HIV incidence [@hayes2019effect]. Default parameter estimates [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/tmb.R) are inspired by [@kretzschmar2013prospects].

# Packages Used

The code in this article uses the following packages, and set an option that keeps the article tidier by suppressing information about optimization.

```{r packages, warning=FALSE, message=FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
options(macpan2_verbose = FALSE)
```


# Model Specification

This model has been specified in the `hiv` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_lib}
spec = mp_tmb_library(
    "starter_models"
  , "hiv"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagram using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/hiv/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 2.5, fig.width = 8.5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_grid(spec
  , east = "^(infection|progression|death)"
  , south = "^(protection)"
  , north = "^(unprotection)"
  , loops = "^(unprotection)"
  , x_gap = 0.3
  , y_gap = 0.3
  , north_south_sep = 0.15
)
(layout
  |> plot_flow_diagram(show_flow_rates = TRUE)
  |> draw_outflows(layout, show_labels = TRUE, lab = "rate")
  |> draw_inflows(layout, show_labels = TRUE, lab = "rate")
)
```


# States

| Variable | Description                                                                                                        |
| -------- | ------------------------------------------------------------------------------------------------------------------ |
| S        | Number of susceptible individuals.                                                                                 |
| I1 - I4  | Numbers of infectious individuals who are not treated. The number gives the state of disease progression from 1-4. |
| A1 - A4  | Numbers of infectious individuals who are treated. The number gives the state of disease progression from 1-4.     |
| D        | Number of individuals who have died from the disease.                                                              |
                                                          |

# Parameters

| Variable    | Description                                                                                                               |
| ----------- | ------------------------------------------------------------------------------------------------------------------------- |
| $\lambda_0$ | Baseline transmission rate.                                                                                 |
| $\alpha$    | Constant in non-linear transmission rate, accounting for heterogeneity in sexual behaviour.                                                                                 |
| $n$         | Constant in non-linear transmission rate, accounting for heterogeneity in sexual behaviour.                                                                                 |
| $\epsilon$  | Constant, in non-linear transmission rate, measuring the relative decrease in transmission caused by treatment. |
| $\beta$     | Per-capita birth rate.                                                                                                    |
| $\mu$       | Per-capita (background) death rate.                                                                                       |
| $\rho$      | Per-capita rate of disease progression for non-treated individuals.                                                       |
| $\sigma$    | Per-capita rate of disease progression for treated individuals.                                                           |
| $\tau$      | Per-capita rate at which individuals become protected.                                                                    |
| $\phi$      | Per-capita rate at which individuals become unprotected.                                                                  |


# Force-of-Infection

This model has the somewhat non-standard functional form for the force-of-infection (per-capita transition rate from `S` to `I1`):

$$
\frac{\lambda J}{N}
$$

The symbols in this expression are defined as follows.

* $\lambda = \lambda_0 e^{-\alpha P^n}$
* $P = I/N$
* $I = \sum_i(I_i + A_i)$
* $J = \sum_i(I_i + \epsilon A_i)$
* $N$ is the total number of alive boxes

The remaining transition rates are constant per-capita rates.

# Solving the ODEs

Here is an example trajectory from this model with small [updates](https://canmod.github.io/macpan2/reference/mp_tmb_update) to the default parameters, using the [Runge-Kutta ODE solver](https://canmod.github.io/macpan2/reference/state_updates).

```{r simulations}
outputs = c(sprintf("I%s", 1:4), sprintf("A%s", 1:4))
sim = (spec
  |> mp_tmb_update(default = list(lambda0 = 0.36, n = 0.2))
  |> mp_rk4()
  |> mp_simulator(time_steps = 50L, outputs)
)
(sim
  |> mp_trajectory()
  |> mutate(matrix = sub("^A([1-4])$", "Infectious and treated, stage \\1", matrix))
  |> mutate(matrix = sub("^I([1-4])$", "Infectious and untreated, stage \\1", matrix))
  |> ggplot()
  + geom_line(aes(time, value))
  + facet_wrap(~ matrix, ncol = 2, scales = 'free', dir = "v")
  + scale_y_continuous(limits = c(0, NA), expand = c(0, 0))
  + theme_bw()
)
```

# Calibration

To illustrate calibration, we simulate 20 years of data from a population of `1e7` people using two variables: the prevalence of infectious individuals who are `treated` and `untreated`. We initialize with `1000` infectious individuals in each of the four disease stages, and zero untreated individuals.  We add Poisson noise to these variables.

```{r simulated_data_for_cal, fig.width = 4}
set.seed(1L)
spec_for_cal = (spec
  |> mp_tmb_update(
      default = list(lambda0 = 0.38, n = 0.2)
    , inits = list(
          S = 1e7 - 4000
        , I1 = 1000, I2 = 1000, I3 = 1000, I4 = 1000
        , A1 = 0   , A2 = 0   , A3 = 0   , A4 = 0
      )
  )
  |> mp_rk4()
  |> mp_tmb_insert(at = Inf, expressions = list(
      treated ~ A1 + A2 + A3 + A4
    , untreated ~ I1 + I2 + I3 + I4
  ))
)
simulated_data = (spec_for_cal
  |> mp_simulator(time_steps = 20L, c("treated", "untreated"))
  |> mp_trajectory()
  |> mutate(value = rpois(n(), value))
)
(simulated_data
  |> rename(`Observation Year` = time)
  |> rename(Value = value)
  |> ggplot()
  + geom_line(aes(`Observation Year`, Value))
  + facet_wrap(~matrix, ncol = 1, scales = "free")
  + theme_bw()
)
```


We calibrate this model data simulated from it, but start away from the true parameter values by specifying that the default `lambda0 = 0.2` and `n = 0.5`. We assume a Poisson likelihood to match the Poisson noise that we used to produce the simulated data. For numerical stability, we optimize `lambda0` and `n` on the log and logit scales respectively.
```{r calibration}
calibrator = (spec_for_cal
  |> mp_tmb_update(default = list(lambda0 = 0.2, n = 0.5))
  |> mp_tmb_calibrator(
        data = simulated_data
      , traj = list(
            treated = mp_poisson()
          , untreated = mp_poisson()
      )
      , par = c("log_lambda0", "logit_n")
  )
)
mp_optimize(calibrator)
```

The convergence code is `0`, which is good.
```{r convergence}
mp_optimizer_output(calibrator)$convergence
```

The true parameters, `lambda0 = 0.38` and `n = 0.2`, are in the confidence intervals, although the confidence interval for `n` is quite wide.
```{r estimates}
(mp_tmb_coef(calibrator, conf.int = TRUE)
 |> select(-term, -row, -col, -type)
)
```

The simulated data (black) that we fit to matches the predictions of the fitted model (red) with 95% confidence intervals for the point prediction).
```{r fit, fig.width = 4}
(calibrator
 |> mp_trajectory_sd(conf.int = TRUE)
 |> ggplot()
 + geom_line(aes(time, value), colour = "red")
 + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "red")
 + geom_line(aes(time, value), data = simulated_data)
 + facet_wrap(~matrix, ncol = 1, scales = "free")
 + theme_bw()
)
```

# References
