---
title: "Awareness Models"
index_entry: "Behaviour modifications in response to death"
bibliography: ../../references.bib
link-citations: TRUE
author: Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "./figures/"
)
```

SEIR-type models with awareness-driven behaviour, inspired by [@weitz2020awareness] and [@hurford2023pandemic].

# Packages Used

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```


# Model Specification

We describe several models here that have been specified in the `awareness` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/awareness/tmb.R) and are accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
specs = mp_tmb_library(
    "starter_models"
  , "awareness"
  , package = "macpan2"
  , alternative_specs = TRUE
)
print(names(specs))
```

These specifications can be used to draw the following flow diagrams using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/awareness/README.Rmd). This is the diagram for the simple awareness model. 

```{r diagram, echo = FALSE, fig.height = 2, fig.width = 5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_paths(specs$awareness_model)
plot_flow_diagram(layout)
```

The other three models in this example have a hospital box as well.

```{r diagram-delayed-death, echo = FALSE, fig.height = 2, fig.width = 5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_paths(specs$delayed_death_awareness_model)
plot_flow_diagram(layout)
```


# States

| variable | description  |
| -------- | ------------ |
| S        | Susceptible  |
| E        | Exposed      |
| I        | Infectious   |
| R        | Recovered    |
| D        | Dead         |
| H        | Hospitalized |
|          |              |


# Parameters

| variable               | description                                                                   |
| ---------------------- | ----------------------------------------------------------------------------- |
| $\beta$                | per capita transmission rate                                                  |
| $\mu$                  | per capita infection rate (average time spent in compartment E is $1/\alpha$) |
| $\gamma$               | per capita recovery/death rate                                                |
| `importation_prob`     | probability that one more infectious individual will arrive on any day.       |
| `memory_length`        | how long do people 'care about' deaths (days)                                 |
| $1/\gamma_h$           | mean time between isolation and death (days)                                  |
| $\delta_c$             | half-saturation constant for death awareness.                                 |
| $\delta_{\text{long}}$ | half-saturation constant for longer-term death awareness.                     |
| $f_D$                  | fraction of $\gamma$ that contributes to death rather than recovery           |
| $k$                    | exponent for half-saturation power law                                        |

# Base Awareness Model

$$
\begin{align*}
\frac{dS}{dt} &= -S\frac{\beta I}{N(1 + (I \gamma f_D / \delta_c)^k)} \\
\frac{dE}{dt} &= S\frac{\beta I}{N(1 + (I \gamma f_D / \delta_c)^k)} - \mu E \\
\frac{dI}{dt} &= \mu E- \gamma (1 - f_D) I \\
\frac{dR}{dt} &= \gamma (1 - f_D) I \\
\frac{dD}{dt} &= I \gamma f_D \\
\end{align*}
$$

# Delayed Death Awareness Model

$$
\begin{align*}
\frac{dS}{dt} &= -S\frac{\beta I}{N(1 + (I \gamma f_D / \delta_c)^k)} \\
\frac{dE}{dt} &= S\frac{\beta I}{N(1 + (I \gamma f_D / \delta_c)^k)} - \mu E \\
\frac{dI}{dt} &= \mu E- \gamma (1 - f_D) I \\
\frac{dR}{dt} &= \gamma (1 - f_D) I \\
\frac{dH}{dt} &= I \gamma f_D - H \gamma_h\\
\frac{dD}{dt} &= H \gamma_h \\
\end{align*}
$$

# Longer Memory Awareness Model

This model is identical to the [delayed death awareness model](#delayed-death-awareness-model), except that the factor $I \gamma f_D$ in the force of infection is replaced with a temporal convolution that takes a weighted average of past values with more recent values being weighted more heavily. See [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/awareness/tmb.R) for details.

This model is able to simulate behavioural cycles, where incidence declines because people change their behaviour not because of a depletion of susceptible individuals. Note that the `plot_traj` function is defined in the [source code for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/awareness/README.Rmd).

```{r, echo = FALSE}
plot_traj = function(traj) {
  p = (traj
    |> mutate(matrix = factor(matrix, levels = outputs))
    |> ggplot()
    + geom_line(aes(time, value))
    + facet_wrap(~matrix
        , scales = "free_y"
        , ncol = 1
        , dir = "v"
      )
    + scale_y_continuous(limits = c(0, NA), expand = c(0, NA))
    + scale_x_continuous(
          minor_breaks = \(x) {
            y = seq(from = x[1], to = x[2], by = 50L)
            y[1:(length(y) - 1)]
          }
        , expand = c(0, 0)
      )
    + ylab("")
    + xlab("")
    + theme_bw()
  )
  return(p)
}
```

```{r behavioural_cycles}
set.seed(8L)
days = 800
outputs = c(
    "infection", "death"
  , "S", "importation"
)
sim = (specs$longer_memory_awareness_model
  |> mp_euler()
  |> mp_simulator(days, outputs)
)
traj = (sim
  |> mp_trajectory(include_initial = TRUE)
)
plot_traj(traj)
```

This code uses the [`mp_euler`](https://canmod.github.io/macpan2/reference/mp_euler.html) function for ensuring that the Euler method is used to solve the differential equations. This method is the default, but we draw attention to it here as we will change it below. Then we do the [standard](https://canmod.github.io/macpan2/articles/quickstart.html) `macpan2` thing of generating simulator (using [mp_simulator](https://canmod.github.io/macpan2/reference/mp_simulator.html)) and then using that simulator to simulate trajectories from the model (using [mp_trajectory](https://canmod.github.io/macpan2/reference/mp_trajectory.html)).

# Importation Awareness Model

This model is identical to the [longer memory awareness model](#longer-memory-awareness-model) except that it includes random importations. These importations are simulated by drawing a Bernoulli random variable at each time step, adding it to the `I` box, and removing from the `R` box.

```{r importation}
set.seed(8L)
days = 800
outputs = c(
    "infection", "death"
  , "S", "importation"
)
sim = (specs$importation_awareness_model
  |> mp_euler()
  |> mp_simulator(days, outputs)
)
traj = (sim
  |> mp_trajectory(include_initial = TRUE)
)
plot_traj(traj)
```

This model is particularly interesting for small populations if we use the [`mp_euler_multinomial`](https://canmod.github.io/macpan2/reference/mp_euler.html) state update method to generate process error from the Euler multinomial distribution.

```{r importation_and_process_error}
set.seed(8L)
days = 800
outputs = c(
    "infection", "death"
  , "S", "importation"
)
sim = (specs$importation_awareness_model
  |> mp_euler_multinomial()
  |> mp_simulator(days, outputs)
)
traj = (sim
  |> mp_trajectory(include_initial = TRUE)
)
plot_traj(traj)
```

The process error combined with the random importation has the effect of irregularly spaced cycles.

# References
