---
title: "Calibrating Compartmental Models to Data"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Calibrating Compartmental Models to Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

[![status](https://img.shields.io/badge/status-working%20draft-red)](https://canmod.github.io/macpan2/articles/vignette-status#working-draft)

```{r pkgs, include = FALSE}
library(macpan2)
library(macpan2helpers)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Hello World Fitting Problem (IN PROGRESS)


First set up the model from the quickstart guide:

```{r sir_setup}
sir = Compartmental(system.file("starter_models", "sir", package = "macpan2"))
sir_simulator = sir$simulators$tmb(
  time_steps = 100,
  state = c(S = 99, I = 1, R = 0),
  flow = c(foi = NA, gamma = 0.1),
  beta = 0.2,
  N = empty_matrix
)
sir_results = sir_simulator$report()
```

Add some noise to the prevalence (`I`) value:
```{r sir_noise}
set.seed(101)
sir_prevalence = (sir_results
    |> dplyr::select(-c(matrix, col))
    |> filter(row == "I")
    |> mutate(obs_val = value + rnorm(n(), sd = 1))
)
ggplot(sir_prevalence, aes(time)) +
    geom_point(aes(y = obs_val)) +
    geom_line(aes(y = value))
```

Step 1:  use `$add$matrices()` to:

(a) add observed data 
(b) declare a matrix storing the simulation history of variables to compare with observed data (if you want you can return these values in the report)
(c) declare a matrix to store the log-likelihood
(d) specify which matrices you want to save and/or return in the report

```{r sim_addmat}
sir_simulator$add$matrices(
  ## observed data
  I_obs = sir_prevalence$obs_val,
  ## simulated trajectory to compare with data
  I_sim = empty_matrix, 
  ## location of I in the state vector
  ## (the `-1L` bit is to get 0-based indices instead of 1-based)
  I_index = match("I", sir$labels$state()) - 1L,
  ## matrix to contain the log likelihood values at each time step
  log_lik = empty_matrix,
  ## need to save the simulation history of each of these matrices
  .mats_to_save = c("I_sim", "log_lik"),
  .mats_to_return = c("I_sim")
)
```


**fixme**: some examples show `sir_simulator$print$matrix_dims()` at this point.  What is this good for/how do we interpret it?


Step 2: collect simulated values into matrices to be
compared with data. the `.at = Inf` and
`.phase = "during"` indicates that this expression
should come at the *end* of the expressions evaluated
during each iteration of the simulation loop.

```{r add_Isim}
sir_simulator$insert$expressions(
  I_sim ~ state[I_index],
  .phase = "during",
  .at = Inf
)
```

Step 3: compute any values that will be part of the
objective function to be optimized. here we
have the log (by default) of the Gaussian density of the
observed `I` values with mean (i.e. predicted)
value of the simulated `I` values. the
`rbind_time` function gathers together the
full simulation history of the `I_sim`
matrix by binding together the rows at each
iteration.

```{r compute_loglik}
sir_simulator$insert$expressions(
  log_lik ~ dnorm(I_obs, rbind_time(I_sim), I_sd),
  .phase = "after"
  )
```

Step 4: specify the objective function (usually
the sum of the negative log-likelihoods)

```{r add_loglik}
sir_simulator$replace$obj_fn(~ -sum(log_lik))
```

Step 5: declare (and maybe transform) parameters to be optimized,
as well as starting values for the parameters to be optimized

```{r step5}
sir_simulator$add$matrices(
                      I_sd = 1
                  )

sir_simulator$add$transformations(Log("I_sd"))
sir_simulator$add$transformations(Log("beta"))
params <- read.delim(sep = "|", header = TRUE, text = "
mat       | row | col | default
log_I_sd  | 0   | 0   | 0
log_beta  | 0	| 0   | 1
")
try(sir_simulator$replace$params_frame(params))
```

**fixme**: on the last line above I get a less-than-perfectly-useful error message:

(Use this chunk to turn off evaluation while working through example:

```{r stop_eval}
knitr::opts_chunk$set(eval = FALSE)
```

```{r start_eval, eval = TRUE}
## turn evaluation back on
knitr::opts_chunk$set(eval = TRUE)
```

## Measles Data

Here is a reasonably difficult problem -- fit an SIR model to weekly measles incidence data from London UK over about six decades.

<!-- FIXME: store data locally, don't rely on IIDA ... -->
```{r sir_plot, fig.width=6}
measles = read.csv(
  file.path(
    "https://raw.githubusercontent.com/davidearn/iidda/master/data",
    "meas_uk__lon_1944-94_wk/source-data/meas_uk__lon_1944-94_wk.csv"
  ),
  comment = "#"
)
measles$date = as.Date(sprintf(
  "%s-%s-%s", measles$year, measles$month, measles$day
))
plot(measles$date, measles$cases, type = "l")
```

We need to slightly extend the standard SIR model to include waning immunity.

```{r sir_waning}
sir = Compartmental(system.file("starter_models", "sir_waning", package = "macpan2"))
sir$flows()
```


We use [radial basis functions](https://canmod.github.io/macpan2/articles/time_varying_parameters.html#radial-basis-functions-for-flexible-time-variation-in-progress) to model time-variation in the transmission rate.  We also make a variety of questionable assumptions (TODO: fix these), but the point at the moment is just to illustrate usage and provide a proof of concept.
```{r rbf}
d = 100
n = nrow(measles)
simulator = sir$simulators$tmb(
      time_steps = n
    , state = c(S = 100000 - 500, I = 500, R = 0)
    , flow = c(foi = NA_real_, gamma = 0.2, wane = 0.01)
    
    ## this beta does not matter because we will overwrite
    ## it with the output of the radial basis functions
    , beta = NA_real_
    
    ## FIXME: this is surely not the population of London at
    ##        all in the series
    , N = 100000 
    
    ## matrices involved in radial basis functions
    , X = rbf(n, d)
    , b = rnorm(d, sd = 0.01)
    , incidence = empty_matrix
    , eta = empty_matrix
    
    , .mats_to_save = c("state", "incidence", "beta")
    , .mats_to_return = c("state", "incidence", "beta")
    
## initial S is a function of initial I, which we
## fit to data below
)$insert$expressions(
    S ~ N - I
  , .phase = "before"
  , .at = 1
  
## radial basis function evaluations
)$insert$expressions(
    eta ~ gamma * exp(X %*% b)
  , .phase = "before"
  , .at = Inf
)$insert$expressions(
    beta ~ eta[time_step(1)] / clamp(S/N, 1/100)
  , .phase = "during"
  , .at = 1
  
## save the simulated incidence trajectory to
## compare with data
)$insert$expressions(
    incidence ~ I
  , .vec_by_states = "total_inflow"
  , .phase = "during"
  , .at = Inf
)
```

Here is an example simulation from this model, before fitting to data.
```{r rbf_ex, fig.width=6}
set.seed(1L)
simulated_incidence = filter(simulator$report(.phases = "during"), matrix == "incidence")$value
plot(measles$date, simulated_incidence, type = "l", xlab = "time")
```

It looks nothing like the observed measles series, but illustrates the ability to generate complex incidence patterns not present in the simple SIR model without radial basis functions and waning immunity.


```{r rbf_model, echo=FALSE}
simulator = ("https://github.com"
  |> file.path("canmod/macpan2")
  |> file.path("raw/main")
  |> file.path("misc/saved-simulators/measles-model-object.rds")
  |> url('rb')
  |> readRDS()
)
```

We modify the simulation object to be able to fit to the measles data.

```{r rbf_modify, eval=FALSE}
simulator$add$matrices(
    reports = measles$cases
  , log_lik = empty_matrix
  , sim_reports = empty_matrix
)
simulator$insert$expressions(
    sim_reports ~ rbind_time(incidence)
  , .phase = "after"
  , .at = Inf
)
simulator$replace$params(
  default = c(
      c(0.2, 0.01)
    , rep(0, d)
    , 500
  )
  , mat = c(
      rep("flow", 2L)
    , rep("b", d)  
    , "state"
  )
  , row = c(
      (1:2)
    , seq_len(d) - 1L
    , 1
  )
)
simulator$replace$obj_fn(~ - sum(dpois(reports, sim_reports)))
```

The optimization takes quite a few minutes, and still doesn't converge in 10000 function evaluations.
```{r rbf_optim_fake, eval=FALSE}
simulator$optimize$nlminb(control = list(eval.max = 10000, iter.max = 10000, trace = 10))
```



```{r rbf_opt_get}
simulator$optimization_history$get()[[3]]  ## the 3 is there because we tried two other times
```

Here the red data are fitted and black observed.
```{r plot_rbf_res, fig.width=6}
simulated_incidence = filter(simulator$report(.phases = "during"), matrix == "incidence")$value
plot(measles$date, measles$cases, xlab = "time", type = "l")
lines(measles$date, simulated_incidence, col = 2)
```

Not a perfect fit, but not bad for now (TODO: work on this, without papering over the real challenges).

## Challenging Logistic Variation in Transmission Rate

Here we consider the problem of fitting an SIR model to a simulated dataset from this model, such that the simulations pose challenges to the fitting machinery.

```{r logistic_ex}
sir = Compartmental(system.file("starter_models", "sir", package = "macpan2"))
sir$flows_expanded()
```

Our simulation model includes a logistically time-varying transmission rate.

```{r logistic_sim}
n = 2500
set.seed(1L)
simulator = sir$simulators$tmb(
      time_steps = n
    , state = c(S = 100000 - 500, I = 500, R = 0)
    , flow = c(foi = NA, gamma = 0.2)#, wane = 0.01)
    , beta = 1
    , N = 100000
    , X = cbind(1, scale(seq_len(n)))
    , b = c(0, 1)
    , incidence = empty_matrix
    , beta_values = empty_matrix
    , .mats_to_save = c("state", "incidence", "beta")
    , .mats_to_return = c("state", "incidence", "beta")
)$insert$expressions(
    beta_values ~ 1 / (1 + exp(-X %*% b))
  , .phase = "before"
  , .at = Inf
)$insert$expressions(
    beta ~ beta_values[time_step(1)]
  , .phase = "during"
  , .at = 1
)$insert$expressions(
    incidence ~ I
  , .vec_by_states = "total_inflow"
  , .phase = "during"
  , .at = Inf
)$replace$params(
    default = c(0, 1)
  , mat = rep("b", 2)
  , row = 0:1
)
```

```{r logistic_plot2, fig.height=8, fig.width=6}
set.seed(5L)
sims = simulator$report(.phases = "during")
(sims
  |> mutate(variable = if_else(matrix == "state", row, matrix))
  |> ggplot()
  + facet_wrap(~ variable, ncol = 1, scales = 'free')
  + geom_line(aes(time, value))
)
```

Fitting to the simulation data, manages to converge, but to the wrong value.
```{r logistic_bad_converge, fig.width=6}
set.seed(3L) ## different seeds do result in convergence on the correct value
reports = filter(sims, matrix == "incidence")$value
simulator$add$matrices(reports = reports, report_sim = empty_matrix)
simulator$insert$expressions(
    report_sim ~ rbind_time(incidence)
  , .phase = "after"
  , .at = Inf
)
simulator$replace$obj_fn(~ -sum(dpois(reports, report_sim)))
simulator$replace$params(
    default = rnorm(2L) ## random starting values for the optimizer
  , mat = rep("b", 2)
  , row = 0:1
)
simulator$optimize$nlminb()
simulator$current$params_frame()
fitted_incidence = (simulator$current$params_vector()
  |> simulator$report()
  |> filter(matrix == "incidence")
  |> pull(value)
)
plot(reports, type = "l")
lines(fitted_incidence, col = 2)
```

The fit is not good!  Why?  To find out we plot the likelihood surface with arrows representing the magnitude and direction of the down-hill gradient towards the optimum. Notice the very flat gradient in the direction along the valley containing the optimum at $(0, 1)$. The gradient is pointing towards the valley but not along it.  I do not understand why.

```{r logistic_plot_surf, results='hide', message=FALSE}
lik_surf = expand.grid(
  intercept_parameter = seq(from = -1, to = 1, by = 0.05),
  slope_parameter = seq(from = 0, to = 2, by = 0.05)
)
gr = t(apply(lik_surf, 1, simulator$gradient))
lik_surf$z = apply(lik_surf, 1, simulator$objective)
gr = 0.1 * gr / max(abs(gr))
lik_surf$gx = gr[,1]
lik_surf$gy = gr[,2]
```



```{r logistic_plot_surf2, fig.height=6, fig.width=7}
(ggplot(lik_surf) 
  + geom_tile(aes(intercept_parameter, slope_parameter, fill = z)) 
  + geom_segment(
    aes(
      intercept_parameter, 
      slope_parameter, 
      xend = intercept_parameter - gx, 
      yend = slope_parameter - gy
    ), 
    arrow = arrow(length = unit(0.05, "inches")), 
    colour = 'white'
  )
  + theme_bw()
)
```

