---
title: "Engine-Agnostic Model Specification Grammar"
output: 
  html_document:
    toc: true
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Engine-Agnostic Model Specification Grammar}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

[![status](https://img.shields.io/badge/status-working%20draft-red)](https://canmod.github.io/macpan2/articles/vignette-status#working-draft)

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/"
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
```

One of the main goals of `macpan2` is to provide a flexible grammar for model specification that reduces friction when building upon and expanding an existing model. This goal complements the standard approach of modelling, which is to start simply and add complexity as needed. 

There is a trade-off between the flexibility and the simplicity of the model grammar: specifying a simple model may not always be very concise, and there is a learning curve to the model grammar. However, it can be very powerful when it comes to specifying **structured** models, especially when they are cast as expansions of simple models. Such [structured models](https://idpjournal.biomedcentral.com/articles/10.1186/s40249-022-01001-y) can include:

- multiple pathogen strains
- multiple infection types (_e.g._, asymptomatic and symptomatic or mild and severe)
- age-structure
- multiple locations (a metapopulation model)
- testing processes to identify infections
- vaccination status

This vignette seeks to explain `macpan2`'s model specification grammar and in particular how one could take a simple model and expand it with additional structure.

# Amuse bouche: a structured SIR model {#amuse-bouche}

A key to `macpan2`'s flexible model grammar is the use of **functional forms** to repeat the same kinds of calculations across model structures. For instance, consider an SIR model that has two pathogen strains (without co-infections):

```{r design-concepts, echo=FALSE}
knitr::include_graphics("../man/figures/two-strain.svg")
```

Here, 

- $S$, $I_x$, and $R$ are the numbers of individuals that are susceptible, infected with strain $x$ ($A$ or $B$), and recovered, respectively,
- $N= S + I_A + I_B + R$ is the total population size,
- $\beta_x$ is the transmission rate for strain $x$,
- $\gamma$ is the recovery rate for infected individuals,^[In this model we are choosing to make the recovery rate (and hence the infectious period) the same for both strains, but it would be easy to relax this assumption in the model specification by following the way in which we stratify the state variable $I$ by strain below.]
- $\lambda_x = \beta_x (I_x)/N$ is the force of infection for strain $x$. 

We can cast this model as a system of difference equations, since this is how we will iterate them numerically in our simulation:

\begin{align}
S_{t+1} & = - [\beta_A (I_A)_t/N + \beta_B (I_B)_t/N] S_t, \\ 
(I_A)_{t+1} &= \phantom{-[} \beta_A (I_A)_t/N - \gamma (I_A)_t, \\ 
(I_B)_{t+1} &= \phantom{-[} \beta_B (I_B)_t/N - \gamma (I_B)_t, \\ 
R_{t+1} &= \phantom{-[}  \gamma (I_A)_t + \gamma (I_B)_t.
\end{align}

Each force of infection, $\lambda_A = \beta_A (I_A)/N$ and  $\lambda_B = \beta_B (I_B)/N$ has the same **functional form**, that is, using an expression like $\lambda = \beta I / N$. When numerically simulating this model, it doesn't take much effort to write out each calculation separately as something like:

```
lambda.A = beta.A * I.A / N
lambda.B = beta.B * I.B / N
```

However, in `macpan2`, we can specify a single functional form for it, for instance

```
lambda = beta * I / N
```

and then attach a **ledger** to the model object that tabulates specific instances of when this functional form is used to define a component of the model. In other words, this ledger should enumerate which specific subscripted `lambda`, `beta`, and `I` to use each time we invoke the associated functional form during the simulation. 

In this case, there would only be two calculations in the force of infection ledger (one calculation per strain), but one can easily imagine a more complicated case. For instance, consider a relatively simple two-city age-structured metapopulation model with 10 age groups within each of two patches: there would be 10x10x2 = `r 10*10*2` force of infection terms of the same form (one per combination of age groups to capture the options for susceptible and infected interaction, repeated for each of the two patches).

**Using functional forms and ledgers allows the modeller to focus on modelling questions**, like the design of the model structure and the choice of expressions for the forces of infection, while **`macpan2` handles the bookkeeping**, matching stratified variables with each other when calculating expressions. This approach cuts down on rote repetition when setting up model calculations, which in turn reduces the opportunity for bugs in the simulation code. It also means that expanding a model can be as simple as updating the calculation ledger, rather than error-prone editing of calculations in the simulation code. 

While a modeller could write their own code to cut down on repetition when expanding a simple model (and many do), `macpan2` provides a ready-made model specification grammar that enables easy model extension, especially when building [product models](https://arxiv.org/abs/2307.10308), and that can readily interface with fast simulation and calibration engines, like [TMB](https://cran.r-project.org/web/packages/TMB/index.html). 

# Appetizer: specifying the basic SIR model

Let's start with specifying the basic SIR model, the foundation of the two-strain model above, in `macpan2`:

\begin{align}
S_{t+1} &= -\beta S_t I_t/N, \\ 
I_{t+1} &= \phantom{-} \beta S_t I_t/N - \gamma I_t, \\ 
R_{t+1} &=  \phantom{-} \gamma I_t.
\end{align}

It will be helpful to set $\lambda = \beta I/N$ and recast the equations as:

\begin{align}
S_{t+1} &= -\lambda S_t, \\ 
I_{t+1} &=  \phantom{-} \lambda S_t - \gamma I_t, \\ 
R_{t+1} &=  \phantom{-} \gamma I_t.
\end{align}

Since the focus of this quickstart guide is `macpan2`'s model specification grammar, we have defined an `SIR_starter()` function to sweep some of the details of initializing a model object under the rug (for now, though we will revisit it [later](#dessert)). All you need to know about `SIR_starter()` at this stage is that we will pass it some inputs to define the model using the model grammar and it will output a model object from which we can build a simulator. Our primary focus for the remainder of this vignette will be how the inputs to `SIR_starter()` are created.

```{r SIR-starter, echo = FALSE}
## helper function to simplify the exposition in this vigette -----------
SIR_starter <- function(
  # index tables for model quantities
  state,
  rate,
  # ledgers tabulating the use of different functional forms
  flow, # list of individual ledgers
  force_of_infection
){
  
  ## Set up expressions list for each functional form --------------
  ## names refer to when the calculation gets performed relative to 
  ## the simulation time-step loop (before, during, ...)
  ## FIXME: we should not be referring to a specific engine in
  ## a vignette about an 'engine-agnostic grammar'
  expr_list <- mp_tmb_expr_list(
    before = list(
      ## aggregations
        N ~ sum(state)
    ),
    during = list(
      ## force of infections
        rate[infection_flow_rates] ~
          state[infectious_states] * rate[transmission_rates] / N
  
      ## unsigned individual flows
      , flow_per_time ~ state[from_states] * rate[flow_rates]
  
      ## state update
      , total_inflow ~ group_sums(flow_per_time, to_states, state)
      , total_outflow ~ group_sums(flow_per_time, from_states, state)
      , state ~ state + total_inflow - total_outflow
    )
  )
  
  ## Ledgers for each specific calculation --------------
  ledgers <- list(
    flow = mp_ledgers(flow),
    force_of_infection = mp_ledgers(force_of_infection)
  )
  
  ## Initialize vectors from index tables (with all zeros for values) --------------
  # used as placeholders for user input
  init_vecs <- list(
    state = mp_structured_vector(state),
    rate = mp_structured_vector(rate)
  )
  
  ## Initialize model object -----------------
  mp_dynamic_model(
    expr_list = expr_list,
    ledgers = ledgers,
    init_vecs = init_vecs
  )
}
```

The inputs to `SIR_starter()` are of two types:

- **index tables** containing indices (labels) of model quantities, 
- **ledgers** that tabulate specific calculations required to simulate the model equations (based on the included functional forms).

The index tables we need to specify fall into two groups:

- `state`: state names, $S$, $I$, and $R$ from the model equations
- `rate`: rate names, $\beta$, $\gamma$, and the derived rate $\lambda$

We have identified two useful **functional forms** that we have baked into `SIR_starter()`. In this case, we're thinking of these forms not necessarily as repeated calculations in this particular model, but as calculations that a modeller may want to repeat down the line, as they expand this simple model with additional structure (as we will do [below](#main-course)). The forms are:

- **flow**: Unsigned flows from one class to another of the form $rX$, with $r>0$ being the *per capita* flow rate and $X$ being the occupancy of the state from which the flow originates. This calculation is repeated for all terms on the right-hand side of the recast system of difference equations above.
- **force of infection**: The prevalence-dependent *per capita* rate of flow from susceptible classes to infectious classes of the form $\lambda = \beta I /N$, used in calculating infection flows.

In this case, the flow form is repeated within these model equations, while the force of infection form is used only once. We've identified the force of infection as a functional form since we will want to repeat it later when [expanding into the two-strain model](#main-course). Either way, these forms are already baked into `SIR_starter()`, so our task will be creating a ledger for each of these forms to input into the function.  

We start by creating the `state` and `rate` index tables:

```{r sir-index-tables}
## index tables to label model quantities -------------------------
state <- mp_index(Epi = c("S", "I", "R"))
rate <- mp_index(Epi = c("beta", "gamma", "lambda"))
```

The `mp_index()` function sets structures like data frames that tabulate the model quantity labels:

```{r sir-state-and-rate}
state
rate
```

The `Epi` column name is unimportant in this simple model, but it will be key to stratifying model quantities with different features (such as epidemiological status, infection type, age group, location) in more complicated models.

For the flow form, we will create two ledgers: `infection` for the flow from $S$ to $I$ and `recovery` for the flow from $I$ to $R$ and then pass these as a list to the `flow` argument of `SIR_starter()`. We specify flows using the name of the state from which it originates (`from_states`), the state to which it goes (`to_states`), and a flow rate name (`flow_rates`).

We use the `mp_join()` function to create the `infection` ledger like so:

```{r sir-infection-ledger}
## infection ledger -------------------------
infection <- mp_join(
  from_states = mp_subset(state, Epi = "S"),
  to_states = mp_subset(state, Epi = "I"), 
  flow_rates = mp_subset(rate, Epi = "lambda")
)
```

The `mp_join()` function takes the options provided in each argument `from_states`, `to_states`, and `flow_rates`, e.g.

```{r sir-infection-ledger-inputs}
mp_subset(state, Epi = "S")
mp_subset(state, Epi = "I")
mp_subset(rate, Epi = "lambda")
```

and by default creates one entry in the ledger for each combination of these values (_i.e._, a [full join](https://dplyr.tidyverse.org/reference/mutate-joins.html#outer-joins)). However, since there is only one value in each column, there is only one entry in the resulting ledger:

```{r sir-infection-ledger-2}
infection
```

The names of the arguments in the `mp_join()` function are tied to how the functional form baked into `SIR_starter()` is specified, but in general modellers can define their functional forms and the corresponding `mp_join()` argument names however they like.^[There is only one `mp_join()` argument name that is not available to the user &mdash; `by`, which has a special role that we will see [later](#main-course).]

We create the `recovery` ledger in a similar way:

```{r sir-recovery-ledger}
## recovery ledger -------------------------
recovery  <- mp_join(
  from_states = mp_subset(state, Epi = "I"),
  to_states = mp_subset(state, Epi = "R"),
  flow_rates = mp_subset(rate, Epi = "gamma")
)

recovery
```

Finally, the `force_of_infection` ledger is slightly different as it corresponds to a different functional form in `SIR_starter()` (so the `mp_join()` argument names are different):

```{r sir-foi-ledger}
## force of infection ledger -------------------------
# infection additionally involves the calculation of a force of infection
force_of_infection <- mp_join(
  infectious_states = mp_subset(state, Epi = "I"),
  transmission_rates = mp_subset(rate, Epi = "beta"),
  infection_flow_rates = mp_subset(rate, Epi = "lambda")
)
```

For this functional form, we need to specify the `transmission_rates` and `infectious_states` involved in computing the force of infection, as well as the names where we want to store the results of this calculation (`infection_flow_rates`) for use in the `infection` flow calculations. 

Now we can use the `SIR_starter()` function to initialize our model object:

```{r sir}
## SIR model object -------------------------
sir <- SIR_starter(
  # index tables
  state = state,
  rate = rate,
  # ledgers
  flow = list(
    infection,
    recovery
  ),
  force_of_infection = force_of_infection
)
```

We can create a model simulator using `mp_dynamic_simulator()`^[`tmb` stands for "template model builder", the underlying simulation engine provided by the [TMB package](https://kaskr.github.io/adcomp/Introduction.html)], giving it the model object (`model`), initial values for each index (`vectors`), as well as the number of total time steps in the simulation (`time_steps`):

```{r sir-simulator}
## SIR model simulator -------------------------
sir_simulator <- mp_dynamic_simulator(
  dynamic_model = sir,
  vectors = list(
    state = c(S = 999, I = 1, R = 0),
    rate = c(beta = 0.25, gamma = 0.1)
  ),
  time_steps = 100
)
```

Note that we've specified `NA` for `lambda` as it will be calculated for us using the force of infection functional form.

Then we can actually simulate the model by passing our model simulator to `mp_trajectory()`:

```{r sir-results}
## SIR model simulation results -------------------------
sir_results <- mp_trajectory(sir_simulator)
```

The output of the simulation is a [long data frame](https://r4ds.had.co.nz/tidy-data.html#longer):

```{r sir-results-head}
head(sir_results)
```

The simulation output has several columns:

- `matrix`: The matrix storing our values internally, corresponding to our two index tables, `state` and `rate`.
- `time`: An internal time index, where `time = 1` is the result after the first step through the simulation loop. 
- `row`: The primary label for the `value` (the row name in the corresponding `matrix`).
- `col`: A secondary label for the `value` (the column name in the corresponding `matrix`). Since the outputs of this model (i.e. states and rates) are specified as vectors and not matrices, this column is empty for all entries. TODO: When would this be useful?
- `value`: The numerical value.

This output can be manipulated and plotted with standard tools, like `dplyr` and `ggplot2`, e.g.:

```{r sir-ggplot-example, fig.width = 6, fig.height = 4}
(sir_results
  |> filter(matrix == "state") # keep just the state variables at each point in time
  |> mutate(state = factor(row, levels = c("S", "I", "R"))) # to enforce logical state ordering in legend
  |> ggplot(aes(time, value, colour = state))
  +  geom_line()
)
```

(Above, we used the [base R pipe operator](https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/#pipes), `|>`.)

If you prefer to make plots in base R, you can convert the long format data to wide format:

```{r pivot_wider}
sir_results_wide <- (sir_results
    |> dplyr::filter(matrix == "state") # keep state variables at each point in time
    ## drop unneeded columns before pivoting
    |> dplyr::select(-c(matrix, col))
    |> tidyr::pivot_wider(id_cols = time, names_from = row)
)

head(sir_results_wide, n = 3)
```

We can plot one state like so

```{r sir-base-plot-ex, fig.width = 6}
with(sir_results_wide,
     plot(x = time,
          y = I,
          type = "l")
)
```

or multiple states on the same plot with

```{r sir-base-matplot-ex, fig.width = 6}
par(las = 1) ## horizontal y-axis ticks
matplot(sir_results_wide[, 1],
        sir_results_wide[,-1],
        type = "l",
        xlab = "time", ylab = "")
legend("left", col = 1:3, lty = 1:3, legend = state$labels())
```

# Main course: expanding the basic SIR with additional structure {#main-course}

As previously noted, we created a force of infection functional form ($\beta I / N$) despite it only being used once to define the SIR model. However, if we consider the two-strain model from [before](#amuse-bouche), we see this calculation is repeated for each strain:

\begin{align}
\lambda_A &= \beta_A I_A/N \\ 
\lambda_B &= \beta_B I_B/N
\end{align}

Since we already have a form for the force of infection, we can easily expand our basic SIR with the strain-related structure to get the two-strain SIR model.

To define the two-strain model, we again must specify our `state` and `rate` index tables, as well as our `infection`, `recovery`, and `force_of_infection` ledgers.

We start by creating a new set of indices for the strains:

```{r strain-indices}
Strain_indices <- c("A", "B")
```

A simple approach would be to define a table of the new state and rate indices directly using the `mp_index()` function, as we did above:

```
state <- mp_index(
  Epi = c("S", rep("I", 2), "R"),
  Strain = c("", Strain_indices, "")
)

rate <- mp_index(
  Epi = c(rep(c("beta", "lambda"), 2), "gamma"),
  Strain = c(rep(c("A", "B"), each = 2), "")
)
```

However, this approach is less flexible if we want to build a complex model or if we already have a simpler, working model (like the SIR above) and want expand it with many strata and/or several different types of strata. We present an alternative approach below that is more verbose but far more flexible.

For the state, we want to cross $I$ with the different strains to create one $I$ compartment name per strain. We can do so using the `mp_cartesian()` function, which takes the [Cartesian product](https://en.wikipedia.org/wiki/Cartesian_product) of indices (all possible combinations across sets)^[`mp_cartesian()` is analogous to `expand.grid()` in base R, or `tidyr::expand()` in the tidyverse]:

```{r strain-expand-I}
I_indices <- mp_cartesian(
  mp_subset(state, Epi = "I"),
  mp_index(Strain = Strain_indices)
)

I_indices
```

This table stores all indices associated with the $I$ compartment.^[In standard mathematical notation, one would typically write $I_A$ and $I_B$, with only $A$ and $B$ referred to as indices, but we've taken the abstraction one step further and chosen to refer to state variable names, like $I$, as indices as well. This choice was made deliberately when developing `macpan2` so that state variable names get treated like any other component used to label a particular compartment (like location or age group).]

We then combine the newly-stratified $I$ indices with the other states that remain unchanged using the `mp_union()` function to make a `state` index table:

```{r two-strain-state}
state <- mp_union(
  mp_subset(state, Epi = "S"),
  I_indices, 
  mp_subset(state, Epi = "R")
)

state
```

We update the `rate` index table similarly:

```{r two-strain-rate}
rate <- 
  mp_union(
  # stratify rates involved in the infection process by strain
  mp_cartesian(
    mp_subset(rate, Epi = c("beta", "lambda")),
    mp_index(Strain = Strain_indices)
  ),
  # recovery rate will be the same across strains
  mp_subset(rate, Epi = "gamma")
)

rate
```

For the `infection` ledger, let's see what our previous code for generating it yields now that we are (partially) stratifying by `Strain`:

```{r two-strain-infection-default}
# infection ledger from before
mp_join(
  from_states = mp_subset(state, Epi = "S"),
  to_states = mp_subset(state, Epi = "I"), 
  flow_rates = mp_subset(rate, Epi = "lambda")
)
```

As before, the default in `mp_join()` is to give all possible combinations for the indices (the full join), where the individual indices, denoted by values in the `Epi` and `Strain` columns, are dot-concatenated for the full quantity labels. 

For this model, we want only two of these flows:

- a flow between `S` and `I.A` with flow rate `lambda.A`
- a flow between `S` and `I.B` with flow rate `lambda.B`

In other words, we want the `Strain` index on `I` to match with the `Strain` index on `lambda`. We can specify this within `mp_join()` when building the ledger like so:

```{r two-strain-infection-ledger}
## new infection ledger -------------------------
infection <- mp_join(
  from_states = mp_subset(state, Epi = "S"),
  to_states = mp_subset(state, Epi = "I"), 
  flow_rates = mp_subset(rate, Epi = "lambda"),
  by = list(
    to_states.flow_rates = "Strain"
  )
)

infection
```

Note the syntax of the `by` argument here. Each `by` list element will correspond to a pairwise join of two of the index tables passed to `mp_join()`. Which indices are involved in the join will correspond to the dot concatenated list element name (`to_states.flow_rates`), with the names coming from `mp_join()`'s argument names (`to_states`, `flow_rates`). The list element value should be a character string corresponding to the index table column name upon which to perform matches. In this case, the value is `"Strain"` because we want the "to state" labels and the "flow rate" labels to match based on the `Strain` index table column (`I.A` with `lambda.A` and `I.B` with `lambda.B`).

For the recovery ledger, we haven't stratified `gamma` or `R`, so the default full join with the `I` labels yields exactly the flows we want:

```{r two-strain-recovery-ledger}
recovery <- mp_join(
    from_states = mp_subset(state, Epi = "I"),
    to_states = mp_subset(state, Epi = "R"),
    flow_rates = mp_subset(rate, Epi = "gamma")
)
recovery
```

For the force of infection ledger, the full join yields many combinations that we don't want:

```{r two-strain-foi-default}
mp_join(
  infection_flow_rates = mp_subset(rate, Epi = "lambda"),
  infectious_states = mp_subset(state, Epi = "I"),
  transmission_rates = mp_subset(rate, Epi = "beta")
)
```

We want the `lambda`, `I`, and `beta` labels all matched on the `Strain` column of the respective index tables. Internally, `mp_join()` performs pairwise joins, so we cannot specify a three-way `by` argument. Instead, we will specify two pairwise joins to the same effect:

```{r two-strain-foi-ledger}
## new force of infection ledger -------------------------
force_of_infection <- mp_join(
  infection_flow_rates = mp_subset(rate, Epi = "lambda"),
  infectious_states = mp_subset(state, Epi = "I"),
  transmission_rates = mp_subset(rate, Epi = "beta"),
  by = list(
    infection_flow_rates.infectious_states = "Strain", # first pairwise join
    infectious_states.transmission_rates = "Strain" # second pairwise join
  )
)

force_of_infection
```

Now we're ready to build the two-strain model object and simulate it:

```{r two-strain-results, fig.width = 6, fig.height = 4}
two_strain_model <- SIR_starter(
  # index tables
  state = state,
  rate = rate,
  # ledgers
  flow = list(
    infection,
    recovery
  ),
  force_of_infection = force_of_infection
)

two_strain_simulator <- mp_dynamic_simulator(
  dynamic_model = two_strain_model,
  vectors = list(
    state = c(S = 998, I.A = 1, I.B = 1, R = 0),
    rate = c(beta.A = 0.25, gamma = 0.1, beta.B = 0.2)
  ),
  time_steps = 100
)

two_strain_results <- (mp_trajectory(two_strain_simulator)
  |> filter(matrix == "state")                    
)

levels <- unique(two_strain_results$row) # get state variables in the desired order

(two_strain_results # keep state variables at each point in time
  |> mutate(state = factor(row, levels = levels)) # to enforce logical state ordering in plot
  |> ggplot(aes(time, value, colour = state))
  +  geom_line()
)
```

# Dessert: understanding model simulation in `macpan2` {#dessert}

As mentioned, we've hidden some of the details of initializing a model object within the `SIR_starter()` function: 

```{r sir-starter-print}
<<SIR-starter>>
```

This function definition shows how all the pieces fit together. The expressions list `expr_list` is perhaps the most interesting as it contains all of the functional forms used to simulate the model, including some we explored above (unsigned flows, force of infection), as well as some that we didn't discuss (total inflow, total outflow, state update). The `ledgers` and `init_vecs` are just set up to ensure that the ledgers and initial conditions for simulation get attached to the model object correctly.

These topics will be discussed fully in a future vignette.

