---
title: "Hello Products"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Hello Products}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

[![status](https://img.shields.io/badge/status-working%20draft-red)](https://img.shields.io/badge/status-working%20draft-red)

```{r, include = FALSE}
library(macpan2)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
cat_file = function(...) {
  cat(readLines(file.path(...)), sep = "\n")
}
prod_dir = file.path("..", "inst", "starter_models", "SI_products", "hello_products")
si_dir = file.path("..", "inst", "starter_models", "SI_products", "hello_si")
age_dir = file.path("..", "inst", "starter_models", "SI_products", "hello_age")
```

## Read Models

Here we read the target product model, `prod`, and its two factor models, `si` and `age`, into R.

```{r}
prod = Compartmental(prod_dir)
si = Compartmental(si_dir)
age = Compartmental(age_dir)
```

The goal of this document is to multiply `si` and `age` to see if we can recover our target, `prod`.

## Types of Variables

Each Model has the following set of variables.

```{r}
si$variables$all()
age$variables$all()
prod$variables$all()
```

These variables are categorized into various types that distinguish their roles in the dynamics.

The state variables represent the nodes in the compartmental diagram.

```{r}
si$variables$state()
age$variables$state()
prod$variables$state()
```

The model definition also recognizes two subsets of these states -- infectious and infected -- which in these models are identical because there is no latency period.

```{r}
si$variables$infectious_state()
age$variables$infectious_state()
prod$variables$infectious_state()
```

```{r}
si$variables$infected_state()
age$variables$infected_state()
prod$variables$infected_state()
```

Note that the `age` model has no infection, and so these types of states are not present.

The flow variables represent the flows between states in the compartmental diagram.

```{r}
si$variables$flow()
age$variables$flow()
prod$variables$flow()
```

And a subset of flows are those involved in infection.

```{r}
si$variables$infection_flow()
age$variables$infection_flow()
prod$variables$infection_flow()
```

But what about other variables that are neither states nor flows? We can list these other variables.

```{r}
si$variables$other()
age$variables$other()
prod$variables$other()
```

We see that the only variables in the `age` model are states and flows. But we also see that the `si` and `prod` models have a `transmission` variable.


## Multiplying Variables

We can recover each kind of variable in the `prod` model by operating on the factor models, `si` and `age`. The state variables in `prod` are a Cartesian product of the state variables in `si` and `age`.

```{r}
cartesian(si$variables$state(), age$variables$state())
prod$variables$state()
```

The flow variables in `prod` are a union of two Cartesian products.

```{r}
union_vars(
  cartesian(si$variables$flow(), age$variables$state()),
  cartesian(si$variables$state(), age$variables$flow())
)
prod$variables$flow()
```

The other variable in `prod`, `transmission.`, can be obtained by taking the union of the other variables in `si` and `age`.

```{r}
union_vars(si$variables$other(), age$variables$other())
prod$variables$other()
```

Putting this all together we have.

```{r}
union_vars(
  cartesian(si$variables$state(), age$variables$state()),
  cartesian(si$variables$flow(), age$variables$state()),
  cartesian(si$variables$state(), age$variables$flow()),
  si$variables$other(),
  age$variables$other()
)
prod$variables$all()
```

## Multiplying Flows

```{r}
si_flows = si$flows()
age_flows = age$flows()
```

```{r}
si_flows$from_to_partition = "Age"
si_flows$from_flow_partition = "Age"
age_flows$from_to_partition = "Epi"
age_flows$from_flow_partition = "Epi"
prod_flows = rbind(si_flows, age_flows)
prod_flows
prod$flows()
```

## Combining the Transmission Matrices

## Combining the Derivations

## Simulating the Resulting Product Model


## Roadmap

Here we outline what development needs to take place in order to complete this document.

* Add transmission matrix step to `StandardExprs`
* Align transmission matrix specifications
* Pass required indices for lining up the transmission matrix to the flow vector and state vector.

## Scratch


```{r}
prod$labels$state()
prod$labels$flow()
```

```{r}
si$variables$all()
si$flows_expanded()
```

```{r}
age$variables$all()
age$flows_expanded()
```


```{r, eval=FALSE}
prod$variables()
prod$flows_expanded()
UserExpr(prod)$expand_vector_expressions()
prod$expr_list()$print_exprs()
```


## Simulating from the Product Model

* Do the model definition by hand so that it works
* Specify numerical inputs
* Run the the simulation
* Verify the results
