---
title: "Quickstart"
output: 
  html_document:
    toc: true
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

[![status](https://img.shields.io/badge/status-stable-green)](https://canmod.github.io/macpan2/articles/vignette-status#stable)

## Hello World

This vignette shows how to simulate from the SI model, which is I think is the simplest possible model of epidemiological transmission. Here is the code for defining an SI model (see [this short article](https://canmod.github.io/macpan2/articles/example_models) on how to find other example models).
```{r hello-world, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
si = mp_tmb_model_spec(
    before = S ~ N - I
  , during = mp_per_capita_flow(
        from     = "S"            # compartment from which individuals flow
      , to       = "I"            # compartment to which individuals flow
      , rate     = "beta * I / N" # expression giving per-capita flow rate
      , abs_rate = "infection"    # name of absolute flow rate = beta * I * S/N
    )
  , default = list(N = 100, I = 1, beta = 0.25)
)
print(si)
```

Simulating from this model requires choosing the number of time-steps to run and the model outputs to generate. Syntax for simulating `macpan2` models is designed to combine with standard data prep and plotting tools in R, as we demonstrate with the following code.
```{r plot-tmb-si, fig.height=3}
library(ggplot2)
library(dplyr)
si_plot = (si
 
 ## macpan2
 |> mp_simulator(
        time_steps = 50
      , outputs = c("I", "infection")
    )
 |> mp_trajectory()
 
 ## dplyr
 |> mutate(quantity = case_match(matrix
      , "I" ~ "Prevalence"
      , "infection" ~ "Incidence"
    ))
 
 ## ggplot2
 |> ggplot() 
 + geom_line(aes(time, value)) 
 + facet_wrap(~ quantity, scales = "free")
 + theme_bw()
)
print(si_plot)
```

(Above, we used the [base R pipe operator](https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/#pipes), `|>`.)

The remainder of this article looks at each step required to create this plot in more detail, and discusses alternative approaches.

## Creating a Simulator

The first step is to produce a simulator object, which can be used to generate simulation results. This object can be produced using the `mp_simulator` function, which takes the following arguments.

- `model` : A model specification object, such as `si`.
- `time_steps`: How many time steps should the epidemic simulator run for? 
- `outputs` : The model variables to return in simulation output. Functions `mp_state_vars()` and `mp_flow_vars()` can be used to programmatically output all state or flow variables respectively.
- `default` (optional) : Allows one to update the default parameter values and initial conditions provided in the model specification (see the argument `default` above in the `mp_tmb_model_spec` function). Any variables that are not specified in `default` will be left at the values in the specification.

```{r tmb-si-simulator}
si_simulator = mp_simulator(
    model = si 
  , time_steps = 50
  , outputs = c("I", "infection")
)
si_simulator
```

This `si_simulator` object contains all of the information required to generate model simulations of `I` and `infection` over 50 time steps, without actually generating the simulations. This more interesting step of simulation is covered in [the next section](#generating-simulations). But before moving on we explain why we separate the step of creating a simulator from the step of running simulations.

The reason for two steps is to optimize performance in more computationally challenging applications of the software than those covered in this article. The step of creating the simulator object is more computationally intensive than the next step of actually generating the simulations. Therefore this separation is useful when a single simulator can be used to repeatedly generate many different simulations. Three common examples requiring repeated simulations from the same simulator are:

* for models with stochasticity so that the output is different for each run,
* for calibrating model parameters to data using iterative optimization tools,
* and for running different scenarios by updating certain parameters before simulation.

Because `macpan2` is primarily developed for such computationally challenging iterative simulation problems, we believe that it is better to introduce these two steps as distinct from the very beginning. Although two steps might seem unnecessarily complex within the context of this article, keeping them separate will make life easier when working with more realistic workflows and models. 

## Generating Simulations

Now that we have a simulation engine object,  `si_simulator`, we use it to generate simulation results using the `mp_trajectory()` function. The results come out in [long (or "narrow") format](https://en.wikipedia.org/wiki/Wide_and_narrow_data), where there is exactly one value per row:

```{r run_sims}
si_results = mp_trajectory(si_simulator)
si_results |> head(8)
```

The simulation results are output as a data frame with the following columns:

- `matrix`: Which matrix does a value come from? All variables are represented as matrices, although in this article we only consider 1-by-1 matrices.
- `time`: The time index from `1` to `time_steps`.
- `row`, `col`: Placeholders for variable components in more complicated structured models (not covered in this article, but useful for example when S and I in different age groups or geographic locations are tracked separately).
- `value`: The simulated value for a particular state and time step.

## Processing Results

`macpan2` does not provide any data manipulation or plotting tools (although there are a few in `macpan2helpers`). The philosophy is to focus on the engine and modelling interface, but to provide outputs in formats that are easy to use with other data processing packages, like `ggplot2`, `dplyr`, and `tidyr`, all of which readily make use of data in long format.

In the graphs above for example, we required a step that renames the model variables into something that makes more sense for the graphical presentation. Here we reproduce this, but take a little more care to produce a tidier dataset.
```{r rename-variables}
si_results = (si_results
  |> mutate(matrix = case_match(matrix
      , "I" ~ "Prevalence"
      , "infection" ~ "Incidence"
    ))
  |> rename(quantity = matrix)
  |> select(time, quantity, value)
)
print(head(si_results, 8L))
```

Here we rename the `I` state variable as the disease prevalence (number of currently infected individuals). Because this is a discrete-time model, we can reinterpret the `infection` as the incidence (number of newly infected individuals) over one time step. Note that this approach to calculating incidence only works if the time period over which incidence is measured corresponds to the length of one time step. If this assumption is not met -- for example if the time step is one day but incidence data are reported every week -- then other approaches to computing incidence that are not covered here must be taken.

We can generate a `ggplot` from this dataset as we did above. If you want to use base R plots, you can convert the long format data to wide format using the `tidyr` package.
```{r pivot_wider}
library(tidyr)
si_results_wide <- (si_results
  |> tidyr::pivot_wider(
      , id_cols = time
      , names_from = quantity
    )
  |> rename(Time = time)
)
head(si_results_wide, n = 3)
```

```{r base_plot_ex, fig.width = 6}
with(si_results_wide,
     plot(x = Time,
          y = Incidence)
)
```

or

```{r base_matplot_ex, fig.width = 6}
par(las = 1) ## horizontal y-axis ticks
matplot(
    si_results_wide[, 1]
  , si_results_wide[,-1]
  , type = "l"
  , xlab = "Time", ylab = ""
)
legend("topleft"
  , col = 1:3
  , lty = 1:3
  , legend = c("Prevalence", "Incidence")
)
```

## Testing Mermaid

```mermaid
flowchart LR
S --> I
I --> R
```
