---
title: "Quickstart Guide, part 2: specifying and simulating a structured compartmental model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart Guide, part 2: specifying and simulating a structured compartmental model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[![status](https://img.shields.io/badge/status-stub-red)](https://canmod.github.io/macpan2/articles/vignette-status#stub)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
cat_file = function(...) cat(readLines(file.path(...)), sep = "\n")
```

```{r setup}
library(macpan2)
library(dplyr)
library(ggplot2)
library(tidyr)
```

This article is the counterpart to `vignette('quickstart', package = 'macpan2')`, but instead of explaining the spec and sim of a simple SIR, we describe the **new** components introduced by trying to specify a product model. While the new "productify" functionality should make most of the content of this article obsolete eventually, this article can serve as a resource for users until then.

## SIR Vaccination model

A good place to start with product models is the ``sir_vax`` starter model which stratifies a standard SIR model to include vaccination status.

```{r sir_vax_model_path}
print(sir_vax_dir <- system.file("starter_models", "sir_vax", package = "macpan2"))
```

As in the case of a simple ``SIR`` model the population is divided into compartments for susceptible, infected, and recovered individuals. However the population is **also** divided into compartments for vaccinated and unvaccinated individuals. So where a simple ``SIR`` model has three compartments the ``SIR_vax`` model has six. This is reflected in the ``variables.csv`` file in the model directory which now has two columns. 

```{r vars, echo = FALSE, eval = TRUE}
cat_file(sir_vax_dir, "variables.csv")
```
The row of the file specifies the name of each column; so the first column is called ``Epi`` and the second column is called ``Vax``. predictably the ``Epi`` column lists compartment and parameter names relevant to epidemiological status and the ``Vax`` column similarly relates to vaccination status. Some parameters (e.g. ``gamma``) which in an ``SIR`` model have a single value now have two values to reflect the differences in behavior in vaccinated and unvaxxinated people. Other parameters (e.g. ``vax_rate``) have only a single value (in this case because only ``susceptible`` people will be vaccinated). Epi parameters that vary based on vax status will have the parameter name repeated in the ``Epi`` column but will have differing status labels in the ``Vax`` column. When referring to specific compartments or parameters we concatenate their labels in each column with a ``.``, so the compartment for the unvaccinated susceptible population is called ``S.unvax``. If a variable has no label in a given column then that entry is left blank but the concatonation dot is still included (so ``vax_rate`` is properly referred to as ``.vax_rate``).


Flows in product models are specified by the ``flows.csv`` file in the model definition directory. 
```{r flows, echo = FALSE, eval=TRUE}
cat_file(sir_vax_dir, "flows.csv")
```
Notice that the three right most columns, which unused in single stratum models now have an important role. In particular they are there to ensure that compartments of one stratum flow into other compartments of the same stratum. Without them the ``S.vax`` compartment would have flows both to ``I.vax`` and `` I.unvax``. There are three such columns ``from_to_partition``, ``from_flow_partition`` and ``to_flow_partition``, in general two of these columns should have entries and the third should be blank or have the ``null_partition`` label specified in ``settings.json``. The ``from_to_partition`` column indicates which partitions should be used to math ``from`` and ``to`` compartments. In the example the flow from ``S`` to ``I`` indicates the ``Vax`` partition should be used so ``S.vax`` flows to ``I.vax`` and ``S.unvax`` to ``I.unvax``. The ``from_flow_partition`` column indicates which partition should be used to match ``from`` compartments with flow variables. In this example the flow from ``I`` to ``R`` has ``Vax`` in the ``from_flow_partition`` so the flow from ``I.unvax`` to ``R.unvax`` is matched with the ``gamma.unvax`` variable. In principle the ``to_flow_partition`` column can be used to match flow parameters to flow via their ``to`` compartment rather than their ``from`` compartment however in this example we have used the other two columns so the ``to_flow_column`` has a ``Null`` entry. Notice that the flow from ``S.unvax`` to ``S.vax`` is different from the other flows because there are no corresponding flows from ``I.unvax`` to ``I.vax`` or from ``R.unvax`` to ``R.vax``. In this case the ``from_partition`` and the ``to_partition`` are both ``Epi.Vax`` because the ``from`` and ``to`` compartments are specified using both partitions, the ``flow_partition`` is specified as ``Vax`` because the rate of vaccination is governed by the ``.dose_rate`` variable which has no entry in the ``Epi`` partition. There is no need to use the final three columns since the compartments involved in the flow are given explicitly rather than as a group as was the case for the other flows. 


 
```{r derivations, echo = FALSE, eval = TRUE}
cat_file(sir_vax_dir, "derivations.json")
```
The ``derivations.json`` file from the model definition directory largely the same as it would be for single stratum models. On key distinction is that now each derivation can correspond to multiple different equations, this is reflected in the existence of multiple entries in the ``group_names`` as well as ``output_names`` fields. If we take the first derivation in the above file as an example we can see that the expression being evaluated is ``sum(S, I, R)``. The ``group_names`` field has entries ``unvax`` and ``vax`` (the ``group_partition`` field defines which partition the ``group_names`` are related to). The ``output_names`` field also has two entries ``N.unvax`` and ``N.vax`` which are the names of the variables this derivation will compute values for. Taken together we see that this single derivation produces two distinct equations,  ``N.unvax = sum(I.unvax, S.unvax, R.unvax)`` and ``N.vax = sum(S.vax, I.vax, R.vax)``.


```{r settings, echo = FALSE, eval= TRUE}
cat_file(sir_vax_dir, "settings.json")
```

The only notable difference between the ``settings.json`` files for single stratum and multi-strata models is that multi-strata models will have multiple ``required_partitions``. It's also worth noting that the ``null_partition`` entry defines what should be entered in whichever of the ``from_to_partition``, ``from_flow_partition`` and ``to_from_partition`` isn't being used in the ``flows.csv`` file.


```{r example_model}
sir_vax = Compartmental(sir_vax_dir)
```








```{r sim_values}
sir_vax_sim = sir_vax$simulators$tmb(time_steps = 100L
  , state = c(S.unvax = 99, I.unvax = 1, R.unvax = 0, S.vax = 0, I.vax = 0, R.vax = 0)
  , flow = c(
        infection.unvax = 0, infection.vax = 0
      , gamma.unvax = 0.1, gamma.vax = 0.1
      , .vax_rate = 0.1
    )
  , sigma.unvax = 1
  , sigma.vax = 0.01
  , beta.unvax = 0.2
  , beta.vax = 0.2
  , foi.unvax = empty_matrix
  , foi.vax = empty_matrix
  , foi. = empty_matrix
  , N.unvax = empty_matrix
  , N.vax = empty_matrix
)
```

```{r simulations}
(sir_vax_sim$report() 
 %>% separate_wider_delim("row", ".", names = c("Epi", "Vax"))
 %>% mutate(Epi = factor(Epi, levels = c("S", "I", "R")))
 %>% ggplot()
 + facet_grid(Epi~Vax, scales = "free")
 + geom_line(aes(time, value))
)
```
