---
title: "Specifying Time-Varying Parameters"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Specifying Time-Varying Parameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

[![status](https://img.shields.io/badge/status-working%20draft-red)](https://canmod.github.io/macpan2/articles/vignette-status#working-draft)

```{r pkgs, include = FALSE}
library(macpan2)
library(macpan2helpers)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Baseline SIR Model

Here we modify an [SIR](https://canmod.github.io/macpan2/articles/quickstart) model so that transmission rate is time-varying.

```{r baseline_sir}
sir = Compartmental(system.file("starter_models", "sir", package = "macpan2"))
simulator = sir$simulators$tmb(time_steps = 50
  , state = c(S = 99, I = 1, R = 0)
  , flow = c(foi = 0, gamma = 0.2)
  , N = empty_matrix
  , beta = 0.8
)
(simulator$report(.phases = "during")
  %>% rename(state = row)
  %>% mutate(state = factor(state, sir$labels$state()))
  %>% ggplot() + geom_line(aes(time, value, colour = state))
)

```

## Piecewise Time Variation

We now change the value of the transmission rate, `beta`, at the beginning of time-step 10 and 15. In the first step we add to the simulator a vector containing these change-points.
```{r change_points}
simulator$add$matrices(beta_changepoints = c(0, 10, 15))
```

Next we add the values to which `beta` changes at these time-steps.
```{r change_values}
simulator$add$matrices(beta_values = c(0.8, 0.01, 0.4))
```

We also need a variable to track the current value of `beta`. This `beta_pointer` starts at time-step equal to 0, and it will be incremented throughout the simulation.
```{r change_pointer}
simulator$add$matrices(beta_pointer = 0)
```

We increment `beta_pointer` using the [`time_group` function](https://canmod.github.io/macpan2/reference/engine_functions.html#time-indexing) that returns either `beta_pointer` or `beta_pointer + 1` depending on whether or not the current time-step is at a change-point in `beta_changepoints`.
```{r update_pointer}
simulator$insert$expressions(
    beta_pointer ~ time_group(beta_pointer, beta_changepoints), 
    .phase = "during"
)
```

We update `beta` at every iteration of the simulation loop using this `beta_pointer`.
```{r update_beta}
simulator$insert$expressions(
  beta ~ beta_values[beta_pointer],
  .phase = "during"
)
```

And that's it.  Now we plot the updated simulations using these change-points, which we highlight with vertical lines.
```{r time_varying_graph}
s = simulator$report(.phases = "during")
(s
  %>% rename(state = row)
  %>% mutate(state = factor(state, sir$labels$state()))
  %>% ggplot()
  + geom_line(aes(time, value, colour = state))
  + geom_vline(
    aes(xintercept = x), 
    linetype = "dashed", 
    alpha = 0.5, 
    data = data.frame(x = simulator$get$initial("beta_changepoints"))
  )
)
```

## Calibrating Time Variation Parameters

First we simulate data to fit our model to, to see if we can recover the time-varying parameters.
```{r noisy_data}
set.seed(1L)
I_observed = rpois(
  50, 
  filter(s, matrix == "state", row == "I")$value
)
plot(I_observed)
```

Then we add a few matrices to the model for keeping tracking of information used in model fitting.
```{r add_matrices}
simulator$add$matrices(
  
  ## observed data
  I_obs = I_observed,
  
  ## simulated trajectory to compare with data
  I_sim = empty_matrix, 
  
  ## location of I in the state vector
  ## (the `-1L` bit is to get 0-based indices instead of 1-based)
  I_index = match("I", sir$labels$state()) - 1L, 
  
  ## matrix to contain the log likelihood values at 
  ## each time step
  log_lik = empty_matrix, 
  
  ## need to save the simulation history of each of these matrices
  .mats_to_save = c("I_sim", "log_lik")
)
```

Now we need some new expressions. The first expression pulls out the I state from the state vector.
```{r pull_out_trajectory}
simulator$insert$expressions(
  I_sim ~ state[I_index],
  .phase = "during"
)
```

The second expression computes a vector of Poisson log-likelihood values -- one for each time step.
```{r log_lik}
simulator$insert$expressions(
  log_lik ~ dpois(I_obs, clamp(rbind_time(I_sim))),
  .phase = "after"
)
simulator$replace$obj_fn(~ -sum(log_lik))
```

Next we declare the beta values as parameters to be optimized on the log scale.
```{r params}
simulator$add$transformations(Log("beta_values"))
simulator$replace$params(
  default = log(mean(simulator$get$initial("beta_values"))),
  mat = rep("log_beta_values", 3L),
  row = 0:2
)
```

Finally we fit the model back to the simulation data.
```{r optimize}
simulator$optimize$nlminb()
```

We can see that the optimizer converges (i.e. `$convergence = 0`) in 26 iterations.

On the log scale we see that the optimizer finds different values (`current`) than it started at (`default`).
```{r log_parameters}
simulator$current$params_frame()
```

More importantly the beta values on the untransformed scale recover to the values used in the simulations, although the second fitted beta value is much smaller than the true value.
```{r recovered_parameters}
data.frame(
  fitted = formatC(
    exp(simulator$current$params_frame()$current),
    format = "e", digits = 2
  ),
  true = simulator$get$initial("beta_values")
)
```


## Smooth Time Variation (TODO)


## Generalized Linear Models (TODO)



## Radial Basis Functions for Flexible Time Variation (In-Progress)

## Getting and the Base Model

Before we can add the fancy radial basis for the transmission rate, we need a base model. We use an SIR model that has been modified to include waning.

```{r}
sir = Compartmental(system.file("starter_models", "sir_waning", package = "macpan2"))
sir$flows()
```


## Radial Basis Functions (In-Progress)

This section uses radial basis functions to generate models with a flexible functional form for smooth changes in the transmission rate.

The `macpan2::rbf` function can be used to produce a matrix giving the values of each basis function (each column) at each time step (each row). Using this matrix, $X$, and a weights vector, $b$, we can get a flexible output vector, $y$, with a shape that can be modified into a wide variety of shapes by changing the weights vector.

$$
y = Xb
$$

The following code illustrates this approach.

```{r, fig.height=8, fig.width=6}
set.seed(1L)
d = 20
n = 2500
X = rbf(n, d)
b = rnorm(d, sd = 0.01)
par(mfrow = c(3, 1))
plot(c(1, n), range(X), type = "n",
     xlab = "time", ylab = "basis functions")
for (i in 1:d) lines(X[,i])
barplot(b, xlab = "", ylab = "weights")
plot(X %*% b, type = "l", xlab = "time", ylab = "output")
```

Here `d` is the dimension of the basis, or number of functions, and `n` is the number of time steps. By multiplying the uniform basis matrix (top panel) to a set of weights (middle panel), one can obtain a non-uniform curve (bottom panel). Note how the peaks in the output are associated with large positive weights.

To transform the output of the matrix multiplication of the radial basis function matrix and the weights vector into a time-series for the transmission rate, $\beta$. Although we could just use the output vector as the $\beta$ time series, it is more convenient to transform it a little so that the $\beta$ values yield more interesting dynamics in an SIR model. In particular, our model for $\beta_t$ as a function of time, $t$, is as follows.

$$
\log(\beta_t) = \log(\gamma_t) + \log(N) - \log(S_t) + x_tb
$$

Here we have the recovery rate, $\gamma_t$, total population, $N_t$, and number of susceptibles, $S_t$, at time, $t$, and the $t$th row of $X$, $x_t$. To better understand the rationale for this equation note that if every element of $b$ is set to zero, we have the following condition.

$$
\frac{\beta_t S_t}{N} = \gamma_t
$$

This condition assures that the number of infected individuals remains constant at time, $t$. This means that positive values of $b$ will tend to generate outbreaks and negative values will tend to reduce transmission.

```{r}
set.seed(1L)
simulator = sir$simulators$tmb(
      time_steps = n
    , state = c(S = 100000 - 500, I = 500, R = 0)
    , flow = c(foi = 0, gamma = 0.2, wane = 0.01)
    , beta = 1
    , N = 100000
    , X = rbf(n, d)
    , b = rnorm(d, sd = 0.01)
    , incidence = empty_matrix
    , eta = empty_matrix
    , .mats_to_save = c("state", "incidence", "beta")
    , .mats_to_return = c("state", "incidence", "beta")
)$insert$expressions(
    eta ~ gamma * exp(X %*% b)
  , .phase = "before"
  , .at = Inf
)$insert$expressions(
    beta ~ eta[time_step(1)] / clamp(S/N, 1/100)
  , .phase = "during"
  , .at = 1
)$insert$expressions(
    incidence ~ I
  , .vec_by_states = "total_inflow"
  , .phase = "during"
  , .at = Inf
)$replace$params(
    default = rnorm(d, sd = 0.01)
  , mat = rep("b", d)
  , row = seq_len(d) - 1L
)
```

```{r, fig.height=8, fig.width=6}
set.seed(5L)
(simulator$report(rnorm(d, sd = 0.01), .phases = "during")
  %>% mutate(variable = if_else(matrix == "state", row, matrix))
  %>% ggplot()
  + facet_wrap(~ variable, ncol = 1, scales = 'free')
  + geom_line(aes(time, value))
)
```

## Generalized Linear Mixed Models (In-Progress)

```{r, eval=FALSE}
#set.seed(6L)
sir = Compartmental(system.file("starter_models", "sir", package = "macpan2"))
simulator_rand_beta = sir$simulators$tmb(time_steps = 50
  , state = c(S = 99, I = 1, R = 0)
  , flow = c(foi = 0, gamma = 0.2)
  , N = empty_matrix
  , beta = 0.4
)
simulator_rand_beta$add$matrices(
  beta_eps = norm_beta,
  beta_mean = log(0.3)
)
simulator_rand_beta$insert$expressions(
    beta ~ 1 / (1 + exp(-beta_mean - beta_eps))
  , .phase = "during"
  , .at = 1
)
(simulator_rand_beta$report(.phases = "during")
  %>% filter(row == "I", matrix == "state")
  %>% ggplot()
  + facet_wrap(~ matrix, ncol = 1, scales = 'free')
  + geom_line(aes(time, value))
)
observed_I = filter(simulator_rand_beta$report(.phases = "during"), row == "I", matrix == "state")$value
```

```{r eval = FALSE}
sir = Compartmental(system.file("starter_models", "sir", package = "macpan2"))
simulator_step_level_re = sir$simulators$tmb(time_steps = 50
  , state = c(S = 99, I = 1, R = 0)
  , flow = c(foi = 0, gamma = 0.2)
  , N = empty_matrix
  , beta = 0.8
)
simulator_step_level_re$add$matrices(
    beta_mean = log(0.5)
  , beta_sd = 1
  , beta_time_step = rnorm(50, 0, 0.2)
  , beta_log_density = empty_matrix
  , I_observed = observed_I
  , log_lik = empty_matrix
  , total_log_lik = empty_matrix
  , .mats_to_save = "log_lik"
)
simulator_step_level_re$insert$expressions(
    beta_log_density ~ sum(dnorm(
      beta_time_step, 
      0, 
      beta_sd
    ))
  , .phase = "before"
)
simulator_step_level_re$insert$expressions(
    beta ~ exp(beta_mean + beta_time_step[time_step(1)])
  , .phase = "during"
)
simulator_step_level_re$insert$expressions(
    log_lik ~ dpois(I_observed[time_step(1)], clamp(I))
  , .phase = "during"
  , .at = Inf
)
simulator_step_level_re$insert$expressions(
    total_log_lik ~ sum(rbind_time(log_lik))
  , .phase = "after"
)
simulator_step_level_re$replace$params(
  default = c(0, 1),
  mat = c("beta_mean", "beta_sd")
)
simulator_step_level_re$replace$random(
  default = rnorm(50, 0, 1),
  mat = rep("beta_time_step", 50),
  row = 0:49
)
simulator_step_level_re$replace$obj_fn(
  ~ -total_log_lik-beta_log_density-dnorm(beta_sd, 0.2, 0.01)
)
simulator_step_level_re$optimize$nlminb()
simulator_step_level_re$optimize$optim()
#simulator_step_level_re$current$random_frame()
#simulator_step_level_re$current$params_frame()
#simulator_step_level_re$ad_fun()$fn(0)
ggplot(simulator_step_level_re$current$random_frame()) +
  geom_line(aes(row, current))
```


```{r eval = FALSE}
#simulator_step_level_re$cache$invalidate()
#simulator_step_level_re$ad_fun()$gr(-1.23)
plot(
  filter(simulator_step_level_re$report(.phases = "during"), row == "I", matrix == "state")$value,
  observed_I
)
abline(0, 1)
```
