---
title: "usecases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{usecases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(macpan2)
```

## List of Use-Cases

Out-of-context it is easy to write down the use-cases, so here they are.  The real (combinatorial) difficulties come when use-cases are mixed in ways that were not considered.

* Flow definition
  * Define unstructured flows
  * Define product model flows
  * Edit product model flows
* Parameter variation
  * Define time-variation of parameters
  * Define product model parameter variation
* State variable initialization
  * Default state
  * Eigenvector
* Integration
  * Time step unit
  * Hazard step
  * ODE solvers
* Simulation
  * Deterministic
  * Parameter uncertainty
  * Observation error
  * Process error
  * Include initial state
  * Include time dimension
* Trajectory processing 
  * Lagged differencing
  * Uneven lagged differencing
  * Convolutions
* Calibration
  * Likelihood
  * Priors
  * Parameter transformations
  
  
  
## Existing Tensions

### Observation error and parameter uncertainty


