---
title: "SIR with demography"
index_entry: "An SIR model with birth and death"
bibliography: ../../references.bib
link-citations: TRUE
author: Jen Freeman, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

This is an extension of the [sir](https://github.com/canmod/macpan2/tree/main/inst/starter_models/sir) model to include birth and death.

# Packages Used

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```

# Model Specification

This model has been specified in the `sir_demog` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_demog/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
spec = mp_tmb_library(
    "starter_models"
  , "sir_demog"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagram using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_demog/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 2, fig.width = 8}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_paths(spec)
(layout
  |> plot_flow_diagram(show_flow_names = TRUE)
  |> draw_inflows(layout, show_labels = TRUE)
  |> draw_outflows(layout, show_labels = TRUE, pattern_mutate = "^([a-z]*)(_.*)", pattern_replace = "\\1")
)
```


# States

| variable | description                       |
| -------- | --------------------------------- |
| S        | Number of susceptible individuals |
| I        | Number of infectious individuals  |
| R        | Number of recovered individuals   |

The size of the total population is, $N = S + I + R$.

# Parameters

| variable | description                  |
| -------- | ---------------------------- |
| $\beta$  | per capita transmission rate |
| $\gamma$ | per capita recovery rate     |
| $\nu$    | per capita birth rate        |
| $\mu$    | per capita mortality rate    |

The SIR model with demography often assumes that the time scale of epidemic changes is much shorter than demographic changes [@earn2008light]. This translates to a constant population size $N$ over time, with $\nu = \mu$. We parameterize birth and mortality rates separately to allow for the general case in which epidemic and demographic dynamics occur on similar time scales. 

# Dynamics 

We assume new individuals (births) join the susceptible compartment, and individuals can leave the population (die) from any compartment.

$$
\begin{align*}
\frac{dS}{dt} &= \nu N -\beta S\frac{I}{N} - \mu S \\
\frac{dI}{dt} &= \beta S\frac{I}{N} - \gamma I - \mu I \\
\frac{dR}{dt} &= \gamma I - \mu R
\end{align*}
$$


# Calibration Example

## Simulate fake data

We modify the specification so that it is different from the default library model that we will calibrate. We will use the [`mp_rk4`](https://canmod.github.io/macpan2/reference/mp_euler.html) ODE solver.
```{r defaults}
spec_for_making_fake_data = mp_tmb_insert(
    spec |> mp_rk4()
  , default = list(mu = 0.1, beta = 0.4)
)
```

We simuate 100 time steps of this model.
```{r simulate}
time_steps = 100L
sim = mp_simulator(  
    model = spec_for_making_fake_data
  , time_steps = time_steps
  , outputs = "infection"
)

# simulate data (known 'true' trajectory)
true_traj = obs_traj = mp_trajectory(sim)

# add noise (simulated observed and noisy trajectory)
set.seed(1L)
obs_traj$value = rpois(time_steps, true_traj$value)
```

## calibrate to fake data

We fit `beta` and `mu` to the simulated infection flow (i.e., incidence). To be consistent we use the [`mp_rk4`](https://canmod.github.io/macpan2/reference/mp_euler.html) ODE solver.
```{r calibrator}
cal = mp_tmb_calibrator(mp_rk4(spec), obs_traj, "infection", c("beta", "mu"))
```

Before calibrating, we capture the 'default' trajectory that we would simulate before our model is calibrated (represents ignorance). This is useful to show that calibration 'did something'.
```{r default_traj}
default_traj = mp_trajectory(cal)
```

Calibrate the model and check for convergence (convergence = 0 is good).
```{r optimize}
mp_optimize(cal)
```

## Explore the calibration

The calibration object now contains the information gained through optimization. We can use this information to check the fitted parameter values.
```{r fits}
mp_tmb_coef(cal, conf.int = TRUE)
```

These fits similar to the true values.
```{r true}
spec_for_making_fake_data$default[c("beta", "mu")]
```

We can also get the calibrated trajectory with confidence intervals.
```{r ci}
cal_traj = mp_trajectory_sd(cal, conf.int = TRUE) 
data = (nlist(true_traj, obs_traj, default_traj, cal_traj)
  |> bind_rows(.id = "data_type")
)
```

The calibrated trajectory and confidence interval are consistent with the true trajectory and go through the observed trajectory. The default trajectory is much different, indicating that calibration really did do what it should do.
```{r plot_fit}
(data
  |> ggplot()
  + geom_line(aes(time, value, colour = data_type))
  + geom_ribbon(aes(x = time, ymin = conf.low, ymax = conf.high)
      , alpha = 0.5
      , colour = "lightgrey"
      , fill = "lightgrey"
      , data = cal_traj
    ) 
  + theme_bw()
)
```

# References
