---
title: "Fitting to Data in the International Infectious Disease Data Archive (IIDDA)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

As usual we need the following packages.

```{r other_packages, message=FALSE}
library(macpan2)
library(dplyr)
library(ggplot2)
library(broom.mixed)
```

In addition, we need the `iidda.api` package for obtaining data from [this CANMOD-funded project](http://canmod.net/digitization).

```{r installation, message=FALSE}
if (!require(iidda.api)) {
  install.packages("iidda.api", repos = c("https://canmod.r-universe.dev", "https://cran.r-project.org"))
}
api_hook = iidda.api::ops_staging
```

And we set a few options for convenience.

```{r api_hook}
options(iidda_api_msgs = FALSE, macpan2_verbose = FALSE)
```


## One Scarlet Fever Outbreak in Ontario

The following code will pull data for a single scarlet fever outbreak in Ontario that ended in 1930.

```{r scarlet_fever_ontario}
scarlet_fever_ontario = api_hook$filter(resource_type = "CANMOD CDI"
  , iso_3166_2 = "CA-ON"  ## get ontario data
  , time_scale = "wk"  ## weekly incidence data only 
  , disease = "scarlet-fever"
  
  # get data between 1929-08-01 and 1930-10-01
  , period_end_date = "1929-08-01/1930-10-01"  
)
print(scarlet_fever_ontario)
```

This is what the data look like.

```{r scarlet_fever_ontario_plot, fig.width=7}
(scarlet_fever_ontario
  |> ggplot(aes(period_mid_date, cases_this_period))
  + geom_line() + geom_point()
  + ggtitle("Scarlet Fever Incidence in Ontario, Canada")
  + theme_bw()
)
```


## SIR Model

We will begin by pulling the simple sir model from the model library.

```{r}
sir = mp_tmb_library("starter_models", "sir", package = "macpan2")
print(sir)
```


## Modify Model for Reality

Once we take any library model off the shelve we need to modify it a bit to be more realistic. We do this with the `mp_tmb_insert` function, which inserts new default values and model expressions.

We first change the population size to the population of Ontario at the time.

```{r}
sir = mp_tmb_insert(sir
  , default = list(N = median(scarlet_fever_ontario$population))
)
```

Second, our SIR model assumes that all of the cases are recorded in the data, but this is not realistic.  We therefore modify the model to include under-reporting, by creating a case `reports` variable that is given by the product of incidence (i.e. weekly `infection` rate) and a reporting probability.

```{r}
sir = mp_tmb_insert(sir
    ## insert this expression ...
  , expressions = list(reports ~ infection * report_prob)
    
    ## at the end (i.e. Infinity) of the expressions evaluated
    ## 'during' each iteration of the simulation loop ...
  , at = Inf  
  , phase = "during"
  
    ## add a new default value for the reporting probability
  , default = list(report_prob = 1/300)
)
```

Here we can print out the modified model to see if our changes were made successfully.

```{r}
print(sir)
```


## Preparing the Data

The first step in preparing data for `macpan2` is to simulate from the model that you are considering. Here we simulate the `"infection"` variable because it corresponds to the incidence (the number of new cases per time-step, which in our case is one week).
```{r}
sir_simulator = mp_simulator(sir
  , time_steps = 5
  , outputs = "reports"
)
head(mp_trajectory(sir_simulator))
```

The next step is to get our data into the above format.

```{r}
observed_data = (scarlet_fever_ontario
  ## select the variables to be modelled -- a time-series of case reports.
  |> select(period_end_date, cases_this_period)
  
  ## change the column headings so that the match the columns
  ## in the simulated trajectories.
  |> mutate(matrix = "reports")
  |> rename(value = cases_this_period)
  
  ## create a `time` column with the time-step IDs that will correspond
  ## to the time-steps in the simulation. this column heading also 
  ## must match the column with the time-steps in the simulated trajectories
  |> mutate(time = seq_along(period_end_date))
)
print(head(observed_data))
```




## Set up the Optimizer

Now we can create an object that can be calibrated.

```{r}
sir_cal = mp_tmb_calibrator(
    spec = sir
  , data = observed_data
  
  ## name the trajectory variable, with a name that
  ## is the same in both the spec and the data
  , traj = "reports"  
  
  ## fit the following parameters
  , par = c("beta", "gamma", "I", "report_prob")
)
```


## Run the Optimization

```{r}
sir_opt = mp_optimize(sir_cal)
```



## Examine the fit

Note convergence ... etc...

```{r}
print(sir_opt)
```

Not off to a good start.  The `convergence` code is `1`, indicating that the model did _not_ converge (`convergence == 0` is good). Examining the parameter estimates, which are stored internally in the calibrator object, things get worse.

```{r}
mp_tmb_coef(sir_cal, conf.int = TRUE)
```

That doesn't look right!  Those are very high `beta` and `gamma` values, and the confidence intervals are enormous and overlap zero.

But the model fit doesn't look all that bad.

```{r, fig.width=7}
fitted_data = mp_trajectory_sd(sir_cal, conf.int = TRUE)
(observed_data
  |> ggplot()
  + geom_point(aes(time, value))
  + geom_line(aes(time, value)
    , data = fitted_data
    , colour = "red"
  )
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
    , data = fitted_data
    , alpha = 0.2
    , colour = "red"
  )
  + theme_bw()
)
```

What is going on?  Can you modify this model and/or fitting procedure to get both reasonable parameter estimates and a reasonable trajectory?
