---
title: "Lotka-Volterra"
index_entry: "simple predator-prey model"
bibliography: ../../references.bib
link-citations: TRUE
author: Jen Freeman, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
system.file("utils", "round-coef.R", package = "macpan2") |> source()
```

Lotka-Volterra predator-prey models with two competing species [@Hastings1997predatorprey]. This is not a compartmental model because individuals of one species do not become individuals of the other, unlike say an SIR model where susceptible individuals become infectious and then recovered. This predator-prey model, and this other [competition model](https://github.com/canmod/macpan2/tree/main/inst/starter_models/lotka_volterra_competition), illustrate how `macpan2` can be used for general dynamical modelling.

# Packages Used and Settings

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```

To keep the optimizer from printing too much in this article, we set the `macpan2_verbose` option to `FALSE`.

```{r options}
options(macpan2_verbose = FALSE)
```


# Model Specifications

These models have been specified in the `lotka_volterra_predator_prey` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/lotka_volterra_predator_prey/tmb.R) and are accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specifications using the `mp_tmb_library` command.
```{r model_spec}
specs = mp_tmb_library(
    "starter_models"
  , "lotka_volterra_predator_prey"
  , package = "macpan2"
  , alternative_specs = TRUE
)
```

# Species

| variable | description         |
| -------- | ------------------- |
| $X$      | number of prey      |
| $Y$      | number of predators |

# Parameters

| variable | description                                                                                                              |
| -------- | ------------------------------------------------------------------------------------------------------------------------ |
| $\alpha$ | per capita growth rate of prey, in the absence of predators (Hastings, 1997)                                             |
| $\gamma$ | per capita rate of predator loss, in the absence of prey                                                                 |
| $\beta$  | per capita mortality rate of prey fom predation                                                                          |
| $\delta$ | per capita growth rate of predators from predation                                                                       |
| $K$      | prey carrying capacity                                                                                                   |
| $f(X)$   | per predator predation rate as a function of the number of prey, also called [functional response](#functional-response) |

# Dynamics

## Simple Predator-Prey

In the simple predator-prey Lotka-Volterra model, we assume exponential growth and decay for prey and predators respectively.

$$
\begin{align*}
\frac{dX}{dt} &= \alpha X - \beta X Y \\
\frac{dY}{dt} &= \delta XY - \gamma Y
\end{align*}
$$

The model specification for this model is as follows.
```{r simple_spec}
specs$holling_1
```

## Logistic Prey Growth

We modify the simple model to include logisitic prey growth in the absence of predators with a prey carrying capacity of $K$.

$$
\begin{align*}
\frac{dX}{dt} &= \alpha X \left(1 - \frac{X}{K}\right)- \beta X Y \\
\frac{dY}{dt} &= \delta XY - \gamma Y
\end{align*}
$$

The model specification for this model is as follows.
```{r logistic_spec}
specs$holling_2
```

## Functional Response

The functional response $f(X)$ describes the predation rate as a function of prey density (Hastings, 1997). In the simplest case, we assume a linear function of prey, $f(X) = aX$ also called a [type I Holling](#holling-type-i) response. The simple predator-prey model includes a type I Holling response. Increasing functions that approach horizontal asymptotes can be used to represent more ecologically realistic predation rates to communicate that predation does not indefinitely increase when prey are abundant. The [type II Holling](#holling-type-ii) response is parameterized by the predator *attack rate* $a$, and the time elapsed by the predator capturing and consuming prey, called the *handling time* $h$ [@bolker2008]. A [Holling type III](#holling-type-iii) response is defined with higher powers of prey density.

### Holling type I

$$ f(X) = a X \:, a>0$$

### Holling type II

$$ f(X) = \frac{a X}{1+ ahX} \:,  a > 0  \:, h > 0$$

### Holling type III

$$ f(X) = \frac{a X^k}{1 + ah X^k} \:, k > 1$$


## General Form

The general predator-prey dynamics incorporating the functional response $f(X)$ is given below.

$$
\begin{align*}
\frac{dX}{dt} &= \alpha X\left(1-\frac{X}{K}\right) - f(X) Y\\
\frac{dY}{dt} &= \delta f(X)Y - \gamma Y 
\end{align*}
$$

Note if we parameterize the general dynamics by the inverse of the carrying capacity $K$, we can recover the exponential prey growth model by setting $K^{-1}=0$ and using the Holling type I response with $\beta$ as the attack rate. 

The model specification for this model is as follows.
```{r func_resp_spec}
specs$holling_3
```

# Calibration

## Simulate fake data

We modify the simple specification so that it is different from the default library model that we will calibrate.
```{r fake_data}
spec = specs$holling_1
set.seed(1L)
# set number of time steps in simulation
time_steps = 100L
# delta value to simulate data with
true = list(delta = 2.5/10)

# simulator object
fake_data = (spec
  |> mp_tmb_insert(
      phase = "during"
    , at = Inf
    , expressions = list(X_noisy ~ rpois(X), Y_noisy ~ rpois(Y))
    , default = true
  )
  |> mp_simulator(  
      time_steps = time_steps
    , outputs = c("X_noisy","Y_noisy")
  )
  |> mp_trajectory()
  |> mutate(matrix = substr(matrix, 1L, 1L))
)

(fake_data
  |> mutate(species = ifelse(matrix == "X", "prey", "predator"))
  |> ggplot()
  + geom_line(aes(time, value, colour = species))
  + theme_bw()
)
```

## Calibration

```{r}
cal = (spec
  |> mp_tmb_calibrator(
      data = fake_data
    , traj = c("X", "Y")
    , par = "delta"
  )
)
mp_optimize(cal)
```

## Explore the fit

The calibration object now contains the information gained through optimization. We can use this information to check the fitted parameter values.
```{r coef}
coef = mp_tmb_coef(cal) |> round_coef_tab()
coef$true = true[coef$mat]
print(coef)
```
The estimate is different enough from the default to indicate that the optimizer did 'something', and the estimate is close enough to the true value to indicate that it did the right thing. Indeed the fit to the data is excellent, although we should expect this given that we fitted the same model used to generate the data. Note well that it is usually not this easy (see the [seir](https://github.com/canmod/macpan2/tree/main/inst/starter_models/seir) model for a peak at what can go wrong).

Here is the fit to the data.

```{r traj_fit} 
comparison_data = list(
    obs = fake_data
  , fit = mp_trajectory_sd(cal, conf.int = TRUE)
) |> bind_rows(.id = "type") |> mutate(species = ifelse(matrix == "X", "prey", "predator"))
(comparison_data
  |> ggplot()
  + geom_line(aes(time, value, colour = type))
  + facet_wrap(~species)
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
    , colour = "red"
    , fill = "red"
    , alpha = 0.5
    , filter(comparison_data, type == "fit")
  )
  + theme_bw()
)
```


# References