---
title: "Macpan Base"
index_entry: "Re-implementation of the McMaster group's COVID-19 model"
bibliography: ../../references.bib
link-citations: TRUE
author: Jennifer Freeman
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/",
  message = FALSE,
  warning = FALSE
)
round_coef_tab = function(x, digits = 4) {
  id_cols = c("term", "mat", "row", "col", "type")
  num_cols = setdiff(names(x), id_cols)
  for (col in num_cols) {
    x[[col]] = round(x[[col]], digits)
  }
  rownames(x) = NULL
  cols_to_drop = c("term", "col", "type")
  x = x[, setdiff(names(x), cols_to_drop), drop = FALSE]
  x
}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
```

The McMasterPandemic model [@macpan], which is  modified SEIR model that incorporates additional infectious compartments to reflect the current knowledge of COVID-19 epidemiology. Exposed individuals enter four infectious compartments characterized by the individuals symptom status (asymptomatic, pre-symptomatic, mild, and severe). Severely infected individuals require care through hospital and/or Intensive Care Unit (ICU) compartments and either recover or die.

This model was used throughout the pandemic to support public health responses (e.g., [https://mac-theobio.github.io/forecasts/outputs/ON_accuracy.html](https://mac-theobio.github.io/forecasts/outputs/ON_accuracy.html)).

# Packages Used

The code in this article uses the following packages.

```{r packages, warning=FALSE, message=FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
```

# Model Specification

This model has been specified in the `macpan_base` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/macpan_base/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_lib}
spec = mp_tmb_library(
    "starter_models"
  , "macpan_base"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagrams using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/macpan_base/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 5, fig.width = 8}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_paths(spec, x_gap = 0.1, y_gap = 0.1)
plot_flow_diagram(layout)
```


# States

| variable | description                                                         |
| -------- | ------------------------------------------------------------------- |
| $S$      | Number of susceptible individuals                                   |
| $E$      | Number of exposed individuals                                       |
| $I_a$    | Number of asymptomatic infectious individuals                       |
| $I_p$    | Number of pre-symptomatic infectious individuals                    |
| $I_m$    | Number of mildly infectious individuals                             |
| $I_s$    | Number of severely infectious individuals                           |
| $H$      | Number of hospitalized individuals (acute care)                     |
| $ICU_s$  | Number of individuals admitted to the ICU with a survival prognosis |
| $ICU_d$  | Number of individuals admitted to the ICU with a death prognosis    |
| $H_2$    | Number of hospitalized individuals (acute care) after ICU stay      |
| $D$      | Number of dead individuals                                          |
| $R$      | Number of recovered individuals                                     |

The size of the total population is, $N = S + E + I_a + I_p + I_m + I_s + H +  ICU_s + ICU_d + H_2 + D + R$.

# Parameters

| variable       | description                                                                         |
| -------------- | ----------------------------------------------------------------------------------- |
| $\beta_0$      | baseline (non-intervention) transmission across categories                          |
| $C_a$          | relative asymptomatic transmission (or contact) proportion                          |
| $C_p$          | relative presymptomatic transmission (or contact) proportion                        |
| $C_m$          | relative mildly transmission (or contact) proportion                                |
| $C_s$          | relative severly transmission (or contact) proportion                               |
| $\alpha$       | fraction of infections that are asymptomatic                                        |
| $\mu$          | fraction of symptomatic infections that are mild                                    |
| $\sigma$       | 1/time in exposed class                                                             |
| $\gamma_a$     | 1/time to recovery for asymptomatic infections                                      |
| $\gamma_p$     | 1/time in pre-symptomatic state                                                     |
| $\gamma_m$     | 1/time to recovery for mildly symptomatic infections                                |
| $\gamma_s$     | 1/time spent in severely symptomatic state before either hospitalization or death   |
| $\rho$         | 1/time in hospital (initial acute care admission)                                   |
| $\delta_{nh}$  | probability of mortality without hospitalization                                    |
| $\text{iso}_m$ | relative self-isolation/distancing of mild cases                                    |
| $\text{iso}_s$ | relative self-isolation/distancing of severe cases                                  |
| $\phi_1$       | fraction of hospitalized infections that only require acute care (no ICU admission) |
| $\phi_2$       | fraction of ICU infections that are fatal                                           |
| $\psi_1$       | 1/time spent in ICU before returning to acute care                                  |
| $\psi_2$       | 1/time spent in ICU before dying                                                    |
| $\psi_3$       | 1/time in post-ICU acute care before hospital discharge                             |

# Differential Equations

<!-- nb. do not remove the spaces in `\delta _ {nh}`. https://github.com/github/markup/issues/1575 -->
$$
\begin{align*}
\frac{dS}{dt} &= -\beta_0\left(C_aI_a +C_pI_p + C_mI_m(1-\text{iso}_m) + C_sI_s(1-\text{iso}_s)\right)S/N \\
\frac{dE}{dt} &= \beta_0\left(C_aI_a +C_pI_p + C_mI_m(1-\text{iso}_m) + C_sI_s(1-\text{iso}_s)\right)S/N - \sigma E \\
\frac{dI_a}{dt} &= \alpha\sigma E- \gamma_a I_a \\
\frac{dI_p}{dt} &= (1-\alpha)\sigma E- \gamma_p I_p \\
\frac{dI_m}{dt} &= \mu\gamma_pI_p- \gamma_m I_m \\
\frac{dI_s}{dt} &= (1-\mu)\gamma_pI_p- (1-\delta _ {nh})\gamma_s I_s \\
\frac{dH}{dt} &= (1-\delta _ {nh})\phi_1\gamma_s I_s - \rho H  \\
\frac{dICU_s}{dt} &= (1-\delta _ {nh})(1-\phi_1)(1-\phi_2)\gamma_s I_s - \psi_1 ICU_s \\
\frac{dICU_d}{dt} &= (1-\delta _ {nh})(1-\phi_1)\phi_2\gamma_s I_s - \psi_2 ICU_d \\
\frac{dH_2}{dt} &= \psi_1 ICU_s - \psi_3 H_2  \\
\frac{dR}{dt} &= \gamma_a I_a + \gamma_m I_m + \rho H + \psi_3 H_2 \\
\frac{dD}{dt} &=  \psi_2 ICU_d
\end{align*}
$$

When there were discrepancies with how the model is expressed in Papst (TBD) versus model definition files, the latter was chosen.

# Calibration

## Observed data prep

We fit this model to observed death and case reports. For convenience we have saved some Ontario COVID-19 data in the `macpan2` package.
```{r get_data}
ts_data  = readRDS(url("https://github.com/canmod/macpan2/releases/download/macpan1.5_data/covid_on.RDS"))
```


To further prepare the time series data for calibration we filter for the appropriate time range and time series variables, create a numeric date field named 'time'.
```{r prep_ts}
prepped_ts_data = (ts_data
  |> rename(matrix = var)
  # dates from base model calibration (Figure 4)
  |> filter(date >= "2020-02-24" & date < "2020-08-31")
  # create unique time identifier
  |> arrange(date)
  |> group_by(date)
  |> mutate(time = cur_group_id())
  |> ungroup()
  # one negative value for daily deaths (removing for now time==178)
  # this explains negative values:
  # https://github.com/ccodwg/Covid19Canada?tab=readme-ov-file#datasets
  |> filter(matrix %in% c("death","report") & value >= 0)

)
```


To further prepare the mobility data for calibration we filter for the appropriate time range, create a numeric date field named 'time' and compute the logarithm of the mobility index.
```{r prep_mob}
prepped_mobility_data = (ts_data
  |> filter(var == "mobility_index")
  # dates from base model calibration (Figure 4)
  |> filter(date >= "2020-02-24" & date < "2020-08-31")
  # create unique time identifier
  |> arrange(date)
  |> group_by(date)
  |> mutate(time = cur_group_id())
  |> ungroup()
  |> mutate(log_mobility_ind = log(value))
)
```


Mobility breakpoints identified for piecewise varying transmission.
```{r mob_breaks}
mobility_breaks = (prepped_ts_data
  |> filter(date %in% c("2020-04-01", "2020-08-07"), matrix=="report")
  |> pull(time)
)
```

To support code below, we define the number of time steps in the model from the data.
```{r time_steps}
time_steps = nrow(prepped_ts_data %>% select(date) %>% unique())
prepped_ts_data = select(prepped_ts_data, -date)
```

## Mobility Sub-Model

The function `S_j` computes the logistic transition curve for mobility breakpoints, and this function is used to produce the model matrix, `X`, describing the temporal change in transmission using mobility data and piecewise breaks smoothed with the logistic curve.
```{r mobility_matrix}
S_j = function(t, tau_j, s) 1/(1 + exp((t - tau_j) / s))
X = cbind(
      prepped_mobility_data$log_mobility_ind
    , S_j(1:time_steps, mobility_breaks[1], 3)
    , S_j(1:time_steps, mobility_breaks[1], 3) * prepped_mobility_data$log_mobility_ind
    , S_j(1:time_steps, mobility_breaks[2], 3)
    , S_j(1:time_steps, mobility_breaks[2], 3) * prepped_mobility_data$log_mobility_ind
    ) %>% as.matrix()
X_sparse = macpan2:::sparse_matrix_notation(X, TRUE)
model_matrix_values = X_sparse$values
row_ind = X_sparse$row_index
col_ind = X_sparse$col_index
```


## Update the Model Specification for Calibration

Update model specification to include additional components described in manuscript and required for calibration.
```{r focal_model}
focal_model = (spec 
   # add variable transformations:
   |> mp_tmb_insert(phase = "before"
      , at = 1L
      , expressions = list(
          zeta ~ exp(log_zeta)
        , beta0 ~ exp(log_beta0)
        , nonhosp_mort ~ 1/(1+exp((-logit_nonhosp_mort)))
        , E ~ exp(log_E)
      )
      , default = list(
          log_zeta = empty_matrix
        , log_beta0 = empty_matrix
        , logit_nonhosp_mort = empty_matrix
        , log_E = empty_matrix
      )

   )
   
   # add accumulator variables:
   # death - new deaths each time step
   |> mp_tmb_insert(phase = "during"
      , at = Inf
      , expressions = list(death ~ ICUd.D + Is.D)
   )
   
   # add phenomenological heterogeneity:
   |> mp_tmb_update(phase = "during"
      , at =1L
      , expressions = list(
        mp_per_capita_flow(
            "S", "E"
          , "((S/N)^zeta) * (beta / N) * (Ia * Ca + Ip * Cp + Im * Cm * (1 - iso_m) + Is * Cs *(1 - iso_s))"
          , "S.E"
        ))

   )
   
   # compute gamma-density delay kernel for convolution:
   |> mp_tmb_insert(phase = "before"
      , at = Inf
      , expressions = list(
          gamma_shape ~ 1 / (c_delay_cv^2)
        , gamma_scale ~ c_delay_mean * c_delay_cv^2
        , gamma ~ pgamma(1:(qmax+1), gamma_shape, gamma_scale)
        , delta ~ gamma[1:(qmax)] - gamma[0:(qmax - 1)]
        , kappa ~ c_prop * delta / sum(delta)
      )
      , default = list(qmax = empty_matrix)
   )
   
   # add convolution to compute case reports from incidence:
   |> mp_tmb_insert(phase = "during"
      , at = Inf
      , expressions = list(report ~ convolution(S.E, kappa))
   )
   
   # add time-varying transmission with mobility data:
   |> mp_tmb_insert(phase = "before"
      , at = Inf
      , expressions = list(relative_beta_values ~ group_sums(model_matrix_values * log_mobility_coefficients[col_ind], row_ind, model_matrix_values))
      , default = list(log_mobility_coefficients = empty_matrix, model_matrix_values = empty_matrix)
      , integers = list(row_ind = row_ind, col_ind = col_ind)
   )
   |> mp_tmb_insert(phase = "during"
      , at = 1L
      , expressions = list(
        beta ~ exp(log_beta0 + relative_beta_values[time_step(1)])
      )
   )

)
```


## Calibration

```{r calibrator}
focal_calib = mp_tmb_calibrator(
    spec = focal_model |> mp_rk4()
  , data = prepped_ts_data
  , traj = c("report","death") 
  , par = c(
      "log_zeta"
    , "log_beta0"
    , "logit_nonhosp_mort"
    , "log_mobility_coefficients"
    , "log_E" 
    # negative binomial dispersion parameters for reports and deaths get added
    # automatically with options(macpan2_default_loss = "neg_bin") set above
  )
  , outputs = c("death","report")
  , default = list(
    
    # states
    # Population of Ontario (2019) from:
    # https://github.com/mac-theobio/macpan_base/blob/main/code/ontario_calibrate_comb.R
      S = 14.57e6 - 5 
    , log_E = log(5)
    , Ia = 0
    , Ip = 0
    , Im = 0
    , Is = 0
    , R = 0
    , H = 0
    , ICUs = 0
    , ICUd = 0
    , H2 = 0
    , D = 0
    
    , model_matrix_values = model_matrix_values
    
    # set initial parameter values for optimizer
    
    # width of convolution kernel computed according to:
    # https://canmod.net/misc/flex_specs#computing-convolutions
    # shape = 1/(mp_default_list(focal_model)$c_delay_cv^2)
    # scale = mp_default_list(focal_model)$c_delay_mean * mp_default_list(focal_model)$c_delay_cv^2
    # qmax = ceiling(qgamma(0.95, shape, scale))
    , qmax = 34
    , log_beta0=log(5)
    , logit_nonhosp_mort = -0.5
    , log_mobility_coefficients = rep(0,5)
    , log_zeta = 1
    
  )
)
# converges
mp_optimize(focal_calib)
```

## Explore Calibration

We can get the trajectory and confidence intervals.
```{r get_trajectory}
fitted_data = mp_trajectory_sd(focal_calib, conf.int = TRUE)
```

```{r coefficients}
# View estimates and their confidence intervals:
# zeta ~ 0(wide CI), do we need phenomenological heterogeneity?
# beta0 ~ 0.18, order of magnitude smaller than manuscript estimate (Table 1)
# mobility coefficients:
#   mobility power                                 ~ 3.6 (huge std err)
#   relative change in transmission (breakpoint 1) ~ 3.9 (3.4,4.6)
#   change in mobility power (breakpoint 1)        ~ 1.04 (0.8,1.4)
#   relative change in transmission (breakpoint 2) ~ 0.64 (0.56,0.74)
#   change in mobility power (breakpoint 2)        ~ 0.21 (wide CI)
# E(0) ~ 173(102,295) plausible for early pandemic (but different from manuscript)
# nohosp_mort ~ 0.14(0.11,0.16) is this too high?
# theta_report ~ 3(2.8,3.3) very different from manuscript
# theta_death ~ 0.5(0.2,0.8) very different from manuscript
mp_tmb_coef(focal_calib, conf.int = TRUE)
```


```{r plot_fit}
(ggplot(prepped_ts_data, aes(time,value))
  + geom_point()
  + geom_line(aes(time, value)
              , data = fitted_data |> filter(matrix %in% c("death","report"))
              , colour = "red"
  )
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
                , data = fitted_data |> filter(matrix %in% c("death","report"))
                , alpha = 0.2
                , colour = "red"
  )
  + facet_wrap(vars(matrix),scales = 'free')
  + theme_bw()
)
```



# Computing $\mathcal{R}_0$ with a Cohort Model

The following is some advanced material that we are planning to include as a feature for calculating $\mathcal{R}_0$ in the package for 

Simulate a single exposed individual through time (N=1, E=1, all other states 0). The transmission kernel is the force of infection at each time step of the simulation. We sum the transmission kernel to get an estimate on $\mathcal{R}_0$ (for a given set of parameters?)

In our case, our focal model includes phenomenological heterogeneity and setting S to 0, leads to 0^0 issues. The way out, I think, is to realize that the kernel method assumes we are at the beginning of the epidemic and therefore that S/N ~ 1.  In this case, FOI reduces to beta * I / N, which presents no issue.  So we do not need safe_power.

Update focal model model to remove inflow to E (ensuring no new susceptible individuals reach E). This update also automatically turns off outflow from S, which is also what we want because this makes S/N ~ 1. In this case we don't initialize S to 0. We have N = N_focal (full population), `N_cohort` = cohort population (1 in this case).
```{r cohort_model}
cohort_model_ph = (
  focal_model
  %>% mp_tmb_update(phase="during"
                    , at = 2L
                    , expressions = list(S.E ~ (S^zeta) * (beta / (N^(zeta) * N_cohort)) * (Ia * Ca + Ip * Cp + Im * Cm * (1 - iso_m) + Is * Cs *(1 - iso_s)))
                    , default = list(S.E = 0))
)
```

This cohort model allows us to simulate the cohort.
```{r simulate_cohort}
cohort_sim_ph = (mp_simulator(cohort_model_ph
                  , time_steps = 100L
                  , outputs = "S.E"
                  , default = list(
                      S = 14.57e6 - 1 
                    , log_E = log(1)
                    , Ia = 0
                    , Ip = 0
                    , Im = 0
                    , Is = 0
                    , R = 0
                    , H = 0
                    , ICUs = 0
                    , ICUd = 0
                    , H2 = 0
                    , D = 0
                    , N_cohort = 1
                    , model_matrix_values = model_matrix_values
                    , qmax = 34
                    , log_beta0=log(1)
                    , logit_nonhosp_mort = -0.5
                    , log_mobility_coefficients = rep(0,5)
                    , log_zeta = 1
                    )
  ) |> mp_trajectory()
) 
```

If we ran our simulations for long enough, the sum of these simulations is $\mathcal{R}_0$.
```{r R0}
R0_ph = sum(cohort_sim_ph$value)
print(R0_ph)
```

Should we turn off phenomenological heterogeneity, or turn off inflow to E? If so, update S.E flow to remove phenomenological heterogeneity add duplicate foi expression because we can't recover the foi = S.E / S, when S is 0. In this case we do initialize S to 0.
```{r cohort_no_ph}
cohort_model = (focal_model
  %>% mp_tmb_update(phase="during"
                    , at = 2L
                    , expressions = list(mp_per_capita_flow(
        "S", "E"
      , "(beta / N) * (Ia * Ca + Ip * Cp + Im * Cm * (1 - iso_m) + Is * Cs * (1 - iso_s))"
      , "S.E" 
    )
  ))
  %>% mp_tmb_insert(phase="during"
                    , at = Inf
                    , expressions = list(foi ~ (beta / N) * (Ia * Ca + Ip * Cp + Im * Cm * (1 - iso_m) + Is * Cs *(1 - iso_s)))
  )
)

cohort_sim = (mp_simulator(cohort_model
                           , time_steps = 100L
                           , outputs = "foi"
                           , default = list(
                               S = 0
                             , log_E = log(1)
                             , Ia = 0
                             , Ip = 0
                             , Im = 0
                             , Is = 0
                             , R = 0
                             , H = 0
                             , ICUs = 0
                             , ICUd = 0
                             , H2 = 0
                             , D = 0
                             , N = 1
                             , model_matrix_values = model_matrix_values
                             , qmax = 34
                             , log_beta0=log(1)
                             , logit_nonhosp_mort = -0.5
                             , log_mobility_coefficients = rep(0,5)
                             , log_zeta = 1) # not actually used
  ) |> mp_trajectory()
) 

R0 = sum(cohort_sim$value)
print(R0)
```

It turns out that we arrive at approximately the same answer with both approaches.
```{r check_R0}
all.equal(R0_ph, R0)
```

To solve for intrinsic growth rate `r`, we find the root of the Euler-Lotka equation.
```{r euler_lotka}
euler_lotka = function(r) sum(cohort_sim$value * exp(-r * cohort_sim$time)) - 1
uniroot(euler_lotka, c(0,10))
```


# References 
