---
title: "Wastewater Model"
index_entry: "Macpan base with an additional wastewater component"
bibliography: ../../references.bib
link-citations: TRUE
author: Jennifer Freeman, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
round_coef_tab = function(x, digits = 4) {
  id_cols = c("term", "mat", "row", "col", "type")
  num_cols = setdiff(names(x), id_cols)
  for (col in num_cols) {
    x[[col]] = round(x[[col]], digits)
  }
  rownames(x) = NULL
  cols_to_drop = c("term", "col", "type")
  x = x[, setdiff(names(x), cols_to_drop), drop = FALSE]
  x
}
```

The McMasterPandemic model [@macpan] modified to include a wastewater component.

# Packages and Settings Used

The code in this article uses the following packages.

```{r packages, warning=FALSE, message=FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
library(lubridate)

```

To keep the optimizer from printing too much in this article, we set the `macpan2_verbose` option to `FALSE`.

```{r options}
options(macpan2_verbose = FALSE)
```

# Model Specification

This model has been specified in the `ww` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/ww/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_lib}
spec = mp_tmb_library(
    "starter_models"
  , "ww"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagrams using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/ww/README.Rmd). For clarity, we first draw the epidemiological components of the model, which are identical to the [macpan_base](https://github.com/canmod/macpan2/tree/main/inst/starter_models/macpan_base) model.

```{r diagram, echo = FALSE, fig.height = 4, fig.width = 8.5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout_base = mp_layout_paths(spec, x_gap = 0.1, y_gap = 0.1, trim_states = c("W", "A"))
layout_ww = mp_layout_paths(spec, x_gap = 0.1, y_gap = 0.1, trim_states = c("A", setdiff(layout_base$states(), c("Ia", "Ip", "Im", "Is"))))
plot_flow_diagram(layout_base)
```

The infectious compartments of the above epidemiological component shed virus into wastewater with the following sub-model.

```{r diagram_shedding, echo = FALSE, fig.height = 2, fig.width = 3}
plot_flow_diagram(layout_ww)
```

# States

| variable | description                                                            |
| -------- | ---------------------------------------------------------------------- |
| $S$      | Number of susceptible individuals                                      |
| $E$      | Number of exposed individuals                                          |
| $I_a$    | Number of asymptomatic infectious individuals                          |
| $I_p$    | Number of pre-symptomatic infectious individuals                       |
| $I_m$    | Number of mildly infectious individuals                                |
| $I_s$    | Number of severely infectious individuals                              |
| $H$      | Number of hospitalized individuals (acute care)                        |
| $ICU_s$  | Number of individuals admitted to the ICU with a survival prognosis    |
| $ICU_d$  | Number of individuals admitted to the ICU with a death prognosis       |
| $H_2$    | Number of hospitalized individuals (acute care) after ICU stay         |
| $D$      | Number of dead individuals                                             |
| $R$      | Number of recovered individuals                                        |
| $W$      | Concentration of viral particles in wastewater                         |
| $A$      | Total accumulated concentration of virus in wastewater over time       |

The size of the total population is, $N = S + E + I_a + I_p + I_m + I_s + H +  ICU_s + ICU_d + H_2 + D + R$.

# Parameters

| variable       | description                                                                         |
| -------------- | ----------------------------------------------------------------------------------- |
| $\beta_0$      | baseline (non-intervention) transmission across categories                          |
| $C_a$          | relative asymptomatic transmission (or contact) proportion                          |
| $C_p$          | relative presymptomatic transmission (or contact) proportion                        |
| $C_m$          | relative mildly transmission (or contact) proportion                                |
| $C_s$          | relative severly transmission (or contact) proportion                               |
| $\alpha$       | fraction of infections that are asymptomatic                                        |
| $\mu$          | fraction of symptomatic infections that are mild                                    |
| $\sigma$       | 1/time in exposed class                                                             |
| $\gamma_a$     | 1/time to recovery for asymptomatic infections                                      |
| $\gamma_p$     | 1/time in pre-symptomatic state                                                     |
| $\gamma_m$     | 1/time to recovery for mildly symptomatic infections                                |
| $\gamma_s$     | 1/time spent in severely symptomatic state before either hospitalization or death   |
| $\rho$         | 1/time in hospital (initial acute care admission)                                   |
| $\delta_{nh}$  | probability of mortality without hospitalization                                    |
| $\phi_1$       | fraction of hospitalized infections that only require acute care (no ICU admission) |
| $\phi_2$       | fraction of ICU infections that are fatal                                           |
| $\psi_1$       | 1/time spent in ICU before returning to acute care                                  |
| $\psi_2$       | 1/time spent in ICU before dying                                                    |
| $\psi_3$       | 1/time in post-ICU acute care before hospital discharge                             |
| $\nu$          | viral shedding rate to wastewater                                                   |
| $\xi$          | rate at which virus is denaturing/removed from wastewater                           |

# Dynamics 

<!-- nb. do not remove the spaces in `\delta _ {nh}`. https://github.com/github/markup/issues/1575 -->
$$
\begin{split}
\frac{dS}{dt} &= -\beta_0\left(C_aI_a +C_pI_p + C_mI_m(1-\text{iso}_m) + C_sI_s(1-\text{iso}_s)\right)S/N \\
\frac{dE}{dt} &= \beta_0\left(C_aI_a +C_pI_p + C_mI_m(1-\text{iso}_m) + C_sI_s(1-\text{iso}_s)\right)S/N - \sigma E \\
\frac{dI_a}{dt} &= \alpha\sigma E- \gamma_a I_a \\
\frac{dI_p}{dt} &= (1-\alpha)\sigma E- \gamma_p I_p \\
\frac{dI_m}{dt} &= \mu\gamma_pI_p- \gamma_m I_m \\
\frac{dI_s}{dt} &= (1-\mu)\gamma_pI_p- (1-\delta _ {nh})\gamma_s I_s \\
\frac{dH}{dt} &= (1-\delta _ {nh})\phi_1\gamma_s I_s - \rho H  \\
\frac{dICU_s}{dt} &= (1-\delta _ {nh})(1-\phi_1)(1-\phi_2)\gamma_s I_s - \psi_1 ICU_s \\
\frac{dICU_d}{dt} &= (1-\delta _ {nh})(1-\phi_1)\phi_2\gamma_s I_s - \psi_2 ICU_d \\
\frac{dH_2}{dt} &= \psi_1 ICU_s - \psi_3 H_2  \\
\frac{dR}{dt} &= \gamma_a I_a + \gamma_m I_m + \rho H + \psi_3 H_2 \\
\frac{dD}{dt} &=  \psi_2 ICU_d \\
\frac{dW}{dt} &= \nu I_a + \nu I_p + \nu I_m + \nu I_s - \xi W \\
\frac{dA}{dt} &= \xi W
\end{split}
$$


# Calibration



## Observed Data Prep

We load in an object containing previous information on default parameters.
```{r default_load}
repo = "https://github.com/canmod/macpan2/releases/download/macpan1.5_data"
macpan1.5 = (repo
  |> file.path("covid_on_macpan1.5_calibration.RDS")
  |> url() 
  |> readRDS()
)
```

We also load in Ontario COVID-19 data that comes with the package.
```{r covid_data}
covid_on = (repo
  |> file.path("covid_on.RDS")
  |> url() 
  |> readRDS()
)
```

Prep the observed incidence and wastewater data to calibrate to.
```{r prep}
obs_data = (covid_on
    |> filter(var %in% c("report","W") & between(date,macpan1.5$settings_sim$start_date,macpan1.5$settings_sim$end_date))
    |> rename(time = date, matrix = var)
    |> mutate(matrix = ifelse(matrix == "report", "reported_incidence", matrix))
    # leading zeroes seem to cause calibration challenges
    |> filter(!is.na(value) & time > as.Date("2020-02-23"))
)
```



## Calibration Model Specification

Update model specification to include additional components required for calibration.
```{r cal_spec_update}
focal_model = (
  # waste water model with hazard correction
  spec |> mp_hazard()
     
  # add variable transformations:
  |> mp_tmb_insert(phase = "before"
     , at = 1L
     , expressions = list(
          beta0 ~ exp(log_beta0)
        , nu ~ exp(log_nu)
        , xi ~ exp(log_xi)
        , report_prob ~ 1 / (1 + exp(-logit_report_prob))
     )
     , default = list(
          log_beta0 = log(1)
        , log_nu = 0
        , log_xi = 0
        , logit_report_prob = qlogis(0.1)
     )
  )

  # add accumulator variables:
  # death - new deaths each time step
  |> mp_tmb_insert(phase = "during"
     , at = Inf
     , expressions = list(
          death ~ ICUd.D + Is.D
       ,  prevalence ~ Ip + Ia + Im + Is
      )
  )
     
  # add phenomenological heterogeneity:
  |> mp_tmb_update(phase = "during"
     , at = 1L
     , expressions = list(mp_per_capita_flow("S", "E", "((S/N)^zeta) * (beta / N) * (Ia * Ca + Ip * Cp + Im * Cm * (1 - iso_m) + Is * Cs *(1 - iso_s))", "incidence"))

  )
     
  # add convolution to compute case reports from incidence:
  |> mp_tmb_insert_reports("incidence"
     , report_prob = NA_real_  # defined above with a logit transform
     , mean_delay = 11
     , cv_delay = 0.25
     , report_prob_name = "report_prob"
  )

  # decompose beta for time-varying transmission
  |> mp_tmb_insert(phase = "during"
     , at = 1L
     , expressions = list(beta ~ beta0 * beta1 * beta2)
  )
  
  |> mp_tmb_insert(phase = "during"
     , at = 1L
     , expressions = list(
          beta2 ~ time_var(beta_changes, beta_changepoints)
     )
     , default  = list(beta_changes      = c(1))
     , integers = list(beta_changepoints = c(0))
  )
  
  # log transform W
  |> mp_tmb_insert_trans("W", mp_log)
)
```

## Calibrate to Data

```{r}
useful_params = names(macpan1.5$params) %in% names(focal_model$default)
macpan1.5_defaults = c(
    macpan1.5$params[useful_params]
  , list(
      S = macpan1.5$params$N
    , log_beta0 = log(macpan1.5$params$beta0)
    , log_nu = log(macpan1.5$params$nu)
    , beta1 = 1
    , E = macpan1.5$params$E0
  )
)

# calibrator
focal_calib = mp_tmb_calibrator(
    spec = focal_model
  , data = obs_data
  , traj = list(
      # set likelihoods for trajectories we are fitting to
      reported_incidence = mp_nbinom(disp = 0.1)
    , W = mp_norm(sd = 1)
  )
  , par = c(
        # parameters to fit
        "log_beta0" # baseline transmission
      , "log_nu"    # viral shedding rate
      , "log_xi"
      #, "logit_report_prob" # reporting probability (only makes sense to fit if you fit to incidence)
  )
    # flexible non-parametric (radial basis function) fit
    # for the time-varying transmission rate
  , tv = mp_rbf("beta1", dimension = 7, sparse_tol = 1e-4)
    
    # return these trajectories so that they can be
    # explored after fitting
  , outputs = c(
        "reported_incidence", "beta"
      , "prevalence", "incidence", "S", "log_W"
    )
  
    # update defaults with macpan1.5 wastewater model defaults
  , default = macpan1.5_defaults
  
  , time = mp_sim_offset(15, 0, "daily")
)
# converges
mp_optimize(focal_calib)
```


## Explore Fits

The fitted model parameters, fitted values, and confidence intervals.
```{r get_fits}
macpan2_fit = mp_trajectory_sd(focal_calib, conf.int = TRUE)
fitted_coefs = mp_tmb_coef(focal_calib, conf.int = TRUE)
print(fitted_coefs)
```

Plot the fitted values.
```{r plot_fits}
plot_fit = function(obs_data, sim_data, ncol = 1L) {
  (ggplot(obs_data, aes(time, value))
   + geom_point()
   + geom_line(aes(
          time
        , value
        , colour = source
      )
      , data = sim_data
   )
   #+ scale_y_log10()
   + geom_ribbon(aes(
          time
        , ymin = conf.low
        , ymax = conf.high
        , colour = source
        , fill = source
      )
      , data = sim_data
      , alpha = 0.2
   )
   + facet_wrap(vars(matrix), scales = 'free', ncol = ncol)
   + theme_bw()
   + theme(legend.position = "top")
   #+ scale_x_date(limits = c(as.Date("2020-11-01"), NA))
  )
}
plot_fit(obs_data, mutate(macpan2_fit, source = "macpan2_fit"), ncol = 2L)
```

# References 
