---
title: "Wastewater model"
index_entry: "Macpan base with an additional wastewater component"
bibliography: ../../references.bib
link-citations: TRUE
author: Jennifer Freeman
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
round_coef_tab = function(x, digits = 4) {
  id_cols = c("term", "mat", "row", "col", "type")
  num_cols = setdiff(names(x), id_cols)
  for (col in num_cols) {
    x[[col]] = round(x[[col]], digits)
  }
  rownames(x) = NULL
  cols_to_drop = c("term", "col", "type")
  x = x[, setdiff(names(x), cols_to_drop), drop = FALSE]
  x
}
```

The McMasterPandemic model [@macpan] modified to include a wastewater component.

# Packages Used

The code in this article uses the following packages.

```{r packages, warning=FALSE, message=FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
```

# Model Specification

This model has been specified in the `ww` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/ww/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_lib}
spec = mp_tmb_library(
    "starter_models"
  , "ww"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagrams using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/ww/README.Rmd). For clarity, we first draw the epidemiological components of the model.

```{r diagram, echo = FALSE, fig.height = 4, fig.width = 8.5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout_base = mp_layout_paths(spec, x_gap = 0.1, y_gap = 0.1, trim_states = c("W", "A"))
layout_ww = mp_layout_paths(spec, x_gap = 0.1, y_gap = 0.1, trim_states = c("A", setdiff(layout_base$states(), c("Ia", "Ip", "Im", "Is"))))
plot_flow_diagram(layout_base)
```

The infectious compartments of the above epidemiological component shed virus into wastewater with the following sub-model.

```{r diagram_shedding, echo = FALSE, fig.height = 2, fig.width = 3}
plot_flow_diagram(layout_ww)
```

# States

| variable | description                                                            |
| -------- | ---------------------------------------------------------------------- |
| $S$      | Number of susceptible individuals                                      |
| $E$      | Number of exposed individuals                                          |
| $I_a$    | Number of asymptomatic infectious individuals                          |
| $I_p$    | Number of pre-symptomatic infectious individuals                       |
| $I_m$    | Number of mildly infectious individuals                                |
| $I_s$    | Number of severely infectious individuals                              |
| $H$      | Number of hospitalized individuals (acute care)                        |
| $ICU_s$  | Number of individuals admitted to the ICU with a survival prognosis    |
| $ICU_d$  | Number of individuals admitted to the ICU with a death prognosis       |
| $H_2$    | Number of hospitalized individuals (acute care) after ICU stay         |
| $D$      | Number of dead individuals                                             |
| $R$      | Number of recovered individuals                                        |
| $W$      | Concentration of viral particles in wastewater                         |
| $A$      | Total accumulated concentration of virus in wastewater over time       |

The size of the total population is, $N = S + E + I_a + I_p + I_m + I_s + H +  ICU_s + ICU_d + H_2 + D + R$.

# Parameters

| variable       | description                                                                         |
| -------------- | ----------------------------------------------------------------------------------- |
| $\beta_0$      | baseline (non-intervention) transmission across categories                          |
| $C_a$          | relative asymptomatic transmission (or contact) proportion                          |
| $C_p$          | relative presymptomatic transmission (or contact) proportion                        |
| $C_m$          | relative mildly transmission (or contact) proportion                                |
| $C_s$          | relative severly transmission (or contact) proportion                               |
| $\alpha$       | fraction of infections that are asymptomatic                                        |
| $\mu$          | fraction of symptomatic infections that are mild                                    |
| $\sigma$       | 1/time in exposed class                                                             |
| $\gamma_a$     | 1/time to recovery for asymptomatic infections                                      |
| $\gamma_p$     | 1/time in pre-symptomatic state                                                     |
| $\gamma_m$     | 1/time to recovery for mildly symptomatic infections                                |
| $\gamma_s$     | 1/time spent in severely symptomatic state before either hospitalization or death   |
| $\rho$         | 1/time in hospital (initial acute care admission)                                   |
| $\delta_{nh}$  | probability of mortality without hospitalization                                    |
| $\phi_1$       | fraction of hospitalized infections that only require acute care (no ICU admission) |
| $\phi_2$       | fraction of ICU infections that are fatal                                           |
| $\psi_1$       | 1/time spent in ICU before returning to acute care                                  |
| $\psi_2$       | 1/time spent in ICU before dying                                                    |
| $\psi_3$       | 1/time in post-ICU acute care before hospital discharge                             |
| $\nu$          | viral shedding rate to wastewater                                                   |
| $\xi$          | rate at which virus is denaturing/removed from wastewater                           |

# Dynamics 

<!-- nb. do not remove the spaces in `\delta _ {nh}`. https://github.com/github/markup/issues/1575 -->
$$
\begin{split}
\frac{dS}{dt} &= -\beta_0\left(C_aI_a +C_pI_p + C_mI_m(1-\text{iso}_m) + C_sI_s(1-\text{iso}_s)\right)S/N \\
\frac{dE}{dt} &= \beta_0\left(C_aI_a +C_pI_p + C_mI_m(1-\text{iso}_m) + C_sI_s(1-\text{iso}_s)\right)S/N - \sigma E \\
\frac{dI_a}{dt} &= \alpha\sigma E- \gamma_a I_a \\
\frac{dI_p}{dt} &= (1-\alpha)\sigma E- \gamma_p I_p \\
\frac{dI_m}{dt} &= \mu\gamma_pI_p- \gamma_m I_m \\
\frac{dI_s}{dt} &= (1-\mu)\gamma_pI_p- (1-\delta _ {nh})\gamma_s I_s \\
\frac{dH}{dt} &= (1-\delta _ {nh})\phi_1\gamma_s I_s - \rho H  \\
\frac{dICU_s}{dt} &= (1-\delta _ {nh})(1-\phi_1)(1-\phi_2)\gamma_s I_s - \psi_1 ICU_s \\
\frac{dICU_d}{dt} &= (1-\delta _ {nh})(1-\phi_1)\phi_2\gamma_s I_s - \psi_2 ICU_d \\
\frac{dH_2}{dt} &= \psi_1 ICU_s - \psi_3 H_2  \\
\frac{dR}{dt} &= \gamma_a I_a + \gamma_m I_m + \rho H + \psi_3 H_2 \\
\frac{dD}{dt} &=  \psi_2 ICU_d \\
\frac{dW}{dt} &= \nu I_a + \nu I_p + \nu I_m + \nu I_s - \xi W \\
\frac{dA}{dt} &= \xi W
\end{split}
$$


# Model Specification

This model has been specified in the `ww` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/ww/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). 

# References 
