---
title: "Mosquito-Vector  SIR"
index_entry: "SIR model for mosquito vectors"
bibliography: ../../references.bib
link-citations: TRUE
author: Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```


This [model](https://homepages.warwick.ac.uk/~masfz/ModelingInfectiousDiseases/Chapter4/Program_4.4/index.html) from [@keeling2011modeling] describes the transmission dynamics between humans and mosquitoes, focusing on a vector-borne disease.

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(macpan2)
library(ggplot2)
```

# States

| **Variable** | **Description**                                  |
|--------------|--------------------------------------------------|
| $S_H$          | Number of susceptible humans                     |
| $I_H$          | Number of infected humans                        |
| $S_M$          | Number of susceptible mosquitoes                 |
| $I_M$          | Number of infected mosquitoes                    |

# Parameters

| **Variable**  | **Description**                                  |
|---------------|--------------------------------------------------|
| $r$             | Rate at which humans are bitten |
| $\beta_{ij}$             | Transmission probability (following a bite) to species $i$ from species $j$.|
| $\gamma_i$             | Recovery rate for host species $i$.                          |
| $\nu_i$             | Birth rate for host species $i$.                                 |
| $\mu_i$             | Death rate for species $i$                    |

# Dynamics

The model is described by the following differential equations:

$$
\begin{align*}
\frac{dS_H}{dt} &= \nu_H - r \beta_{HM} S_H I_M - \mu_H S_H \\
\frac{dI_H}{dt} &= r \beta_{HM} S_H I_M - \gamma_H I_H - \mu_H I_H \\
\frac{dS_M}{dt} &= \nu_M - r \beta_{MH} S_M I_H - \mu_M S_M \\
\frac{dI_M}{dt} &= r \beta_{MH} S_M I_H - \gamma_M I_M - \mu_M I_M \\
\end{align*}
$$

# Model Specification

This model has been specified in the `sir_mosquito` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_mosquito/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). 

```{r library}
spec = mp_tmb_library("starter_models", "sir_mosquito", package = "macpan2")
```

This specification can be used to draw the following flow diagram using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_mosquito/README.Rmd). The dashed lines show the state dependence of infection, which is the only way that the human and mosquito sub-models interact (e.g., mosquitos never become humans, they just infect them). The per-capita infection rate for humans depends on the number of infectious mosquitos, and vice versa.

```{r diagram, echo = FALSE, fig.height = 3, fig.width = 8.5}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
repl = "^([a-z]*)(_.*)" ## remove after unscore of flow names
layout = mp_layout_paths(spec)
(layout
  |> plot_flow_diagram(show_flow_names = TRUE
    , pattern_edge_mutate = repl
    , pattern_edge_replace = "\\1"
    , state_dependence = mp_state_dependence_frame(spec)
  ) 
  |> draw_outflows(layout, show_labels = TRUE
    , x = "xmax", y = "y"
    , x_dir = "east", y_dir = "central"
    , pattern_filter = "^recovery"
    , pattern_mutate = repl
    , pattern_replace = "\\1"
  )
  |> draw_outflows(layout, show_labels = TRUE
    , y = "ymax", y_dir = "north"
    , pattern_filter = "^death_(S|I)_H"
    , pattern_mutate = repl
    , pattern_replace = "\\1"
  )
  |> draw_outflows(layout, show_labels = TRUE
    , y = "ymin", y_dir = "south"
    , pattern_filter = "^death_(S|I)_M"
    , pattern_mutate = repl
    , pattern_replace = "\\1"
  )
  |> draw_inflows(layout, show_labels = TRUE
    , pattern_mutate = repl
    , pattern_replace = "\\1"
  )
)
```


# Simulation

Simulation of this and other models depends on the kind of [state update](https://canmod.github.io/macpan2/reference/mp_euler) that you use. Here we use RK4, which is a standard ODE solver.


```{r simulations}
(spec
  |> mp_rk4()
  |> mp_simulator(time_steps = 500, outputs = c("S_H", "I_H", "S_M", "I_M"))
  |> mp_trajectory(include_initial = TRUE)
  |> ggplot()
  + facet_wrap(~matrix, scales = "free")
  + geom_line(aes(time, value))
  + theme_bw()
)
```

# References
