---
title: "Age-stratified SIR"
index_entry: "An age-stratified SIR model"
bibliography: ../../references.bib
link-citations: TRUE
author: Irena Papst, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "./figures/"
)
```

This is an expansion of the [sir](https://github.com/canmod/macpan2/tree/main/inst/starter_models/sir) model to include age-stratification; each state variable is subdivided by age, and transmission is governed by a age-based contact matrix, as described in [@Mistry2021].

# Packages Used and Settings

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(forcats)
library(scales)
library(macpan2)
```

# Model Specification

This model has been specified in the `sir_age` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_age/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
spec = mp_tmb_library(
    "starter_models"
  , "sir_age"
  , package = "macpan2"
)
```



# States

| variable | description                       |
| -------- | --------------------------------- |
| S        | Number of susceptible individuals |
| I        | Number of infectious individuals  |
| R        | Number of recovered individuals   |

Each state variable is a vector with one entry per age group. 

The size of the population in each age groups is given by the components of $N = S + I + R$.

# Parameters

| variable | description                  |
| -------- | ---------------------------- |
| $\tau$  | transmissibility, _i.e._, the proportion of contacts with infecteds that yield infection |
| $\gamma$ | per capita recovery rate     |
| $M$    | matrix of contact rates between age groups, _e.g._ $M_{ij}$ is the rate of contact between susceptibles of age $i$ and infecteds of age $j$        |

$\tau$ and $\gamma$ can be specified by age (as vectors), though in the example below we use a scalar value that applies to all age groups.

# Dynamics 

$$
\begin{align*}
\frac{dS}{dt} &= - \lambda S \\
\frac{dI}{dt} &= \lambda S - \gamma I \\
\frac{dR}{dt} &= \gamma I
\end{align*}
$$

Here $\lambda = \tau M \frac{I}{N}$.

# Simulation Example

```{r plot_sim}

# variable labellers
matrix_labeller <-  c(
  "S" = "Susceptible",
  "I" = "Infectious",
  "R" = "Recovered"
)
row_labeller <- c(
  "0" = "Children (ages 0-17)",
  "1" = "Adults (ages 18-64)",
  "2" = "Seniors (ages 65+)"
)

I0 = mp_default_list(spec)$I[1]
N = mp_default_list(spec)$N

# simulate & visualize
(spec
  # simulate
  |> mp_simulator(
    time_steps = 150
  , outputs = mp_state_vars(spec)
  )
  |> mp_trajectory()
  # enforce natural order of variables for plotting
  |> dplyr::mutate(
    matrix = factor(matrix, levels = mp_state_vars(spec))
   , row = forcats::as_factor(row)
   )
  # visualize trajectory
  |> ggplot2::ggplot(
    mapping = ggplot2::aes(x = time, y = value, colour = row)
  ) 
  + ggplot2::geom_line()
  + ggplot2::facet_wrap(
      ~ matrix, 
      labeller = ggplot2::labeller(
        matrix = matrix_labeller
      ),
      scales = "free_y")
  + ggplot2::scale_colour_discrete(
      labels = row_labeller
    )
  + ggplot2::labs(
      title = "Simulation of the age-structured SIR model with two age groups",
      subtitle = paste("Infection is seeded with", I0, "infectious children in a total population of", scales::label_number(scale_cut = scales::cut_short_scale())(sum(N))),
      x = "Days from beginning of outbreak",
      y = "Number of individuals"
    )
  + ggplot2::theme(
      legend.position = "bottom"
    , legend.title = ggplot2::element_blank()
    )
)
```

# Limitations

Models with vector-valued state variables cannot be simulated with the [mp_euler_multinomial](https://canmod.github.io/macpan2/reference/state_updates) state update method. If you try you will get the following error.
```{r error, echo = FALSE}
x = mp_euler_multinomial(spec)
try(mp_simulator(x, 1, "S") |> mp_trajectory())
```

# References
