---
title: "NFDS and Vaccine Design"
index_entry: "An ecological model using population genomics to design optimal vaccines as implemented in Colijn et al. (2020)"
bibliography: references.bib
link-citations: TRUE
author: Jennifer Freeman
output: 
  github_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

```{r packages, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```


<!-- population-->
*Streptococcus pneumoniae* (pneumococcus) is a bacteria that can cause *invasive pneumococcal disease* (IPD), such as meningitis and bacterial pnuemonia, in vulnerable populations (ie. young children and elderly) [@phac_2023]. Pneumococci typically colonize the respiratory tract and its subsequent occupation is called *carriage* [@coughtrie_2019]. Carriage pneumococcus is common and individuals are often asymptomatic, becoming *carriers* of the bacterium [@phac_2023]. Pneumococci populations are further subdivided by their epidemiological characteristics (*serotypes*) and genetic variation (*genotypes* and the *loci* they encode).


<!-- vaccine/motivation -->
*Protein–polysaccharide conjugate vaccines* (PCVs) were introduced to target small subsets of pneumococcal serotypes to reduce IPD burden [@colijn_2020]. In a carriage population after PCV introduction, vaccine serotypes are removed through herd immunity [@colijn_2020]. Pneumococcal carriage prevalence is known to be relatively stable over time as *serotype replacement* occurs where non-vaccine serotypes compete and replace vaccine types [@corander_2017]. The effect of PCVs in a post-vaccine carriage population can be problematic when non-vaccine serotypes that are highly *invasive*, "the rate at which they progress from carriage to cause IPD", or anti-microbial resistant become more prevalent [@colijn_2020]. 

<!-- the problem -->
@colijn_2020 formulated optimal vaccine designs that minimized IPD burden in post-vaccine carriage populations. Two distinct carriage population datasets were included in model experiments. One from a vaccinated group of children in Massachusetts and the second, an unvaccinated group from the Maela refugee camp in Thailand. In simulations, a hypothetical vaccine was introduced into the population, and a duration of 10 years was selected to evaluate the effect of this vaccine on the carriage community make-up of pneumococcus.

<!-- the model -->
A multi-locus *negative frequency-dependent selection* (NFDS) model was implemented as a first order ODE to simulate the dynamics of genotype prevalence in carriage over time. This model assumes "alleles are most beneficial to genotypes when they are rare" and the fitness of each genotype is  reflected in its dynamic reproductive rate [@corander_2017]. This study ignores genetic *recombination*, assuming the set of unique genotypes in each population is fixed over time [@colijn_2020]. Genotype prevalences in @colijn_2020 were simulated using an ODE solver in Matlab.

<!-- the optimization -->
IPD burden was computed using serotype invasiveness estimates, derived from a meta-analysis, and the post-vaccine carriage prevalences (the ODE solution). Vaccine design was optimized using Bayesian optimization and a genetic algorithm, with additional constraints on vaccine valency and forced serotype inclusion [@colijn_2020].


# States


*Note:* Input values provided here are for the Massachusetts population ($M=603$, $L=1090$), to get corresponding Maela input values see [write_data.m](https://github.com/canmod/macpan2/blob/7126c2217272367eca8e3ea9b9a1fe8e0c6d3d59/inst/starter_models/nfds/data/write_data.m).

| variable   | description                            | input value                                                                                                                                      |
| ---------- | -------------------------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------ |
| $Y_{i}(t)$ | prevalence of genotype $i$ at time $t$ | [intial conditions (t=0)](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/ics.csv) |



# Parameters

*Note:* Input values provided here are for the Massachusetts population ($M=603$, $L=1090$), to get corresponding Maela input values see [write_data.m](https://github.com/canmod/macpan2/blob/7126c2217272367eca8e3ea9b9a1fe8e0c6d3d59/inst/starter_models/nfds/data/write_data.m).

## Fixed


| variable  | description                                                                                                                                                               | csv input file                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
| --------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| $w_l$     | NFDS weight for locus $l$                                                                                                                                                 | [locusweights](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/locusweights.csv)                                                                                                                                                                                                                                                                                                                                                                                           |
| $e_l$     | equilibrium frequency for locus $l$                                                                                                                                       | [locusfreq](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/locusfreq.csv)                                                                                                                                                                                                                                                                                                                                                                                                 |
| $\kappa$  | carriage population carrying capacity                                                                                                                                     | hardcoded, $\kappa = 10,000$                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                             |
| $G_{i,l}$ | $G=$ genotype-loci matrix (size $M$ by $L$). $G_{i,l}=1$ if genotype $i$ encodes locus $i$, 0 otherwise.                                                                  | [G](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/G.csv)                                                                                                                                                                                                                                                                                                                                                                                                                 |
| $\rho$    | constant representing the "overall strength of NFDS" [@colijn_2020], computed as $\log(1 + \sigma)$ where $\sigma$ is taken from @corander_2017.           | [sigma](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/sigma.csv)                                                                                                                                                                                                                                                                                                                                                                                                         |
| $v$       | vaccine efficacy parameter                                                                                                                                                | [v](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/v.csv)                                                                                                                                                                                                                                                                                                                                                                                                                 |
| $m$       | uniform migration rate per genotype                                                                                                                                       | [m](https://github.com/canmod/macpan2/blob/b162156ca3d2787de90f5c080827ee9547b14fab/inst/starter_models/nfds/data/m.csv)                                                                                                                                                                                                                                                                                                                                                                                                                 |
| $r_i$     | vaccine strategy, $r_i=r$ "if genotype $i$ encodes an antigen included in the vaccine" [@colijn_2020], 0 otherwise. Computed as $r_i = -\log(1- (v \text{vax}_i))$ | The user specifies a vector of serotypes to be included, excluded, and *varied* (varied means these serotypes will be optimized i.e. optimizer decides if each serotype in this vector will be included/excluded) in the vaccine. This vaccine vector is then mapped to the level of genotypes, creating a binary vector $\text{vax}$. (See an example of user input for the `macpan2` model implementation [here](https://github.com/canmod/macpan2/blob/7126c2217272367eca8e3ea9b9a1fe8e0c6d3d59/inst/starter_models/nfds/tmb.R#L77)). |



## Time Varying
| variable  | description                                              | calculation                                                 |
| --------- | -------------------------------------------------------- | ----------------------------------------------------------- |
| $N(t)$    | carriage population of pneumococci organisms at time $t$ | $N(t) = \sum\limits_{i=1}^M Y_i$                            |
| $f_{l,t}$ | instantaneous frequency for locus $l$ at time $t$        | $f_{l,t} = \frac{1}{N(t)} \sum\limits_{i=1}^{M} Y_i G_{il}$ |


# Dynamics 

$$\frac{dY_i}{dt} = \left( \log\left( \frac{\kappa}{N(t)}\right) -r_i + \rho\left( \sum_{l=1}^{L} w_l G_{il} (e_l - f_{l,t})\right) \right) Y_i + m, \quad i \in \\{1, M\\}$$

# Model Specification

This model has been specified in the `nfds` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/nfds/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). 

```{r model_spec}
spec = mp_tmb_library("starter_models","nfds", package="macpan2")

```

# Model Validation


## Simulating Data

As a first step in validating the model, we can simulate data from the model to visualize the behaviour of genotype prevalence curves.

```{r simulated_prevalence}
# set number of time steps in simulation
# time_steps = 120L is equal to 10 years (as in Colijn et al. 2020)
# step size = months
time_steps = 120L 

# create simulator
nfds = mp_simulator(  
    model = spec
  , time_steps = time_steps
  , outputs = c("Y","vax_strain","vax_serotypes")
) 

# simulate Y (genotype prevalence state vector) and plot
(nfds
  |> mp_trajectory()
  |> filter(matrix=='Y')
  |> mutate(genotype=as.factor(row))
  |> ggplot()
  + geom_line(aes(time,value,col=genotype))
  + theme_bw()
  + theme(legend.position="none")
  + ylab("prevalence")
  + xlab("time (months)")
  + ggtitle("Prevalence of all genotypes (M=603)",subtitle = "macpan2 implementation")
)
```


We expect that prevalences of all genotypes should reach equilibirium prevalences (level off) with sufficient time steps (although @colijn_2020 mentions curves may not reach mathematical equilibrium, but invasiveness equilibriates faster). These curves look reasonable because there is some asymptotic behaviour in the later years. 

If we colour code the genotype curves by whether they are included in the vaccine (In vaccine = 1) or not (In vaccine = 0) we gain more information about the dynamics.

```{r simulated_prevalence_by_vax}
# get vaccine mapping by strain/genotype
in_vax = (nfds
          |> mp_trajectory()
          |> filter(matrix %in% c("vax_strain"))
          |> mutate(in_vax=as.factor(value))
)

# simulate Y (genotype prevalence) and plot, coloured by vaccine inclusion
(nfds
  |> mp_trajectory()
  |> filter(matrix=='Y')
  |> rename(prevalence=value)
  |> left_join(in_vax, by=c("time", "col", "row"))
  |> ggplot()
  + geom_line(aes(time,prevalence,group=row,col=in_vax),alpha=0.8)
  + scale_colour_manual(name = 'In vaccine',values=c('darkgrey','#670000'))
  + theme_bw()
  + ylab("prevalence")
  + xlab("time (months)")
  + ggtitle("Prevalence of all genotypes (M=603)",subtitle = "macpan2 implementation")
) 

```

We expect serotype replacement to take place, so that genotypes not included in the vaccine will dominate the population after vaccine perturbation (with sufficient time steps), and this graphically seems to be happening. Vaccine type strains decrease in prevalence as they are eliminated from carriage.

### A quick note on run-time when constructing simulators

In general, we know constructing a simulator in `macpan2` is more computationally intensive than simulating from it (see [Creating a Simulator](https://canmod.github.io/macpan2/articles/quickstart.html#creating-a-simulator)). On top of this, we make choices about how our model is specified and this can also contribute to lengthy run-times.

With earlier versions of the NFDS `macpan2` model specification, the wall-clock time to construct the simulator was significant. Experimentation attributed a large portion of this time to the multiplication of the matrix $G$, when computing the instantaneous locus frequencies, $f_{l,t}$, and the NFDS term, $\left( \sum_{l=1}^{L} w_l G_{il} (e_l - f_{l,t}) \right)$, referred to as $\pi_{i,t}(G,Y)$ in @colijn_2020. Model specification changes were made to fix this and a short demonstration is given to highlight these differences when computing $f_{l,t}$.

<!-- -->

```{r get_params, include=FALSE}
# get input parameters and initial conditions
G = system.file("starter_models","nfds","data","G.csv", package="macpan2") %>% read.csv(header=FALSE) %>% as.matrix() %>% unname()
w = system.file("starter_models","nfds","data","locusweights.csv", package="macpan2") %>% read.csv(header=FALSE)  %>% as.matrix() %>% t() %>% unname()
e = system.file("starter_models","nfds","data","locusfreq.csv", package="macpan2") %>% read.csv(header=FALSE)  %>% as.matrix() %>% t() %>% unname()
Y = system.file("starter_models","nfds","data","ics.csv", package="macpan2")%>% read.csv(header=FALSE)  %>% as.matrix() %>% t() %>% unname()
N=1e5
```

The expression for $f_{l,t}$ in the old model specification involved the matrix multiplication of $G$.

```{r old_expr}
# old expression
old_expr = list(
  # compute f_{i,t}
  f ~ (1/N) * (Y %*% G)
)
```

The current model spec computes $f_{l,t}$ differently by using the fact that $G$ is a relatively sparse matrix containing only zeroes and ones. The idea being that we can represent $G$ with two vectors that contain the column and row indices of $G$ that have an entry of 1, and this eliminates the need to provide $G$ to the model specification explicitly. 

```{r new_expr}
# new expression
new_expr = list(
  # compute f_{i,t}
  f ~ (1/N) * group_sums(Y[Gt_i], Gt_j, f)
)
```

We define a `binary_matrix_notation` function that returns these indices as integer vectors named `col_index` and `row_index` for a binary matrix `M`. If `M` is not square, its multiplication by a vector on the right versus on the left will determine whether we need indices for `M` or its transpose `t(M)`. Note `binary_matrix_notation` returns zero-based indices (see `?macpan2::group_sums` for details).

```{r binary_matrix_notation}
binary_matrix_notation <- function(M){
  
  # get col indices
  col_index = c(t(col(M)*M))
  col_index = as.integer(col_index[col_index!=0]-1) 
  
  # get row indices
  row_index = as.integer(rep(1:nrow(M), times=rowSums(M))-1)
  
  return(macpan2:::nlist(col_index,row_index))
}

# get non-zero indices of G transpose (loci-genotype matrix)
# because we have a left multiplication of G (Y %*% G)
Gt_ind <- binary_matrix_notation(t(G)) 

# save indices to a list
integers = list(
    Gt_i = Gt_ind$col_index                                          
  , Gt_j = Gt_ind$row_index                                         
)

```

$G$ has `r ncol(G)*nrow(G)` entries, where `r length(which(G!=0))` entries have a value of 1. Since we need a row and column index for each non-zero entry, this means we need twice the number of non-zero entries in $G$ to be able to represent it with integer vectors. This new specification reduces the dimension of the computation because we need smaller objects to store the same information.

```{r dim_reduction}
# number of entries in integer indices
length(integers$Gt_i)+length(integers$Gt_i)
```

We now define the model specifications. In the old version, we needed to provide the matrix $G$ in the list of defaults. The new version no longer needs $G$, and we can introduce the integer indices in the model using the `integers` argument.

```{r old_vs_new_spec}
# old specification
old_spec = mp_tmb_model_spec(
    during = old_expr
  , default = list(
      N = N
    , G = G
    , Y = Y
  )
)
# new specification
new_spec = mp_tmb_model_spec(
    during = new_expr
  , default = list(
       N = N
     , Y = Y
     # group_sums needs 'n' argument (see ?macpan2::group_sums)
     # so we need to initialize f with length = L (# intermedifate frequency loci)
     , f = rep(0,dim(G)[2]) 
    ) 
  , integers = integers
)
```


We can now compare the wall-clock time of creating a simulator object for both specifications. 

```{r new_time}
# old spec
start_time <- Sys.time()
old_sim = mp_simulator(old_spec,time_steps = 1L,outputs = "Y")
end_time <- Sys.time()
end_time - start_time

# new spec
start_time <- Sys.time()
new_sim = mp_simulator(new_spec,time_steps = 1L,outputs = "Y")
end_time <- Sys.time()
end_time - start_time
```

Creating a simulator with one time step results in the new model specification being twice as fast as the old, and this difference grows with increasing time steps.

## Comparing Model Implementations

To further validate the `macpan2` NFDS implementation, we can compare simulation results with simulated data from @colijn_2020.

We first choose a simulation scenario, using the Massachusets data set with the PCV7 vaccine. Note, serotype '4' is not present in this data set. I suspect this serotype had low initial carriage prevalence and was excluded from the population.

With simulated data from both models, we compare each genotypes simulated prevalence at the final time point (10 years).

```{r model_sims, include=FALSE}
# simulated data from Colijn et al. is generated from
# nfds/data/simulate_data.m
sim_colijn = (
  read.csv(
    system.file(
        "starter_models"
      , "nfds"
      , "data"
      , "simdata_PCV7.csv"
      , package = "macpan2"
    )
    , header = FALSE
  )
  |> as.matrix()
  |> t()
  |> data.frame()
  |> mutate(genotype=row_number())
  |> pivot_longer(cols= -c(genotype), names_to = "time", values_to = "prevalence")
  |> mutate(time=as.integer(gsub("^X","",time,perl=TRUE)),model='colijn')
  |> arrange(time,genotype)
)

# simulated data from macpan2
# Note: the PCV7 vax is the default vaccine in the model spec tmb.R file
# so no changes are required here
sim_macpan2 = (nfds
               |> mp_trajectory()
               |> filter(matrix=='Y')
               |> mutate(genotype=as.integer(row)+1L,
                          model='macpan2')
               |> rename(prevalence="value")
               |> dplyr::select(-c(row,col,matrix))
               |> arrange(time,genotype)
)

# arrange data
# long
sim_long = (union(sim_colijn, sim_macpan2))

# wide
sim_wide = (sim_colijn 
            |> rename(colijn=prevalence) 
            |> select(-model)
            |> inner_join(
              (sim_macpan2 
               |> rename(macpan2=prevalence) 
               |> select(-model)
               )
              , by = join_by(genotype, time)
              ) 
            |> mutate(`colijn - macpan2`= colijn - macpan2)
)
```


```{r prevalence_10years, echo=FALSE}
# prevalence of both model implementations at final time step
(ggplot((sim_wide %>% filter(time==time_steps)), aes(x=colijn,y=macpan2))
  + geom_point(pch=1)
  + theme_bw()
  + ggtitle("Prevalence by model implementation at 10 years")
)

```

This plot confirms that the `macpan2` implementation of this model generates very similar results to those simulated from the @colijn_2020 model implementation in Matlab. 

If we look further at prevalence differences by genotype we can see that the simulated data from both models are not identical.

```{r difference_10years, echo=FALSE}
# prevalence difference of both model implementations at final time step
(ggplot((sim_wide %>% filter(time==time_steps)), aes(x=genotype,y=`colijn - macpan2`))
  + geom_col()
  + theme_bw()
  + ggtitle("Prevalence difference per genotype at 10 years")
  + xlab("genotype ID")
)
```

The largest absolute difference between the two models is approximately `r round(max(abs(subset(sim_wide,time==time_steps)[,"colijn - macpan2"])))` individuals in a population of roughly $N(t)=10000$. This equates to approximately `r (round(max(abs(subset(sim_wide,time==time_steps)[,"colijn - macpan2"])))/1e5)*100` % of the population, and I suspect this difference is not biologically significant. 


```{r prep_data, include=FALSE}
# choose random subset of genotypes so we can visualize
# differences over time easier
set.seed(123)
n_genotypes = 8 # choosing 8
dark2=c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D", "#666666")
# genotypes with high prevalence initial conditions
high_ics = (sim_wide 
  |> filter(time==1 & colijn > 500)
  |> select(genotype)
  |> unique()
  |> sample_n(n_genotypes)
  |> mutate(initial_prevalence="high",
            line_col=dark2)
)
# genotypes with low prevalence initial conditions
low_ics = (sim_wide
   |> filter(time==1 & colijn < 500)
   |> select(genotype)
   |> unique()
   |> anti_join(high_ics, by=join_by(genotype))
   |> sample_n(n_genotypes)
   |> mutate(initial_prevalence="low"
             ,line_col=dark2)
)
# union all
all_ics = union(high_ics, low_ics)


rand_genotypes = ((sim_wide 
           %>% inner_join(all_ics, by=join_by(genotype)) 
           %>% select(-c(colijn, macpan2))
           %>% rename(value=`colijn - macpan2`) 
           %>% mutate(value_desc='colijn - macpan2', model=value_desc)
           ) |> union(sim_long 
             %>% inner_join(all_ics, by=join_by(genotype)) 
             %>% rename(value=prevalence)
             %>% mutate(value_desc='prevalence')
           )
)

# set colours
genotype_cols = c(dark2,dark2)
names(genotype_cols) = rand_genotypes |> select(genotype) |> unique() |> pull()
```

Now that we know there are some differences between the two model implementations, it makes sense to visualize how these models compare over the entire 10 year time period. As an initial hypothesis, we might expect differences to grow over time as floating point rounding errors accumulate.

We select a random subset of genotypes to visualize (`r n_genotypes` with high starting prevalence, and `r n_genotypes` with low starting prevalence) because it's difficult to distinguish between curves when all genotypes (M=603) are graphed.


```{r plots_over_time, echo=FALSE}

 (ggplot(rand_genotypes %>% filter(value_desc=="prevalence"))
  + geom_line(aes(x=time, y=value, col=as.factor(genotype), linetype=model))
  + theme_bw()
  + facet_wrap(vars(initial_prevalence),labeller = "label_both",ncol=2,scales="free_y")
  + xlab("time (months)")
  + ylab("prevalence")
  # +scale_y_log10()
  + scale_colour_manual(values = genotype_cols)
  + guides(col="none",linetype=guide_legend(nrow = 1))
  + ggtitle("Genotype prevalence over time",subtitle="random subset of genotypes (n=16)")
  + theme(legend.position='bottom')
 )

 (ggplot(rand_genotypes %>% filter(value_desc=='colijn - macpan2'))
  + geom_line(aes(x=time, y=value, colour=as.factor(genotype)))
  + theme_bw()
  + facet_wrap(vars(initial_prevalence),labeller = "label_both",ncol=2,scales="free_y")
  + xlab("time (months)")
  + ylab('colijn - macpan2')
  + scale_colour_manual(values = genotype_cols)
  + guides(col="none")
  + ggtitle("Genotype prevalence difference over time",subtitle="random subset of genotypes (n=16)")
  + geom_hline(yintercept = 0, col="black", linetype="dashed")
 )


```

We see some minimal differences when looking at the top panel of prevalence over time. In the difference comparison, we can see larger differences at earlier time points and these differences seem to get smaller on average over time, with one exception. The sea green coloured curve in the bottom right plot shows that prevalence differences are growing over time. When we look at this genotype in the top right plot, we can see that this specific genotype is growing in prevalence much faster than any of the others. This highlights that the difference in prevalences could be better represented as a relative measure.

## Summary

In general, we conclude that there are some differences between the two model implementations but they are not biologically significant. We can attribute these numerical differences to the approximations made in each model implementation. In the `macpan2` implementation we are using the first-order Euler method with the C++ TMB engine for ODE solving as compared to the implementation in @colijn_2020, who use the ODE solver `ode15s` in Matlab.

# Optimization

No optimization was performed in the `macpan2` implementation, as the discrete variable being optimized is not well suited for the TMB engine. A possible future direction for optimization could be [rBayesianOptimization](https://cran.r-project.org/web/packages/rBayesianOptimization/rBayesianOptimization.pdf).

# References

