---
title: "Basic SI"
index_entry: "A very simple epidemic model"
bibliography: ../../references.bib
link-citations: TRUE
author: Jennifer Freeman
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

This is the simplest possible epidemic model [@earn2008light].

```{r flow_diagram, echo = FALSE, fig.height=1, fig.width=4, message=FALSE, warning=FALSE}
library(macpan2)
library(ggplot2)
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
si = mp_tmb_library("starter_models", "si", package = "macpan2")
(si 
  |> mp_layout_paths()
  |> plot_flow_diagram(show_flow_rates = TRUE)
)
```

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(macpan2)
library(ggplot2)
library(dplyr)
```

# States

| variable | description                       |
| -------- | --------------------------------- |
| S        | Number of susceptible individuals |
| I        | Number of infectious individuals  |

The size of the total population is, $N = S + I$.

# Parameters

| variable | description                  |
| -------- | ---------------------------- |
| $\beta$  | per capita transmission rate |

# Dynamics 

$$
\begin{align*}
\frac{dS}{dt} &= -\beta S\frac{I}{N} \\
\frac{dI}{dt} &= \beta S\frac{I}{N}\\
\end{align*}
$$

# Model Specification

This model has been specified in the `si` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/si/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). 

# Simulation

Simulation of this and other models depends on the kind of [state update](https://canmod.github.io/macpan2/reference/mp_euler) that you use. Here we compare Euler, Euler-multinomial, hazard, and RK4.

```{r simulation}
specs = mp_tmb_library("starter_models"
  , "si"
  , package = "macpan2"
  , alternative_specs = TRUE
)
set.seed(1)
(specs
  |> lapply(mp_simulator, 50L, "I")
  |> lapply(mp_trajectory)
  |> bind_rows(.id = "integrator")
  |> rename(prevalence = value)
  |> ggplot()
  + geom_line(aes(time, prevalence, colour = integrator))
  + theme_bw()
)
```

# References
