---
title: "SIR with waning immunity"
index_entry: "a basic SIR model with a flow from R back to S"
bibliography: ../../references.bib
link-citations: TRUE
author: Jen Freeman, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
system.file("utils", "round-coef.R", package = "macpan2") |> source()
```

Endemic pathogens can sometimes be modelled by sending R back to S, thereby controlling susceptible depletion such that new infections keep arising indefinitely. This is an extension of the [sir](https://github.com/canmod/macpan2/tree/main/inst/starter_models/sir) model to include such immunity waning.

# Packages Used and Settings

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```

To keep the optimizer from printing too much in this article, we set the `macpan2_verbose` option to `FALSE`.

```{r options}
options(macpan2_verbose = FALSE)
```


# Model Specification

This model has been specified in the `sir_waning` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_waning/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
spec = mp_tmb_library(
    "starter_models"
  , "sir_waning"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagram using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_waning/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 2, fig.width = 8}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_paths(spec
  , ignore = "waning_immunity"
  , loops = "waning_immunity"
)
p = (layout
  |> plot_flow_diagram()
  + geom_curve(aes(x = xmin_from, y = ymin_from, xend = xmax_to, yend = ymin_from)
    , data = layout$edges_flows_ignored()
    , curvature = -0.2
    , colour = "blue"
    , arrow = mp_flow_arrow
  )
  + ylim(c(0.6, NA))
)
print(p)
```

# States

| variable | description                       |
| -------- | --------------------------------- |
| S        | Number of susceptible individuals |
| I        | Number of infectious individuals  |
| R        | Number of recovered individuals   |

The size of the total population is, $N = S + I + R$.

# Parameters

| variable | description                     |
| -------- | ------------------------------- |
| $\beta$  | per capita transmission rate    |
| $\gamma$ | per capita recovery rate        |
| $\phi$   | per capita waning immunity rate |

# Dynamics 

$$
\begin{align*}
\frac{dS}{dt} &= -\beta S\frac{I}{N} + \phi R\\
\frac{dI}{dt} &= \beta S\frac{I}{N} - \gamma I \\
\frac{dR}{dt} &= \gamma I  - \phi R
\end{align*}
$$

# Calibration

## Simulate fake data

We modify the specification so that it is different from the default library model that we will calibrate. Then we simulate prevalence data from this model, and add noise.

```{r simulation}
set.seed(1L)
time_steps = 100L
true = list(
    beta = 0.3 # beta value to simulate data with
  , phi = 0.05 # phi value to simulate data with
)
sir_waning = (spec
  |> mp_tmb_insert(
      phase = "during", at = Inf
    , expressions = list(noisy_I ~ rpois(I))
    , default = true
  )
  |> mp_simulator(  
      time_steps = time_steps
    , outputs = c("noisy_I", "I")
  )
)
  
observed_data = (sir_waning
  |> mp_trajectory() 
  |> mutate(matrix = ifelse(matrix == "I", "true_I", "I"))
)
(observed_data
  |> ggplot()
  + geom_line(aes(time, value, colour = matrix))
  + theme_bw()
)
```


## Calibrate the model

```{r calibration}
cal = mp_tmb_calibrator(
    spec
  , data = filter(observed_data, matrix == "I")
  , traj = "I"
  , par = c("beta", "phi")
)
mp_optimize(cal)
```

## Explore the fit

The calibration object now contains the information gained through optimization. We can use this information to check the fitted parameter values.
```{r coef}
coef = mp_tmb_coef(cal) |> round_coef_tab()
coef$true = true[coef$mat]
print(coef)
```
The estimate is different enough from the default to indicate that the optimizer did 'something', and the estimate is close enough to the true value to indicate that it did the right thing. Indeed the fit to the data is excellent, although we should expect this given that we fitted the same model used to generate the data. Note well that it is usually not this easy (see the [seir](https://github.com/canmod/macpan2/tree/main/inst/starter_models/seir) model for a peak at what can go wrong).

```{r traj_fit} 
comparison_data = list(
    obs = observed_data
  , fit = mp_trajectory_sd(cal, conf.int = TRUE)
) |> bind_rows(.id = "type")
(comparison_data
  |> filter(matrix == "I")
  |> ggplot()
  + geom_line(aes(time, value, colour = type))
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
    , colour = "red"
    , fill = "red"
    , alpha = 0.5
    , filter(comparison_data, type == "fit")
  )
  + theme_bw()
)
```

# References
