---
title: "SIR with waning immunity"
index_entry: "a basic SIR model with a flow from R back to S"
bibliography: ../../references.bib
link-citations: TRUE
author: Jen Freeman, Steve Walker
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

Endemic pathogens can sometimes be modelled by sending R back to S, thereby controlling susceptible depletion such that new infections keep arising indefinitely. This is an extension of the [sir](https://github.com/canmod/macpan2/tree/main/inst/starter_models/sir) model to include such immunity waning.

# Packages Used

The code in this article uses the following packages.

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(macpan2)
```

# Model Specification

This model has been specified in the `sir_waning` directory [here](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_waning/tmb.R) and is accessible from the `macpan2` model library (see [Example Models](https://canmod.github.io/macpan2/articles/example_models.html) for details). We can read in the model specification using the `mp_tmb_library` command.
```{r model_spec}
spec = mp_tmb_library(
    "starter_models"
  , "sir_waning"
  , package = "macpan2"
)
```

This specification can be used to draw the following flow diagram using code found in the [source for this article](https://github.com/canmod/macpan2/blob/main/inst/starter_models/sir_waning/README.Rmd).

```{r diagram, echo = FALSE, fig.height = 2, fig.width = 8}
system.file("utils", "box-drawing.R", package = "macpan2") |> source()
layout = mp_layout_paths(spec
  , ignore = "waning_immunity"
  , loops = "waning_immunity"
)
p = (layout
  |> plot_flow_diagram(show_flow_names = TRUE)
  + geom_curve(aes(x = xmin_from, y = ymin_from, xend = xmax_to, yend = ymin_from)
    , data = layout$edges_flows_ignored()
    , curvature = -0.2
    , colour = "blue"
    , arrow = mp_flow_arrow
  )
  + ylim(c(0.6, NA))
)
print(p)
```

# States

| variable | description                       |
| -------- | --------------------------------- |
| S        | Number of susceptible individuals |
| I        | Number of infectious individuals  |
| R        | Number of recovered individuals   |

The size of the total population is, $N = S + I + R$.

# Parameters

| variable | description                     |
| -------- | ------------------------------- |
| $\beta$  | per capita transmission rate    |
| $\gamma$ | per capita recovery rate        |
| $\phi$   | per capita waning immunity rate |

# Dynamics 

$$
\begin{align*}
\frac{dS}{dt} &= -\beta S\frac{I}{N} + \phi R\\
\frac{dI}{dt} &= \beta S\frac{I}{N} - \gamma I \\
\frac{dR}{dt} &= \gamma I  - \phi R
\end{align*}
$$

# References
